---
title: "2. Joint Learning of F in Poisson SuSiE"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This page extends the single-observation experiment to a count matrix. It turns
the simulation at the end of `code/joint-learn-susie-poi-F.R` into a
reproducible workflowr analysis, and fits matrix-factorization Poisson SuSiE
while updating the shared factor matrix `F`.

The key question is whether the model can recover both pieces of structure:
the dictionary rows \(F_{k\cdot}\) and the sparse loading supports for each
observation.

```{r source-code}
source("code/joint-learn-susie-poi-F.R")
```

## Fitted Model

For a count matrix \(Y \in \mathbb{N}^{N \times M}\), the fitted model uses a
shared nonnegative factor matrix \(F \in \mathbb{R}_+^{K \times M}\) and a
small number \(D\) of single-effect components per observation:
\[
  Y_{im} \sim \mathrm{Poisson}\left(
    \sum_{d=1}^D \beta_{id} F_{\gamma_{id}m}
  \right).
\]
Each effect chooses one factor row through
\(\gamma_{id} \in \{1,\ldots,K\}\), and its size is \(\beta_{id} > 0\).
The variational fit estimates posterior assignment probabilities
`gamma_bar[i, d, k]`, Gamma variational parameters for the effect sizes, and,
in the unknown-`F` fit, the shared factor matrix `F`.

The simulation below first checks recovery when the true factor matrix `F0` is
given to the algorithm, then repeats the fit with `F` initialized by Poisson NMF
and learned jointly.

## Simulate Data

We create eight observation groups. Groups 1 through 5 each use one factor, and
groups 6 through 8 use three-factor combinations.

```{r simulate}
set.seed(42)

N = 1600
M = 2000
K = 5
D = 3

F0 = matrix(rgamma(M * K, shape = 0.1, rate = 0.01), nrow = K, ncol = M)
F0 = F0 / rowSums(F0)

n_per_group = N / 8
L = matrix(0, nrow = N, ncol = K)
grp = rep(1:8, each = n_per_group)

for (g in 1:5) {
  L[grp == g, g] = matrix(rgamma(n_per_group, 10, 1), n_per_group, 1)
}

L[grp == 6, c(1, 2, 3)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 7, c(1, 2, 4)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 8, c(2, 3, 5)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)

rate = L %*% F0
Y = matrix(rpois(N * M, rate), nrow = N, ncol = M)

data.frame(
  N = N,
  M = M,
  K = K,
  D = D,
  total_count = sum(Y),
  nonzero_entries = sum(Y > 0)
)
```

```{r true-patterns}
true_patterns = vapply(1:8, function(g) {
  paste(which(colSums(L[grp == g, , drop = FALSE]) > 0), collapse = ",")
}, character(1))

data.frame(group = 1:8, true_active_factors = true_patterns)
```

```{r true-factor-shapes}
matplot(t(F0[, 1:150]), type = "l", lty = 1,
        xlab = "m", ylab = "F0[k, m]",
        main = "True factor rows")
legend("topright", legend = paste("factor", 1:K), col = 1:K, lty = 1,
       bty = "n")
```

```{r diagnostic-helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)

  if (K <= 8) {
    perms = all_permutations(seq_len(K))
    scores = apply(perms, 1, function(p) {
      sum(similarity[cbind(seq_len(K), p)])
    })
    learned_to_true = perms[which.max(scores), ]
  } else {
    learned_to_true = rep(NA_integer_, K)
    available = seq_len(K)
    for (k in seq_len(K)) {
      best = available[which.max(similarity[k, available])]
      learned_to_true[k] = best
      available = setdiff(available, best)
    }
  }

  data.frame(
    learned_factor = seq_len(K),
    matched_true_factor = learned_to_true,
    cosine = similarity[cbind(seq_len(K), learned_to_true)]
  )
}

support_labels = function(gamma_bar, support_sizes, learned_to_true = NULL) {
  pip = apply(gamma_bar, c(1, 3), sum)
  K = ncol(pip)

  if (!is.null(learned_to_true)) {
    pip_matched = matrix(0, nrow = nrow(pip), ncol = K)
    for (k in seq_len(K)) {
      pip_matched[, learned_to_true[k]] = pip_matched[, learned_to_true[k]] +
        pip[, k]
    }
    pip = pip_matched
  }

  vapply(seq_len(nrow(pip)), function(i) {
    active = order(pip[i, ], decreasing = TRUE)[seq_len(support_sizes[i])]
    paste(sort(active), collapse = ",")
  }, character(1))
}

gamma_entropy = function(gamma_bar) {
  entropy = -gamma_bar * log(gamma_bar)
  entropy[is.nan(entropy)] = 0
  mean(rowSums(entropy, dims = 2)) / log(dim(gamma_bar)[3])
}

plot_elbo = function(fit, main) {
  idx = fit$elbo_iter
  plot(idx, fit$elbo[idx], type = "l", xlab = "Iteration", ylab = "ELBO",
       main = main)
  points(idx, fit$elbo[idx], pch = 19, cex = 0.5)
}

true_support_size = rowSums(L > 0)
true_support_label = apply(L > 0, 1, function(x) {
  paste(which(x), collapse = ",")
})
```

## Fit With True F0

As a baseline, we first fit the model with the true factor matrix fixed. This
isolates recovery of the posterior labels from the harder problem of learning
`F`.

```{r fit-true-F0}
fit_true_F0 = poisson_susie_nmf_fixed_F(
  Y = Y,
  F = F0,
  D = D,
  max_iters = 100,
  update_prior = TRUE,
  break_symmetry = FALSE,
  init_seed = 1,
  gamma_step_init = 0.1,
  gamma_step_ramp = 50
)
```

```{r elbo-true-F0}
plot_elbo(fit_true_F0, "ELBO with true F0")
```

```{r recovered-labels-true-F0}
top_k_true_F0 = matrix(0, N, D)
for (d in 1:D) {
  top_k_true_F0[, d] = apply(fit_true_F0$gamma_bar[, d, ], 1, which.max)
}

recovered_label_true_F0 = apply(top_k_true_F0, 1, function(x) {
  paste(sort(x), collapse = ",")
})

recovered_factor_mat_true_F0 = matrix(0, nrow = K, ncol = N)
for (i in 1:N) {
  recovered_factor_mat_true_F0[top_k_true_F0[i, ], i] =
    recovered_factor_mat_true_F0[top_k_true_F0[i, ], i] + 1
}

image(x = seq_len(N), y = seq_len(K), z = t(recovered_factor_mat_true_F0),
      axes = FALSE, xlab = "Sample", ylab = "Recovered factor",
      main = "Recovered factors with true F0",
      col = c("white", "firebrick", "darkred"))
axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
     labels = paste("G", 1:8))
axis(2, at = seq_len(K), labels = seq_len(K), las = 1)
abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
       col = "gray80", lty = 3)
box()
```

```{r recovered-table-true-F0}
recovered_by_group_true_F0 = do.call(rbind, lapply(1:8, function(g) {
  idx = which(grp == g)
  tab = table(recovered_label_true_F0[idx])
  data.frame(group = g, recovered_pattern = names(tab), n = as.integer(tab),
             row.names = NULL)
}))

recovered_by_group_true_F0
```

## Fit With Unknown F

The fit initializes `F` with Poisson NMF, initializes the row-level sparse
assignments from the Poisson-NMF loading matrix, then updates `F` inside the
variational loop with damping.

```{r fit}
fit = poisson_susie_nmf(
  Y = Y,
  K = K,
  D = D,
  max_iters = 100,
  update_prior = TRUE,
  break_symmetry = FALSE,
  init_seed = 1,
  init_F = "poisson_nmf",
  nmf_iters = 200,
  init_gamma_from_nmf = TRUE,
  gamma_step_init = 0.1,
  gamma_step_ramp = 50,
  F_step_init = 0.05,
  F_step_ramp = 75,
  elbo_every = 5,
  tol = 5e-4,
  min_iters = 20,
  patience = 2
)
```

```{r elbo}
plot_elbo(fit, "Variational objective")

data.frame(
  converged = fit$converged,
  n_iter = fit$n_iter,
  final_elbo = tail(fit$elbo[fit$elbo_iter], 1)
)
```

## Factor Recovery

Rows of the learned `F` are identifiable only up to permutation, so we first
match learned rows to true rows using cosine similarity.

```{r factor-recovery}
factor_match = match_factor_rows(fit$F, F0)
learned_to_true = factor_match$matched_true_factor

factor_match

data.frame(
  mean_matched_cosine = mean(factor_match$cosine),
  min_matched_cosine = min(factor_match$cosine)
)
```

## Recovery

For each observation and effect, we summarize the factor with the largest
posterior assignment probability, then remap learned factor labels to the
matched true labels.

```{r top-k}
top_k = matrix(0, N, D)
for (d in 1:D) {
  top_k[, d] = apply(fit$gamma_bar[, d, ], 1, which.max)
}
top_k_matched = matrix(learned_to_true[top_k], nrow = N, ncol = D)

recovered_by_group = do.call(rbind, lapply(1:8, function(g) {
  idx = which(grp == g)
  tab = table(apply(top_k_matched[idx, ], 1, function(x) {
    paste(sort(x), collapse = ",")
  }))
  data.frame(group = g, recovered_pattern = names(tab), n = as.integer(tab),
             row.names = NULL)
}))

recovered_by_group
```

```{r representative-posterior}
representative_idx = c(
  which(grp == 1)[1],
  which(grp == 2)[1],
  which(grp == 3)[1],
  which(grp == 4)[1],
  which(grp == 5)[1],
  which(grp == 6)[4],
  which(grp == 7)[1],
  which(grp == 8)[1]
)

representative_gamma = lapply(seq_along(representative_idx), function(j) {
  i = representative_idx[j]
  out = round(fit$gamma_bar[i, , ], 2)
  rownames(out) = paste0("effect_", seq_len(D))
  colnames(out) = paste0("factor_", seq_len(K))
  out
})
names(representative_gamma) = paste0("group_", 1:8)

representative_gamma
```

```{r recovered-factors-learned-F}
recovered_factor_mat = matrix(0, nrow = K, ncol = N)
for (i in 1:N) {
  recovered_factor_mat[top_k_matched[i, ], i] =
    recovered_factor_mat[top_k_matched[i, ], i] + 1
}

image(x = seq_len(N), y = seq_len(K), z = t(recovered_factor_mat),
      axes = FALSE, xlab = "Sample", ylab = "Recovered factor",
      main = "Recovered factors with learned F, matched to F0",
      col = c("white", "firebrick", "darkred"))
axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
     labels = paste("G", 1:8))
axis(2, at = seq_len(K), labels = seq_len(K), las = 1)
abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
       col = "gray80", lty = 3)
box()
```

```{r support-accuracy}
support_recovery = data.frame(
  fit = c("true_F0", "learned_F"),
  exact_support_accuracy = c(
    mean(support_labels(fit_true_F0$gamma_bar, true_support_size) ==
           true_support_label),
    mean(support_labels(fit$gamma_bar, true_support_size, learned_to_true) ==
           true_support_label)
  ),
  mean_normalized_gamma_entropy = c(
    gamma_entropy(fit_true_F0$gamma_bar),
    gamma_entropy(fit$gamma_bar)
  )
)

support_recovery
```

The support accuracy uses the true number of active factors for each simulated
observation and asks whether the top posterior inclusion scores recover that
support. This avoids penalizing one-factor observations only because the model
has `D = 3` possible effects.

## Takeaways

With the true dictionary fixed, the experiment isolates posterior support
recovery. With `F` learned, the same support task becomes coupled to factor
recovery and label matching. This page is therefore the bridge between
single-observation Poisson SuSiE and the later model-selection experiments:
once \(F\) is learned, overfitting \(K\) can create redundant factor rows, and
overfitting \(D\) can create extra single-effect dimensions.
