---
title: "5. MF Poisson SuSiE Tree Scenario"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This page creates a harder simulation for MF-Poisson-SuSiE. The true factor
matrix has `K_true = 7`, and every observation group uses at least two factors.
There are no single-factor anchor groups. The support patterns form a simple
tree: all groups share factor 1, then split into branch factors 2 and 5, and
some groups add one leaf factor.

We fit the model with the correct number of factors, `K_fit = 7`, but with an
overspecified number of effects, `D_fit = 5`.

The purpose of this page is to move beyond anchor-style simulations. If no
group is generated from a single factor, recovery must come from repeated
co-occurrence patterns rather than from easy one-factor examples.

```{r source-code}
source("code/joint-learn-susie-poi-F.R")
```

## Simulate Data

```{r simulate}
set.seed(47)

N = 1800
M = 2000
K_true = 7
K_fit = 7
D_true = 3
D_fit = 5

group_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)
n_groups = length(group_supports)
n_per_group = N / n_groups
grp = rep(seq_len(n_groups), each = n_per_group)

F0 = matrix(rgamma(M * K_true, shape = 0.1, rate = 0.01),
            nrow = K_true, ncol = M)
F0 = F0 / rowSums(F0)

L = matrix(0, nrow = N, ncol = K_true)
for (g in seq_len(n_groups)) {
  idx = which(grp == g)
  support = group_supports[[g]]
  L[idx, support] = matrix(rgamma(length(idx) * length(support),
                                  shape = 18, rate = 1),
                           nrow = length(idx), ncol = length(support))
}

rate = L %*% F0
Y = matrix(rpois(N * M, rate), nrow = N, ncol = M)

data.frame(
  N = N,
  M = M,
  K_true = K_true,
  K_fit = K_fit,
  D_true = D_true,
  D_fit = D_fit,
  total_count = sum(Y),
  nonzero_entries = sum(Y > 0)
)
```

```{r true-patterns}
true_patterns = vapply(group_supports, function(support) {
  paste(support, collapse = ",")
}, character(1))

data.frame(group = seq_len(n_groups), true_active_factors = true_patterns)
```

```{r true-factor-shapes}
matplot(t(F0[, 1:150]), type = "l", lty = 1,
        xlab = "m", ylab = "F0[k, m]",
        main = "True factor rows")
legend("topright", legend = paste("factor", seq_len(K_true)),
       col = seq_len(K_true), lty = 1, bty = "n")
```

## Diagnostics

```{r diagnostic-helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)
  perms = all_permutations(seq_len(K))
  scores = apply(perms, 1, function(p) {
    sum(similarity[cbind(seq_len(K), p)])
  })
  learned_to_true = perms[which.max(scores), ]

  data.frame(
    learned_factor = seq_len(K),
    matched_true_factor = learned_to_true,
    cosine = similarity[cbind(seq_len(K), learned_to_true)]
  )
}

loading_scores = function(fit) {
  E_lambda = fit$alpha / fit$beta
  apply(fit$gamma_bar * E_lambda, c(1, 3), sum)
}

posterior_inclusion_scores = function(fit) {
  1 - apply(1 - fit$gamma_bar, c(1, 3), prod)
}

effect_sizes = function(fit) {
  E_lambda = fit$alpha / fit$beta
  rowSums(fit$gamma_bar * E_lambda, dims = 2)
}

effect_gamma_entropy = function(fit) {
  entropy = -fit$gamma_bar * log(fit$gamma_bar)
  entropy[is.nan(entropy)] = 0
  rowSums(entropy, dims = 2) / log(dim(fit$gamma_bar)[3])
}

effect_max_gamma = function(fit) {
  apply(fit$gamma_bar, c(1, 2), max)
}

effect_top_factor = function(fit) {
  apply(fit$gamma_bar, c(1, 2), which.max)
}

rank_by_effect_size = function(effect_mat, size_mat) {
  out = matrix(0, nrow = nrow(effect_mat), ncol = ncol(effect_mat))
  for (i in seq_len(nrow(effect_mat))) {
    ord = order(size_mat[i, ], decreasing = TRUE)
    out[i, ] = effect_mat[i, ord]
  }
  out
}

rank_gamma_bar_by_size = function(fit, size_mat) {
  N = dim(fit$gamma_bar)[1]
  D = dim(fit$gamma_bar)[2]
  K = dim(fit$gamma_bar)[3]
  out = array(0, dim = c(N, D, K))
  for (i in seq_len(N)) {
    ord = order(size_mat[i, ], decreasing = TRUE)
    out[i, , ] = fit$gamma_bar[i, ord, ]
  }
  out
}

map_scores_to_true = function(scores, learned_to_true) {
  K = length(learned_to_true)
  out = matrix(0, nrow = nrow(scores), ncol = K)
  for (k in seq_len(K)) {
    out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
  }
  out
}

support_labels_from_scores = function(scores, support_sizes) {
  vapply(seq_len(nrow(scores)), function(i) {
    active = order(scores[i, ], decreasing = TRUE)[seq_len(support_sizes[i])]
    paste(sort(active), collapse = ",")
  }, character(1))
}

normalized_entropy = function(scores) {
  prob = scores / rowSums(scores)
  entropy = -prob * log(prob)
  entropy[is.nan(entropy)] = 0
  mean(rowSums(entropy)) / log(ncol(scores))
}

extra_effect_mass_fraction = function(size_mat, true_support_size) {
  vapply(seq_len(nrow(size_mat)), function(i) {
    sorted_sizes = sort(size_mat[i, ], decreasing = TRUE)
    extra_idx = seq.int(true_support_size[i] + 1, length(sorted_sizes))
    sum(sorted_sizes[extra_idx]) / sum(sorted_sizes)
  }, numeric(1))
}

plot_cosine_heatmap = function(cosine_mat, main) {
  image(x = seq_len(ncol(cosine_mat)), y = seq_len(nrow(cosine_mat)),
        z = t(cosine_mat), axes = FALSE, xlab = "True factor",
        ylab = "Fitted factor", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(cosine_mat)),
       labels = paste0("true ", seq_len(ncol(cosine_mat))))
  axis(2, at = seq_len(nrow(cosine_mat)),
       labels = paste0("fit ", seq_len(nrow(cosine_mat))), las = 1)
  for (k_fit in seq_len(nrow(cosine_mat))) {
    for (k_true in seq_len(ncol(cosine_mat))) {
      text(k_true, k_fit, sprintf("%.2f", cosine_mat[k_fit, k_true]),
           cex = 0.65)
    }
  }
  box()
}

plot_recovered_support = function(scores_true_space, main) {
  recovered = matrix(0, nrow = K_true, ncol = N)
  for (i in seq_len(N)) {
    active = order(scores_true_space[i, ], decreasing = TRUE)[
      seq_len(true_support_size[i])
    ]
    recovered[active, i] = 1
  }

  image(x = seq_len(N), y = seq_len(K_true), z = t(recovered),
        axes = FALSE, xlab = "Sample", ylab = "Recovered true factor",
        main = main, col = c("white", "firebrick"))
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", seq_len(n_groups)))
  axis(2, at = seq_len(K_true), labels = seq_len(K_true), las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  box()
}

plot_group_factor_heatmap = function(group_factor_mat, main, ylab) {
  image(x = seq_len(ncol(group_factor_mat)), y = seq_len(nrow(group_factor_mat)),
        z = t(group_factor_mat), axes = FALSE, xlab = "True factor",
        ylab = ylab, main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(group_factor_mat)),
       labels = seq_len(ncol(group_factor_mat)))
  axis(2, at = seq_len(nrow(group_factor_mat)),
       labels = paste("G", seq_len(nrow(group_factor_mat))), las = 1)
  for (g in seq_len(nrow(group_factor_mat))) {
    for (k in seq_len(ncol(group_factor_mat))) {
      text(k, g, sprintf("%.2f", group_factor_mat[g, k]), cex = 0.65)
    }
  }
  box()
}

plot_rank_heatmap = function(rank_mat, main, ylab, col) {
  image(x = seq_len(N), y = seq_len(ncol(rank_mat)), z = rank_mat,
        axes = FALSE, xlab = "Sample", ylab = ylab,
        main = main, col = col)
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", seq_len(n_groups)))
  axis(2, at = seq_len(ncol(rank_mat)), labels = seq_len(ncol(rank_mat)),
       las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  box()
}

plot_factor_id_heatmap = function(rank_top_factor, main) {
  image(x = seq_len(N), y = seq_len(ncol(rank_top_factor)),
        z = rank_top_factor, axes = FALSE, xlab = "Sample",
        ylab = "Effect rank", main = main, col = seq_len(K_fit))
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", seq_len(n_groups)))
  axis(2, at = seq_len(ncol(rank_top_factor)),
       labels = seq_len(ncol(rank_top_factor)), las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  legend("topright", legend = paste("factor", seq_len(K_fit)),
         fill = seq_len(K_fit), bty = "n", cex = 0.8)
  box()
}

plot_single_gamma_heatmap = function(gamma_mat, main) {
  D = nrow(gamma_mat)
  K = ncol(gamma_mat)
  image(x = seq_len(K), y = seq_len(D), z = t(gamma_mat),
        axes = FALSE, xlab = "Fitted factor", ylab = "Effect rank",
        main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(K), labels = seq_len(K))
  axis(2, at = seq_len(D), labels = seq_len(D), las = 1)
  box()
}

plot_representative_gamma = function(rank_gamma_bar, representative_idx) {
  op = par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
  on.exit(par(op))
  for (j in seq_along(representative_idx)) {
    i = representative_idx[j]
    plot_single_gamma_heatmap(
      rank_gamma_bar[i, , ],
      paste("Group", grp[i], "sample", i)
    )
  }
}

true_support_size = rowSums(L > 0)
true_support_label = apply(L > 0, 1, function(x) {
  paste(which(x), collapse = ",")
})
```

## Fit Tree Scenario

```{r fit}
fit = poisson_susie_nmf(
  Y = Y,
  K = K_fit,
  D = D_fit,
  max_iters = 100,
  update_prior = TRUE,
  break_symmetry = FALSE,
  init_seed = 1,
  init_F = "poisson_nmf",
  nmf_iters = 200,
  init_gamma_from_nmf = TRUE,
  gamma_step_init = 0.1,
  gamma_step_ramp = 50,
  F_step_init = 0.05,
  F_step_ramp = 75,
  elbo_every = 5,
  tol = 5e-4,
  min_iters = 20,
  patience = 2
)
```

```{r elbo}
idx = fit$elbo_iter
plot(idx, fit$elbo[idx], type = "l", xlab = "Iteration", ylab = "ELBO",
     main = "Variational objective")
points(idx, fit$elbo[idx], pch = 19, cex = 0.5)

data.frame(
  converged = fit$converged,
  n_iter = fit$n_iter,
  final_elbo = tail(fit$elbo[fit$elbo_iter], 1)
)
```

## Factor Recovery

```{r factor-recovery}
factor_match = match_factor_rows(fit$F, F0)
learned_to_true = factor_match$matched_true_factor

factor_match

data.frame(
  mean_matched_cosine = mean(factor_match$cosine),
  min_matched_cosine = min(factor_match$cosine)
)
```

```{r factor-cosine-heatmap}
cosine_fit_true = row_cosine(fit$F, F0)
plot_cosine_heatmap(
  cosine_fit_true,
  "Cosine similarity between fitted and true factors"
)
```

```{r learned-factor-shapes}
matplot(t(fit$F[, 1:150]), type = "l", lty = 1,
        xlab = "m", ylab = "F[k, m]",
        main = "Learned factor rows")
legend("topright", legend = paste("fit", seq_len(K_fit)),
       col = seq_len(K_fit), lty = 1, bty = "n")
```

## Support Recovery

```{r support-recovery}
fit_loading_scores = loading_scores(fit)
fit_pip_scores = posterior_inclusion_scores(fit)
scores_true_space = map_scores_to_true(fit_loading_scores, learned_to_true)
pip_true_space = map_scores_to_true(fit_pip_scores, learned_to_true)

sizes = effect_sizes(fit)
extra_mass = extra_effect_mass_fraction(sizes, true_support_size)

support_recovery = data.frame(
  exact_support_accuracy = mean(
    support_labels_from_scores(scores_true_space, true_support_size) ==
      true_support_label
  ),
  mean_normalized_loading_entropy = normalized_entropy(scores_true_space),
  mean_extra_effect_mass_fraction = mean(extra_mass),
  median_extra_effect_mass_fraction = median(extra_mass),
  q90_extra_effect_mass_fraction = as.numeric(quantile(extra_mass, 0.9))
)

support_recovery
```

```{r recovered-support-plot}
plot_recovered_support(
  scores_true_space,
  "Recovered supports in the tree scenario"
)
```

```{r recovered-patterns-by-group}
recovered_label = support_labels_from_scores(scores_true_space,
                                             true_support_size)
recovered_by_group = do.call(rbind, lapply(seq_len(n_groups), function(g) {
  tab = table(recovered_label[grp == g])
  data.frame(group = g, recovered_pattern = names(tab), n = as.integer(tab),
             row.names = NULL)
}))

recovered_by_group
```

## Group-Level Loading Structure

These heatmaps ask whether the fitted model recovers the tree structure after
matching learned factors to true factors. The first uses expected loading mass;
the second uses PIPs derived from `gamma_bar`.

```{r group-loading-heatmap}
group_loading_mean = do.call(rbind, lapply(seq_len(n_groups), function(g) {
  colMeans(scores_true_space[grp == g, , drop = FALSE])
}))
group_loading_share = group_loading_mean / rowSums(group_loading_mean)

plot_group_factor_heatmap(
  group_loading_share,
  "Group mean posterior loading share",
  "Group"
)
```

```{r group-pip-heatmap}
group_pip_mean = do.call(rbind, lapply(seq_len(n_groups), function(g) {
  colMeans(pip_true_space[grp == g, , drop = FALSE])
}))
group_pip_share = group_pip_mean / rowSums(group_pip_mean)

plot_group_factor_heatmap(
  group_pip_share,
  "Group mean posterior PIP share",
  "Group"
)
```

```{r branch-contrast}
branch_scores = data.frame(
  group = seq_len(n_groups),
  true_pattern = true_patterns,
  root_1 = group_loading_share[, 1],
  branch_2 = group_loading_share[, 2],
  branch_5 = group_loading_share[, 5],
  leaf_3 = group_loading_share[, 3],
  leaf_4 = group_loading_share[, 4],
  leaf_6 = group_loading_share[, 6],
  leaf_7 = group_loading_share[, 7]
)

branch_scores
```

The hardest separations are expected to be between sibling groups: groups 1 and
2 share factors 1 and 2 and differ only by factor 3 versus 4; groups 3 and 4
share factors 1 and 5 and differ only by factor 6 versus 7. Groups 5 and 6 test
whether the model can recover branch-level supports without any leaf factor.

## Overfitted Effect Behavior

```{r effect-rank-diagnostics}
gamma_entropy = effect_gamma_entropy(fit)
max_gamma = effect_max_gamma(fit)
top_factor = effect_top_factor(fit)
rank_gamma_bar = rank_gamma_bar_by_size(fit, sizes)

rank_size = rank_by_effect_size(sizes, sizes)
rank_entropy = rank_by_effect_size(gamma_entropy, sizes)
rank_max_gamma = rank_by_effect_size(max_gamma, sizes)
rank_top_factor = rank_by_effect_size(top_factor, sizes)

rank_summary = do.call(rbind, lapply(seq_len(D_fit), function(r) {
  data.frame(
    effect_rank = r,
    mean_expected_size = mean(rank_size[, r]),
    median_expected_size = median(rank_size[, r]),
    mean_gamma_entropy = mean(rank_entropy[, r]),
    median_gamma_entropy = median(rank_entropy[, r]),
    mean_max_gamma = mean(rank_max_gamma[, r]),
    median_max_gamma = median(rank_max_gamma[, r]),
    selected_as_extra_fraction = mean(r > true_support_size)
  )
}))

rank_summary
```

```{r effect-rank-size-boxplot}
boxplot(
  split(as.vector(rank_size), col(rank_size)),
  xlab = "Effect rank within observation",
  ylab = "Posterior expected effect size",
  main = "Effect sizes after fitting D = 5"
)
abline(v = D_true + 0.5, col = "gray50", lty = 2)
```

```{r effect-rank-entropy-boxplot}
boxplot(
  split(as.vector(rank_entropy), col(rank_entropy)),
  xlab = "Effect rank within observation",
  ylab = "Normalized entropy of gamma_bar",
  main = "Posterior factor-assignment uncertainty by effect rank"
)
abline(v = D_true + 0.5, col = "gray50", lty = 2)
```

```{r effect-rank-size-entropy}
plot(
  as.vector(rank_size),
  as.vector(rank_entropy),
  pch = 20,
  col = rgb(0.6, 0, 0, 0.25),
  xlab = "Posterior expected effect size",
  ylab = "Normalized entropy of gamma_bar",
  main = "Effect size versus gamma_bar uncertainty"
)
```

```{r effect-rank-size-heatmap}
plot_rank_heatmap(
  rank_size,
  "Posterior expected effect size by rank",
  "Effect rank",
  hcl.colors(100, "YlOrRd")
)
```

```{r effect-rank-entropy-heatmap}
plot_rank_heatmap(
  rank_entropy,
  "Normalized gamma_bar entropy by effect rank",
  "Effect rank",
  hcl.colors(100, "YlGnBu")
)
```

```{r effect-rank-top-factor-heatmap}
plot_factor_id_heatmap(
  rank_top_factor,
  "Top posterior fitted factor by effect rank"
)
```

## Representative Posterior Assignment Heatmaps

Rows are fitted effects sorted by posterior expected size. Columns are fitted
factors. Since `D_fit = 5` but the true support size is two or three, the lower
effect ranks should ideally be small and diffuse.

```{r representative-gamma-heatmaps, fig.width=10, fig.height=6}
representative_idx = vapply(seq_len(n_groups), function(g) {
  which(grp == g)[1]
}, integer(1))

plot_representative_gamma(rank_gamma_bar, representative_idx)
```

## Interpretation

This setting removes the easiest anchors: no observation group is generated from
a single factor. Recovery therefore depends on separating factors through shared
and contrasting co-occurrence patterns. Factor 1 is a root factor present in
every group; factors 2 and 5 are branch factors; factors 3, 4, 6, and 7 are leaf
factors that only appear in one group each.

For fixed learned `F`, `gamma_bar` quantifies posterior uncertainty in the
loading decomposition. In this tree scenario, diffuse `gamma_bar` in lower-ranked
effects indicates that the extra fitted dimensions are not confidently assigned
to a factor. Sizable lower-ranked effects with low entropy are more concerning:
they mean the overspecified `D` is acting as additional confident components,
which can blur the intended tree support patterns.

## Takeaways

This scenario is a stress test for support recovery without single-factor
anchors. It motivates the MiSo layer: the important structure is no longer
only which factors appear in each observation, but which sparsity patterns
repeat across groups. The posterior quantities from MF-Poisson-SuSiE remain
useful diagnostics, but the next modeling step should share motif-level
structure across observations rather than fitting each observation independently.
