---
title: "6. MiSo with Anchor and Tree Scenarios"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(stringsAsFactors = FALSE)
```

This page revisits the two support-pattern simulations introduced in Analyses
2 and 5, now using the full MiSo model. The first scenario contains
single-factor anchor motifs. The second has a tree structure and no
single-factor motif. MiSo learns the factor matrix in both cases. The fitted
numbers of factors $K$, motifs $S$, and maximum motif dimension $D$ are set
separately at the beginning of each scenario section.

```{r source-code}
source("code/miso-benchmark-utils.R")
```

## Simulation design

The anchor supports are the eight groups from Analysis 2. The tree supports
are the six groups from Analysis 5. The early pages used larger matrices; here
we keep the same structural scenarios but use $N=480$ observations and
$M=500$ features so the demonstration is easy to rerun.

```{r design}
anchor_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

scenario_design = data.frame(
  scenario = c("anchor", "tree/no-anchor"),
  N = c(480, 480),
  M = c(500, 500),
  K = c(5, 7),
  S = c(length(anchor_supports), length(tree_supports)),
  D = c(3, 3),
  support_sizes = c("1 or 3", "2 or 3")
)
scenario_design
```

Here $D=3$ is the true maximum dimension in both scenarios. Some individual
motifs have dimension smaller than three; MiSo still uses the common maximum
$D$, with the Gamma prior parameters available to diagnose inactive
dimensions.

```{r simulate}
anchor_data = simulate_support_scenario(
  N = 480,
  M = 500,
  K = 5,
  group_supports = anchor_supports,
  shape_vec = c(rep(10, 5), rep(20, 3)),
  seed = 42
)

tree_data = simulate_support_scenario(
  N = 480,
  M = 500,
  K = 7,
  group_supports = tree_supports,
  shape_vec = rep(18, length(tree_supports)),
  seed = 47
)

data.frame(
  scenario = c("anchor", "tree/no-anchor"),
  total_count = c(sum(anchor_data$Y), sum(tree_data$Y)),
  nonzero_entries = c(sum(anchor_data$Y > 0), sum(tree_data$Y > 0))
)
```

### Shared fitting helper

The factors are unknown to the fitting algorithm. Each scenario is fitted
three times from the same MF-Poisson-SuSiE preliminary fit logic but with different
motif initializations:

1. `distinct` selects the top $D$ center factors without replacement;
2. `threshold` retains center factors with at least 10% loading share and
   recycles those factors when fewer than $D$ are retained;
3. `aligned_mf` uses the top $D$ distinct factors as alignment anchors, aligns
   the exchangeable MF slots within each preliminary cluster, and initializes
   each motif slot from a loading-weighted average of the corresponding
   MF factor-selection posteriors.

Specifically, let $p_i(d)$ be the MF slot aligned to motif slot $d$ for
observation $i$, and let $w_{i,p_i(d)}$ be that slot's share of the
observation's expected total loading. For the observations in preliminary
cluster $C_s$, the new initialization is

$$
\bar\gamma^{(0)}_{sdk}
=
(1-\varepsilon_\gamma)
\frac{
  \sum_{i\in C_s}w_{i,p_i(d)}
  \bar\gamma^{\mathrm{MF}}_{i,p_i(d),k}
}{
  \sum_{i\in C_s}w_{i,p_i(d)}
}
+\frac{\varepsilon_\gamma}{K}.
$$

The one-to-one alignment maximizes the MF posterior expected loading assigned
to the distinct anchor factors. Normalizing the loading weights within each
observation prevents high-count observations from dominating the cluster
average.

The true factor matrix and support labels are used only after fitting to
evaluate recovery. Cache filenames are generated from the scenario, fitted
$K,S,D$, seed, and initialization, so changing a setting creates a new fit.

```{r fit-helper}
fit_miso_scenario = function(dat, scenario, K, S, D, seed,
                             motif_initialization) {
  motif_initialization = match.arg(
    motif_initialization, c("distinct", "threshold", "aligned_mf")
  )
  cache_dir = file.path("analysis", "cache", "6.miso-anchor-tree-v2")
  dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
  cache_name = sprintf(
    "%s-K%d-S%d-D%d-seed%d-%s-iter30.rds",
    scenario, K, S, D, seed, motif_initialization
  )
  cache_file = file.path(cache_dir, cache_name)

  if (file.exists(cache_file)) return(readRDS(cache_file))
  legacy_cache_file = file.path(
    "analysis", "cache", "6.miso-anchor-tree-v1",
    sprintf("%s-K%d-S%d-D%d-seed%d-iter30.rds", scenario, K, S, D, seed)
  )
  if (motif_initialization == "distinct" && file.exists(legacy_cache_file)) {
    return(readRDS(legacy_cache_file))
  }

  fit = miso(
    Y = dat$Y,
    K = K,
    S = S,
    D = D,
    max_iters = 30,
    n_inner = 3,
    mf_max_iters = 30,
    mf_nmf_iters = 60,
    init_seed = seed,
    update_prior = TRUE,
    update_F = TRUE,
    update_gamma = TRUE,
    update_slot_scale = FALSE,
    motif_initialization = motif_initialization,
    motif_min_share = 0.10,
    surplus_slots = "repeat",
    tol = 1e-5,
    min_iters = 5,
    patience = 2,
    block_size = 100
  )
  saveRDS(fit, cache_file)
  fit
}
```

### Recovery summaries

Factor and motif labels are permutation invariant. The following helpers match
learned factors to true factors by cosine similarity and match fitted motifs to
true groups by total posterior responsibility.

#### How the plotted quantities are computed

The plots use the following fitted variational quantities:

- `omega[n, s]` is the posterior probability that observation $n$ belongs to
  motif $s$, $\omega_{ns}=q(z_n=s)$.
- `gamma_bar[s, d, k]` is the posterior probability that dimension $d$ of
  motif $s$ selects factor $k$, $\bar\gamma_{sdk}=q(c_{sd}=k)$.
- `alpha[n, s, d]` and `beta_sd[s, d]` are the shape and rate of the
  variational Gamma distribution for the loading in dimension $d$. Therefore
  its posterior mean is $\alpha_{nsd}/\beta_{sd}$. Here $\beta$ is the Gamma
  rate; the object called `lambda_mean` inside `motif_factor_scores()` is an
  average *loading*, not a rate parameter.
- `slot_scale[s, d]`, denoted $\tau_{sd}$ below, is an optional multiplicative
  scale for that dimension. This page sets `update_slot_scale = FALSE`, so
  $\tau_{sd}=1$ for every motif and dimension.

For each motif and dimension, `motif_factor_scores()` first computes the
responsibility-weighted mean loading

$$
\bar\ell_{sd}
=
\frac{\sum_{n=1}^N \omega_{ns}
      \alpha_{nsd}/\beta_{sd}}
     {\sum_{n=1}^N \omega_{ns}}.
$$

It then aggregates the $D$ dimensions into the unnormalized motif--factor
strength

$$
A_{sk}=\sum_{d=1}^D
       \tau_{sd}\,\bar\ell_{sd}\,\bar\gamma_{sdk}.
$$

The fitted $S\times K$ heatmap displays the row-normalized matrix
$\widetilde A_{sk}=A_{sk}/\sum_{j=1}^K A_{sj}$. Thus a cell is the estimated
share of a motif's total loading strength carried by a factor. It is not merely
$\sum_d\bar\gamma_{sdk}$: dimensions with larger expected loadings contribute
more. Every row of the displayed matrix sums to one.
The implementation applies a $10^{-12}$ floor to denominators and fitted
strengths only to avoid numerical division by zero.

For comparison, if the true support of group $g$ is ${\cal A}_g$, the true
$S\times K$ map is

$$
T_{gk}=\frac{\mathbf{1}\{k\in {\cal A}_g\}}{|{\cal A}_g|}.
$$

This gives equal share to every truly active factor and also makes each row sum
to one. It records support membership, whereas the fitted map records estimated
loading strength; the comparison therefore emphasizes whether the fitted mass
falls on the correct support rather than requiring equal fitted weights.

Because both factor and motif labels are arbitrary, the display applies two
permutations after fitting. Learned factors are matched to true factors by the
one-to-one assignment that maximizes the sum of cosine similarities between
rows of the learned and true factor matrices. Fitted motifs are matched to true
groups by the one-to-one assignment that maximizes
$\sum_n\omega_{n,s(g)}\mathbf{1}\{g_n=g\}$. These steps only reorder rows or
columns for evaluation; the true labels are never supplied to `miso()`.

The posterior-responsibility heatmap displays

$$
R_{gh}=\frac{1}{N_g}\sum_{n:g_n=g}\omega_{n,s(h)},
$$

where $s(h)$ is the fitted motif aligned to true group $h$. Its rows sum to
one. A large diagonal entry means that observations from true group $g$ place
most of their posterior mass on the corresponding fitted motif; off-diagonal
mass shows uncertainty or merging between motifs. All heatmaps use the same
fixed color range from zero to one, and the printed cell values are these
matrices rounded to two decimal places.

```{r evaluation-helpers}
maximum_weight_assignment = function(score) {
  n = max(nrow(score), ncol(score))
  padded = matrix(0, nrow = n, ncol = n)
  padded[seq_len(nrow(score)), seq_len(ncol(score))] = score

  if (requireNamespace("clue", quietly = TRUE)) {
    return(as.integer(clue::solve_LSAP(padded, maximum = TRUE)))
  }

  assignment = rep(NA_integer_, n)
  available_rows = seq_len(n)
  available_columns = seq_len(n)
  for (j in seq_len(n)) {
    candidate = which(
      padded[available_rows, available_columns, drop = FALSE] ==
        max(padded[available_rows, available_columns, drop = FALSE]),
      arr.ind = TRUE
    )[1, ]
    row = available_rows[candidate[1]]
    column = available_columns[candidate[2]]
    assignment[row] = column
    available_rows = setdiff(available_rows, row)
    available_columns = setdiff(available_columns, column)
  }
  assignment
}

match_factor_rows = function(F_hat, F_true) {
  similarity = row_cosine(F_hat, F_true)
  assignment = maximum_weight_assignment(similarity)
  learned_to_true = rep(NA_integer_, nrow(F_hat))
  matched = assignment[seq_len(nrow(F_hat))]
  learned_to_true[matched <= nrow(F_true)] =
    matched[matched <= nrow(F_true)]

  unmatched = which(is.na(learned_to_true))
  if (length(unmatched) > 0) {
    learned_to_true[unmatched] = max.col(
      similarity[unmatched, , drop = FALSE], ties.method = "first"
    )
  }
  attr(learned_to_true, "true_K") = nrow(F_true)

  list(
    table = data.frame(
      learned_factor = seq_len(nrow(F_hat)),
      matched_true_factor = as.integer(learned_to_true),
      cosine = similarity[cbind(seq_len(nrow(F_hat)), learned_to_true)]
    ),
    learned_to_true = learned_to_true,
    similarity = similarity
  )
}

map_scores_to_true = function(scores, learned_to_true) {
  true_K = attr(learned_to_true, "true_K")
  if (is.null(true_K)) true_K = max(learned_to_true)
  out = matrix(0, nrow = nrow(scores), ncol = true_K)
  for (k in seq_along(learned_to_true)) {
    out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
  }
  out
}

match_motifs_to_groups = function(omega, true_group) {
  groups = sort(unique(true_group))
  S_fit = ncol(omega)

  score = vapply(groups, function(g) {
    colSums(omega[true_group == g, , drop = FALSE])
  }, numeric(S_fit))

  assignment = maximum_weight_assignment(score)
  fit_to_true = assignment[seq_len(S_fit)]
  unmatched_fits = which(fit_to_true > length(groups))
  if (length(unmatched_fits) > 0) {
    fit_to_true[unmatched_fits] = max.col(
      score[unmatched_fits, , drop = FALSE], ties.method = "first"
    )
  }

  true_to_fit = match(seq_along(groups), fit_to_true)
  unmatched_groups = which(is.na(true_to_fit))
  if (length(unmatched_groups) > 0) {
    true_to_fit[unmatched_groups] = vapply(unmatched_groups, function(g) {
      which.max(score[, g])
    }, integer(1))
  }

  list(
    fit_to_true = fit_to_true,
    true_to_fit = true_to_fit,
    score = score
  )
}

true_motif_matrix = function(supports, K) {
  # T[g, k] = 1 / |A_g| for a factor in the true support A_g, and zero otherwise.
  out = matrix(0, nrow = length(supports), ncol = K)
  for (s in seq_along(supports)) {
    out[s, supports[[s]]] = 1 / length(supports[[s]])
  }
  out
}

distinct_slot_winners = function(gamma_bar) {
  winners = apply(gamma_bar, c(1, 2), which.max)
  apply(winners, 1, function(x) length(unique(x)))
}

evaluate_miso = function(fit, dat, supports, scenario) {
  # Match learned labels to truth only for evaluation and plotting.
  factor_match = match_factor_rows(fit$F, dat$F0)
  motif_match = match_motifs_to_groups(fit$omega, dat$grp)

  # Observation-level expected factor loadings are used for support metrics.
  observation_scores = miso_observation_factor_scores(fit)

  # A[s, k] = sum_d tau[s, d] * ell_bar[s, d] * gamma_bar[s, d, k].
  motif_scores = motif_factor_scores(fit)$scores
  # Reorder factor columns into true-factor order, normalize each motif to sum 1,
  # and reorder motif rows into true-group order.
  motif_scores = map_scores_to_true(
    motif_scores, factor_match$learned_to_true
  )
  motif_scores = normalize_scores(motif_scores)
  motif_scores = motif_scores[motif_match$true_to_fit, , drop = FALSE]

  # R[g, s] is the mean posterior responsibility for fitted motif s among
  # observations generated from true group g.
  group_responsibility = do.call(rbind, lapply(seq_along(supports), function(g) {
    colMeans(fit$omega[dat$grp == g, , drop = FALSE])
  }))
  group_responsibility = group_responsibility[
    , motif_match$true_to_fit, drop = FALSE
  ]

  soft_mass = mean(fit$omega[cbind(
    seq_len(nrow(fit$omega)), motif_match$true_to_fit[dat$grp]
  )])

  metrics = data.frame(
    scenario = scenario,
    true_K = nrow(dat$F0),
    true_S = length(supports),
    true_D = max(lengths(supports)),
    fitted_K = nrow(fit$F),
    fitted_S = ncol(fit$omega),
    fitted_D = dim(fit$gamma_bar)[2],
    median_initial_distinct_factors = median(
      distinct_slot_winners(fit$gamma_init$gamma_bar)
    ),
    median_final_distinct_factors = median(
      distinct_slot_winners(fit$gamma_bar)
    ),
    mean_initial_gamma_entropy = mean(
      motif_gamma_entropy(fit$gamma_init$gamma_bar)
    ),
    mean_initial_top_gamma = mean(
      apply(fit$gamma_init$gamma_bar, c(1, 2), max)
    ),
    soft_motif_accuracy = soft_mass,
    exact_support_accuracy = support_accuracy(
      observation_scores, factor_match$learned_to_true, dat
    ),
    support_jaccard = support_jaccard(
      observation_scores, factor_match$learned_to_true, dat
    ),
    mean_factor_cosine = mean(factor_match$table$cosine),
    mean_assignment_certainty = mean(apply(fit$omega, 1, max))
  )

  motif_summary = do.call(rbind, lapply(seq_along(supports), function(g) {
    n_active = length(supports[[g]])
    estimated = order(motif_scores[g, ], decreasing = TRUE)[seq_len(n_active)]
    data.frame(
      true_group = paste0("G", g),
      matched_fitted_motif = paste0("S", motif_match$true_to_fit[g]),
      true_factors = support_label(supports[[g]]),
      estimated_top_factors = support_label(estimated),
      posterior_mass_on_true_factors = round(
        sum(motif_scores[g, supports[[g]]]), 3
      )
    )
  }))

  list(
    metrics = metrics,
    factor_match = factor_match,
    motif_match = motif_match,
    group_responsibility = group_responsibility,
    true_motif_share = true_motif_matrix(supports, nrow(dat$F0)),
    fitted_motif_share = motif_scores,
    motif_summary = motif_summary
  )
}

plot_heatmap = function(mat, main, xlab, ylab, xlabels, ylabels,
                        digits = 2) {
  # Every displayed matrix is on [0, 1]; text reports its cell values directly.
  image(
    x = seq_len(ncol(mat)), y = seq_len(nrow(mat)), z = t(mat),
    axes = FALSE, xlab = xlab, ylab = ylab, main = main,
    zlim = c(0, 1), col = hcl.colors(100, "YlOrRd")
  )
  axis(1, at = seq_len(ncol(mat)), labels = xlabels)
  axis(2, at = seq_len(nrow(mat)), labels = ylabels, las = 1)
  for (i in seq_len(nrow(mat))) {
    for (j in seq_len(ncol(mat))) {
      text(j, i, sprintf(paste0("%.", digits, "f"), mat[i, j]), cex = 0.65)
    }
  }
  box()
}

plot_scenario = function(result, scenario) {
  S = nrow(result$true_motif_share)
  K = ncol(result$true_motif_share)
  group_labels = paste0("G", seq_len(S))
  factor_labels = paste0("F", seq_len(K))

  op = par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  on.exit(par(op))
  plot_heatmap(
    result$true_motif_share,
    paste(scenario, "true motif-factor map"),
    "True factor", "True group", factor_labels, group_labels
  )
  plot_heatmap(
    result$fitted_motif_share,
    paste(scenario, "fitted motif-factor map"),
    "Matched true factor", "Matched true group", factor_labels, group_labels
  )
}
```

## Anchor scenario

The anchor scenario contains five one-factor motifs and three three-factor
motifs. It checks whether MiSo can learn elementary factor directions and reuse
them inside composite motifs.

### Choose the fitted dimensions

Edit the first three values in the following chunk to choose the fitted
$K,S,D$ for the anchor analysis. The defaults are the true values. The
`distinct` and `aligned_mf` require
`ANCHOR_FIT_D <= ANCHOR_FIT_K`.

```{r anchor-fit-settings}
ANCHOR_FIT_K = 5
ANCHOR_FIT_S = 8
ANCHOR_FIT_D = 3
ANCHOR_FIT_SEED = 1

data.frame(
  fitted_K = ANCHOR_FIT_K,
  fitted_S = ANCHOR_FIT_S,
  fitted_D = ANCHOR_FIT_D,
  seed = ANCHOR_FIT_SEED
)
```

### Fit with three initializations

```{r anchor-fit-models}
anchor_fit_distinct = fit_miso_scenario(
  anchor_data,
  scenario = "anchor",
  K = ANCHOR_FIT_K,
  S = ANCHOR_FIT_S,
  D = ANCHOR_FIT_D,
  seed = ANCHOR_FIT_SEED,
  motif_initialization = "distinct"
)

anchor_fit_threshold = fit_miso_scenario(
  anchor_data,
  scenario = "anchor",
  K = ANCHOR_FIT_K,
  S = ANCHOR_FIT_S,
  D = ANCHOR_FIT_D,
  seed = ANCHOR_FIT_SEED,
  motif_initialization = "threshold"
)

anchor_fit_aligned_mf = fit_miso_scenario(
  anchor_data,
  scenario = "anchor",
  K = ANCHOR_FIT_K,
  S = ANCHOR_FIT_S,
  D = ANCHOR_FIT_D,
  seed = ANCHOR_FIT_SEED,
  motif_initialization = "aligned_mf"
)

stopifnot(
  all(dim(anchor_fit_distinct$gamma_bar) ==
        c(ANCHOR_FIT_S, ANCHOR_FIT_D, ANCHOR_FIT_K)),
  all(dim(anchor_fit_threshold$gamma_bar) ==
        c(ANCHOR_FIT_S, ANCHOR_FIT_D, ANCHOR_FIT_K)),
  all(dim(anchor_fit_aligned_mf$gamma_bar) ==
        c(ANCHOR_FIT_S, ANCHOR_FIT_D, ANCHOR_FIT_K))
)
```

### Compare recovery

```{r anchor-evaluate}
anchor_result_distinct = evaluate_miso(
  anchor_fit_distinct, anchor_data, anchor_supports, "anchor"
)
anchor_result_threshold = evaluate_miso(
  anchor_fit_threshold, anchor_data, anchor_supports, "anchor"
)
anchor_result_aligned_mf = evaluate_miso(
  anchor_fit_aligned_mf, anchor_data, anchor_supports, "anchor"
)

anchor_recovery_summary = rbind(
  transform(anchor_result_distinct$metrics, initialization = "distinct"),
  transform(anchor_result_threshold$metrics,
            initialization = "threshold/recycle"),
  transform(anchor_result_aligned_mf$metrics,
            initialization = "aligned MF posterior")
)
numeric_columns = vapply(anchor_recovery_summary, is.numeric, logical(1))
anchor_recovery_summary[numeric_columns] = round(
  anchor_recovery_summary[numeric_columns], 3
)
anchor_recovery_summary
```

```{r anchor-motif-summary}
rbind(
  cbind(initialization = "distinct",
        anchor_result_distinct$motif_summary),
  cbind(initialization = "threshold/recycle",
        anchor_result_threshold$motif_summary),
  cbind(initialization = "aligned MF posterior",
        anchor_result_aligned_mf$motif_summary)
)
```

```{r anchor-factor-maps, fig.height=4.5}
plot_scenario(anchor_result_distinct, "Anchor, distinct initialization")
plot_scenario(anchor_result_threshold, "Anchor, threshold/recycle initialization")
plot_scenario(anchor_result_aligned_mf, "Anchor, aligned MF initialization")
```

```{r anchor-responsibilities, fig.width=6, fig.height=5, fig.show='hold'}
plot_heatmap(
  anchor_result_distinct$group_responsibility,
  "Anchor, distinct: mean posterior motif responsibility",
  "Fitted motif, aligned to group", "True group",
  paste0("G", seq_len(ncol(anchor_result_distinct$group_responsibility))),
  paste0("G", seq_len(nrow(anchor_result_distinct$group_responsibility)))
)
plot_heatmap(
  anchor_result_threshold$group_responsibility,
  "Anchor, threshold/recycle: mean posterior motif responsibility",
  "Fitted motif, aligned to group", "True group",
  paste0("G", seq_len(ncol(anchor_result_threshold$group_responsibility))),
  paste0("G", seq_len(nrow(anchor_result_threshold$group_responsibility)))
)
plot_heatmap(
  anchor_result_aligned_mf$group_responsibility,
  "Anchor, aligned MF: mean posterior motif responsibility",
  "Fitted motif, aligned to group", "True group",
  paste0("G", seq_len(ncol(anchor_result_aligned_mf$group_responsibility))),
  paste0("G", seq_len(nrow(anchor_result_aligned_mf$group_responsibility)))
)
```

The diagonal responsibility pattern checks motif recovery directly, while the
factor maps separate dictionary recovery from motif recovery. The fitted map is
shown after label matching only; the true labels were not used during fitting.

## Tree scenario

Every tree motif contains factor 1 and then follows one of two branches,
through factor 2 or factor 5. Four motifs add a branch-specific leaf. Because
there are no one-factor groups, recovery must use repeated co-occurrence across
motifs rather than isolated anchors.

### Choose the fitted dimensions

Edit the first three values in the following chunk to choose the fitted
$K,S,D$ for the tree analysis. The defaults are the true values. The
`distinct` and `aligned_mf` require `TREE_FIT_D <= TREE_FIT_K`.

```{r tree-fit-settings}
TREE_FIT_K = 7
TREE_FIT_S = 6
TREE_FIT_D = 3
TREE_FIT_SEED = 2

data.frame(
  fitted_K = TREE_FIT_K,
  fitted_S = TREE_FIT_S,
  fitted_D = TREE_FIT_D,
  seed = TREE_FIT_SEED
)
```

### Fit with three initializations

```{r tree-fit-models}
tree_fit_distinct = fit_miso_scenario(
  tree_data,
  scenario = "tree",
  K = TREE_FIT_K,
  S = TREE_FIT_S,
  D = TREE_FIT_D,
  seed = TREE_FIT_SEED,
  motif_initialization = "distinct"
)

tree_fit_threshold = fit_miso_scenario(
  tree_data,
  scenario = "tree",
  K = TREE_FIT_K,
  S = TREE_FIT_S,
  D = TREE_FIT_D,
  seed = TREE_FIT_SEED,
  motif_initialization = "threshold"
)

tree_fit_aligned_mf = fit_miso_scenario(
  tree_data,
  scenario = "tree",
  K = TREE_FIT_K,
  S = TREE_FIT_S,
  D = TREE_FIT_D,
  seed = TREE_FIT_SEED,
  motif_initialization = "aligned_mf"
)

stopifnot(
  all(dim(tree_fit_distinct$gamma_bar) ==
        c(TREE_FIT_S, TREE_FIT_D, TREE_FIT_K)),
  all(dim(tree_fit_threshold$gamma_bar) ==
        c(TREE_FIT_S, TREE_FIT_D, TREE_FIT_K)),
  all(dim(tree_fit_aligned_mf$gamma_bar) ==
        c(TREE_FIT_S, TREE_FIT_D, TREE_FIT_K))
)
```

### Compare recovery

```{r tree-evaluate}
tree_result_distinct = evaluate_miso(
  tree_fit_distinct, tree_data, tree_supports, "tree/no-anchor"
)
tree_result_threshold = evaluate_miso(
  tree_fit_threshold, tree_data, tree_supports, "tree/no-anchor"
)
tree_result_aligned_mf = evaluate_miso(
  tree_fit_aligned_mf, tree_data, tree_supports, "tree/no-anchor"
)

tree_recovery_summary = rbind(
  transform(tree_result_distinct$metrics, initialization = "distinct"),
  transform(tree_result_threshold$metrics,
            initialization = "threshold/recycle"),
  transform(tree_result_aligned_mf$metrics,
            initialization = "aligned MF posterior")
)
numeric_columns = vapply(tree_recovery_summary, is.numeric, logical(1))
tree_recovery_summary[numeric_columns] = round(
  tree_recovery_summary[numeric_columns], 3
)
tree_recovery_summary
```

```{r tree-motif-summary}
rbind(
  cbind(initialization = "distinct",
        tree_result_distinct$motif_summary),
  cbind(initialization = "threshold/recycle",
        tree_result_threshold$motif_summary),
  cbind(initialization = "aligned MF posterior",
        tree_result_aligned_mf$motif_summary)
)
```

```{r tree-factor-maps, fig.height=4.5}
plot_scenario(tree_result_distinct, "Tree, distinct initialization")
plot_scenario(tree_result_threshold, "Tree, threshold/recycle initialization")
plot_scenario(tree_result_aligned_mf, "Tree, aligned MF initialization")
```

```{r tree-responsibilities, fig.width=6, fig.height=5, fig.show='hold'}
plot_heatmap(
  tree_result_distinct$group_responsibility,
  "Tree, distinct: mean posterior motif responsibility",
  "Fitted motif, aligned to group", "True group",
  paste0("G", seq_len(ncol(tree_result_distinct$group_responsibility))),
  paste0("G", seq_len(nrow(tree_result_distinct$group_responsibility)))
)
plot_heatmap(
  tree_result_threshold$group_responsibility,
  "Tree, threshold/recycle: mean posterior motif responsibility",
  "Fitted motif, aligned to group", "True group",
  paste0("G", seq_len(ncol(tree_result_threshold$group_responsibility))),
  paste0("G", seq_len(nrow(tree_result_threshold$group_responsibility)))
)
plot_heatmap(
  tree_result_aligned_mf$group_responsibility,
  "Tree, aligned MF: mean posterior motif responsibility",
  "Fitted motif, aligned to group", "True group",
  paste0("G", seq_len(ncol(tree_result_aligned_mf$group_responsibility))),
  paste0("G", seq_len(nrow(tree_result_aligned_mf$group_responsibility)))
)
```

### Interpretation across scenarios

The anchor scenario is the easier sanity check: the singleton motifs expose
each factor separately, and the composite motifs test whether those parts are
recombined correctly. The tree scenario is more diagnostic of MiSo because its
factors are identifiable only through shared motif structure. Within each
scenario, the three fits hold $K,S,D$ and the random seed fixed, so differences
in recovery can be attributed to the motif initialization. Changing the
scenario-specific settings above makes it possible to study under- or
over-specification without modifying the fitting and evaluation code.

In the default runs, `aligned_mf` begins with three distinct factor directions
per motif in both scenarios. In the anchor scenario it attains the same high
soft motif accuracy as threshold/recycle, while its median final number of
distinct directions is two rather than one. In the tree scenario it retains a
median of three distinct directions and combines high motif accuracy with the
best support recovery of the three initializations. This is the intended
compromise: retain the dimensional coverage of distinct initialization while
using the uncertainty already learned by MF-Poisson-SuSiE.
