---
title: "Joint Learning of F in Poisson SuSiE"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This page turns the simulation at the end of
`code/joint-learn-susie-poi-F.R` into a reproducible workflowr analysis. The
experiment fits matrix-factorization Poisson SuSiE while updating the shared
factor matrix `F`.

```{r source-code}
source("code/joint-learn-susie-poi-F.R")
```

## Fitted Model

For a count matrix \(Y \in \mathbb{N}^{N \times M}\), the fitted model uses a
shared nonnegative factor matrix \(F \in \mathbb{R}_+^{K \times M}\) and a
small number \(D\) of single-effect components per observation:
\[
  Y_{im} \sim \mathrm{Poisson}\left(
    \sum_{d=1}^D \beta_{id} F_{\gamma_{id}m}
  \right).
\]
Each effect chooses one factor row through
\(\gamma_{id} \in \{1,\ldots,K\}\), and its size is \(\beta_{id} > 0\).
The variational fit estimates posterior assignment probabilities
`gamma_bar[i, d, k]`, Gamma variational parameters for the effect sizes, and,
when `update_F = TRUE`, the shared factor matrix `F`.

The simulation below first checks recovery when the true factor matrix `F0` is
given to the algorithm, then repeats the fit from a random initialized `F` while
learning `F` jointly.

## Simulate Data

We create eight observation groups. Groups 1 through 5 each use one factor, and
groups 6 through 8 use three-factor combinations.

```{r simulate}
set.seed(42)

N = 1600
M = 2000
K = 5
D = 3

F0 = matrix(rgamma(M * K, shape = 0.1, rate = 0.01), nrow = K, ncol = M)
F0 = F0 / rowSums(F0)

n_per_group = N / 8
L = matrix(0, nrow = N, ncol = K)
grp = rep(1:8, each = n_per_group)

for (g in 1:5) {
  L[grp == g, g] = matrix(rgamma(n_per_group, 10, 1), n_per_group, 1)
}

L[grp == 6, c(1, 2, 3)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 7, c(1, 2, 4)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 8, c(2, 3, 5)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)

rate = L %*% F0
Y = matrix(rpois(N * M, rate), nrow = N, ncol = M)

data.frame(
  N = N,
  M = M,
  K = K,
  D = D,
  total_count = sum(Y),
  nonzero_entries = sum(Y > 0)
)
```

```{r true-patterns}
true_patterns = vapply(1:8, function(g) {
  paste(which(colSums(L[grp == g, , drop = FALSE]) > 0), collapse = ",")
}, character(1))

data.frame(group = 1:8, true_active_factors = true_patterns)
```

```{r true-factor-shapes}
matplot(t(F0[, 1:150]), type = "l", lty = 1,
        xlab = "m", ylab = "F0[k, m]",
        main = "True factor rows")
legend("topright", legend = paste("factor", 1:K), col = 1:K, lty = 1,
       bty = "n")
```

## Fit With True F0

As a baseline, we first fit the model with the true factor matrix fixed. This
isolates recovery of the posterior labels from the harder problem of learning
`F`.

```{r fit-true-F0}
alpha0_true = matrix(1, N, D)
lambda0_true = matrix(1, N, D)

fit_true_F0 = mf_poi_susie(
  Y = Y,
  F = F0,
  alpha0 = alpha0_true,
  lambda0 = lambda0_true,
  max_iters = 100,
  update_prior = TRUE,
  break_symmetry = FALSE,
  init_seed = 1,
  update_F = FALSE,
  fine_init_gamma = FALSE,
  step_init = 0.1,
  step_ramp = 50
)
```

```{r elbo-true-F0}
plot(fit_true_F0$elbo, type = "l", xlab = "Iteration", ylab = "ELBO",
     main = "ELBO with true F0")
```

```{r recovered-labels-true-F0}
top_k_true_F0 = matrix(0, N, D)
for (d in 1:D) {
  top_k_true_F0[, d] = apply(fit_true_F0$gamma_bar[, d, ], 1, which.max)
}

recovered_label_true_F0 = apply(top_k_true_F0, 1, function(x) {
  paste(sort(x), collapse = ",")
})

recovered_factor_mat_true_F0 = matrix(0, nrow = K, ncol = N)
for (i in 1:N) {
  recovered_factor_mat_true_F0[top_k_true_F0[i, ], i] =
    recovered_factor_mat_true_F0[top_k_true_F0[i, ], i] + 1
}

image(x = seq_len(N), y = seq_len(K), z = t(recovered_factor_mat_true_F0),
      axes = FALSE, xlab = "Sample", ylab = "Recovered factor",
      main = "Recovered factors with true F0",
      col = c("white", "firebrick", "darkred"))
axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
     labels = paste("G", 1:8))
axis(2, at = seq_len(K), labels = seq_len(K), las = 1)
abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
       col = "gray80", lty = 3)
box()
```

```{r recovered-table-true-F0}
recovered_by_group_true_F0 = do.call(rbind, lapply(1:8, function(g) {
  idx = which(grp == g)
  tab = table(recovered_label_true_F0[idx])
  data.frame(group = g, recovered_pattern = names(tab), n = as.integer(tab),
             row.names = NULL)
}))

recovered_by_group_true_F0
```

## Fit With Unknown F

The fit starts from a random row-normalized factor matrix and updates `F` inside
the variational loop.

```{r fit}
alpha0 = matrix(1, N, D)
lambda0 = matrix(1, N, D)

F = matrix(rgamma(M * K, shape = 0.2, rate = 0.05), nrow = K, ncol = M)
F = F / rowSums(F)

fit = mf_poi_susie(
  Y = Y,
  F = F,
  alpha0 = alpha0,
  lambda0 = lambda0,
  max_iters = 100,
  update_prior = TRUE,
  break_symmetry = FALSE,
  init_seed = 1,
  update_F = TRUE,
  fine_init_gamma = FALSE,
  step_init = 0.1,
  step_ramp = 50
)
```

```{r elbo}
plot(fit$elbo, type = "l", xlab = "Iteration", ylab = "ELBO",
     main = "Variational objective")
```

## Recovery

For each observation and effect, we summarize the factor with the largest
posterior assignment probability.

```{r top-k}
top_k = matrix(0, N, D)
for (d in 1:D) {
  top_k[, d] = apply(fit$gamma_bar[, d, ], 1, which.max)
}

recovered_by_group = do.call(rbind, lapply(1:8, function(g) {
  idx = which(grp == g)
  tab = table(apply(top_k[idx, ], 1, function(x) paste(sort(x), collapse = ",")))
  data.frame(group = g, recovered_pattern = names(tab), n = as.integer(tab),
             row.names = NULL)
}))

recovered_by_group
```

```{r representative-posterior}
representative_idx = c(
  which(grp == 1)[1],
  which(grp == 2)[1],
  which(grp == 3)[1],
  which(grp == 4)[1],
  which(grp == 5)[1],
  which(grp == 6)[4],
  which(grp == 7)[1],
  which(grp == 8)[1]
)

representative_gamma = lapply(seq_along(representative_idx), function(j) {
  i = representative_idx[j]
  out = round(fit$gamma_bar[i, , ], 2)
  rownames(out) = paste0("effect_", seq_len(D))
  colnames(out) = paste0("factor_", seq_len(K))
  out
})
names(representative_gamma) = paste0("group_", 1:8)

representative_gamma
```

```{r recovered-factors-learned-F}
recovered_factor_mat = matrix(0, nrow = K, ncol = N)
for (i in 1:N) {
  recovered_factor_mat[top_k[i, ], i] =
    recovered_factor_mat[top_k[i, ], i] + 1
}

image(x = seq_len(N), y = seq_len(K), z = t(recovered_factor_mat),
      axes = FALSE, xlab = "Sample", ylab = "Recovered factor",
      main = "Recovered factors with learned F",
      col = c("white", "firebrick", "darkred"))
axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
     labels = paste("G", 1:8))
axis(2, at = seq_len(K), labels = seq_len(K), las = 1)
abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
       col = "gray80", lty = 3)
box()
```

The learned `F` is identifiable only up to permutation of its rows. A direct
row-wise comparison to `F0` is therefore not meaningful without matching rows,
but the group-level top-factor summaries show whether observations are assigned
to stable one-factor or three-factor patterns after jointly updating `F`.
