---
title: "MiSo Baseline Benchmark"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This file compares soft MiSo with two simple baselines:

- Poisson NMF followed by k-means clustering on the fitted loading matrix.
- MF-Poisson-SuSiE followed by k-means clustering on posterior expected loading
  scores.

The goal is not to produce the final paper benchmark. It is a compact
side-by-side table for checking whether motif-level modeling improves
clustering and sparse support recovery in the current simulations.

```{r source-code}
source("code/miso.R")
```

## Helpers

```{r helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)
  perms = all_permutations(seq_len(K))
  scores = apply(perms, 1, function(p) {
    sum(similarity[cbind(seq_len(K), p)])
  })
  learned_to_true = perms[which.max(scores), ]

  list(
    table = data.frame(
      learned_factor = seq_len(K),
      matched_true_factor = learned_to_true,
      cosine = similarity[cbind(seq_len(K), learned_to_true)]
    ),
    learned_to_true = learned_to_true,
    similarity = similarity
  )
}

support_label = function(support) paste(sort(unique(support)), collapse = ",")

simulate_support_scenario = function(N, M, K, group_supports, shape_vec,
                                      factor_shape = 0.1,
                                      factor_rate = 0.01, seed = 1) {
  set.seed(seed)
  S = length(group_supports)
  n_per_group = N / S
  grp = rep(seq_len(S), each = n_per_group)

  F0 = matrix(rgamma(M * K, shape = factor_shape, rate = factor_rate),
              nrow = K, ncol = M)
  F0 = normalize_rows(F0)

  L = matrix(0, nrow = N, ncol = K)
  for (s in seq_len(S)) {
    idx = which(grp == s)
    support = group_supports[[s]]
    L[idx, support] = matrix(
      rgamma(length(idx) * length(support), shape = shape_vec[s], rate = 1),
      nrow = length(idx),
      ncol = length(support)
    )
  }

  Y = matrix(rpois(N * M, L %*% F0), nrow = N, ncol = M)

  list(
    Y = Y,
    F0 = F0,
    L = L,
    grp = grp,
    group_supports = group_supports,
    true_support_size = rowSums(L > 0),
    true_support_label = apply(L > 0, 1, function(x) {
      paste(which(x), collapse = ",")
    }),
    true_patterns = vapply(group_supports, support_label, character(1))
  )
}

normalize_scores = function(scores, eps = 1e-12) {
  scores = pmax(scores, eps)
  scores / rowSums(scores)
}

kmeans_cluster_scores = function(scores, S, seed = 1) {
  set.seed(seed)
  kmeans(normalize_scores(scores), centers = S, nstart = 20)$cluster
}

map_scores_to_true = function(scores, learned_to_true) {
  K = length(learned_to_true)
  out = matrix(0, nrow = nrow(scores), ncol = K)
  for (k in seq_len(K)) {
    out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
  }
  out
}

support_labels_from_scores = function(scores_true_space, support_sizes) {
  vapply(seq_len(nrow(scores_true_space)), function(i) {
    active = order(scores_true_space[i, ], decreasing = TRUE)[
      seq_len(support_sizes[i])
    ]
    paste(sort(active), collapse = ",")
  }, character(1))
}

support_accuracy = function(scores, learned_to_true, dat) {
  scores_true = map_scores_to_true(scores, learned_to_true)
  mean(support_labels_from_scores(scores_true, dat$true_support_size) ==
         dat$true_support_label)
}

gamma_uncertainty_summary = function(gamma_bar) {
  ent = motif_gamma_entropy(gamma_bar)
  top = apply(gamma_bar, c(1, 2), max)
  data.frame(
    mean_gamma_entropy = mean(ent),
    max_gamma_entropy = max(ent),
    mean_top_gamma = mean(top),
    min_top_gamma = min(top)
  )
}

fit_nmf_baseline = function(dat, K, S, seed, nmf_iters = 80) {
  nmf_fit = poisson_nmf_init(dat$Y, K = K, max_iters = nmf_iters,
                             init_seed = seed)
  factor_match = match_factor_rows(nmf_fit$F, dat$F0)
  scores = nmf_fit$L
  z_hat = kmeans_cluster_scores(scores, S = S, seed = seed)

  data.frame(
    method = "Poisson NMF + kmeans",
    cluster_accuracy = cluster_accuracy(z_hat, dat$grp),
    support_accuracy = support_accuracy(scores, factor_match$learned_to_true,
                                        dat),
    mean_factor_cosine = mean(factor_match$table$cosine),
    min_factor_cosine = min(factor_match$table$cosine),
    mean_max_responsibility = NA_real_,
    mean_gamma_entropy = NA_real_,
    max_gamma_entropy = NA_real_,
    mean_top_gamma = NA_real_,
    min_top_gamma = NA_real_
  )
}

fit_mf_baseline = function(dat, K, S, D, seed, max_iters = 35,
                            nmf_iters = 60) {
  mf_fit = poisson_susie_nmf(
    Y = dat$Y,
    K = K,
    D = D,
    max_iters = max_iters,
    update_prior = TRUE,
    init_seed = seed,
    init_F = "poisson_nmf",
    nmf_iters = nmf_iters,
    init_gamma_from_nmf = TRUE,
    elbo_every = 5,
    tol = 5e-4,
    min_iters = 15,
    patience = 2
  )

  factor_match = match_factor_rows(mf_fit$F, dat$F0)
  scores = loading_scores_from_mf(mf_fit)
  z_hat = kmeans_cluster_scores(scores, S = S, seed = seed)

  data.frame(
    method = "MF-Poisson-SuSiE + kmeans",
    cluster_accuracy = cluster_accuracy(z_hat, dat$grp),
    support_accuracy = support_accuracy(scores, factor_match$learned_to_true,
                                        dat),
    mean_factor_cosine = mean(factor_match$table$cosine),
    min_factor_cosine = min(factor_match$table$cosine),
    mean_max_responsibility = NA_real_,
    mean_gamma_entropy = NA_real_,
    max_gamma_entropy = NA_real_,
    mean_top_gamma = NA_real_,
    min_top_gamma = NA_real_
  )
}

fit_miso_method = function(dat, K, S, D, seed, max_iters = 18,
                                n_inner = 3, mf_max_iters = 35,
                                mf_nmf_iters = 60, block_size = 100) {
  fit = miso(
    Y = dat$Y,
    K = K,
    S = S,
    D = D,
    max_iters = max_iters,
    n_inner = n_inner,
    mf_max_iters = mf_max_iters,
    mf_nmf_iters = mf_nmf_iters,
    init_seed = seed,
    update_prior = TRUE,
    update_F = TRUE,
    update_gamma = TRUE,
    gamma_init_floor = 0.05,
    gamma_step_init = 0.5,
    gamma_step_ramp = 8,
    F_step_init = 0.2,
    F_step_ramp = 15,
    tol = 1e-5,
    min_iters = 5,
    patience = 2,
    block_size = block_size
  )

  factor_match = match_factor_rows(fit$F, dat$F0)
  scores = fit$omega %*% motif_factor_scores(fit)$scores
  uncertainty = gamma_uncertainty_summary(fit$gamma_bar)

  cbind(
    data.frame(
      method = "Soft MiSo",
      cluster_accuracy = cluster_accuracy(fit$z_hat, dat$grp),
      support_accuracy = support_accuracy(scores, factor_match$learned_to_true,
                                          dat),
      mean_factor_cosine = mean(factor_match$table$cosine),
      min_factor_cosine = min(factor_match$table$cosine),
      mean_max_responsibility = mean(apply(fit$omega, 1, max))
    ),
    uncertainty
  )
}

run_benchmark = function(dat, K, S, D, seed) {
  rbind(
    fit_nmf_baseline(dat, K = K, S = S, seed = seed),
    fit_mf_baseline(dat, K = K, S = S, D = D, seed = seed),
    fit_miso_method(dat, K = K, S = S, D = D, seed = seed)
  )
}

plot_benchmark_metric = function(results, metric, main) {
  vals = results[[metric]]
  barplot(vals, names.arg = results$method, las = 2, ylim = c(0, 1),
          ylab = metric, main = main, col = "gray70")
  text(seq_along(vals) * 1.2 - 0.5, pmin(vals + 0.04, 0.96),
       sprintf("%.2f", vals), cex = 0.8)
}
```

## Block Scenario

The block scenario has five one-factor groups and three three-factor groups.
This is the easier setting where anchor-like single-factor groups help factor
recovery.

```{r block-simulate}
block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

block = simulate_support_scenario(
  N = 320,
  M = 350,
  K = 5,
  group_supports = block_supports,
  shape_vec = c(rep(10, 5), rep(20, 3)),
  seed = 42
)

data.frame(group = seq_along(block_supports),
           true_active_factors = block$true_patterns)
```

```{r block-fit}
block_results = run_benchmark(
  dat = block,
  K = 5,
  S = length(block_supports),
  D = 3,
  seed = 1
)

block_results
```

```{r block-plots}
plot_benchmark_metric(block_results, "cluster_accuracy",
                      "Block scenario: cluster accuracy")
plot_benchmark_metric(block_results, "support_accuracy",
                      "Block scenario: support accuracy")
plot_benchmark_metric(block_results, "mean_factor_cosine",
                      "Block scenario: mean factor cosine")
```

## Tree Scenario

The tree scenario is harder because no group occupies a single factor. All
groups share factor 1 and branch through factors 2 or 5 before adding optional
leaf factors.

```{r tree-simulate}
tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

tree = simulate_support_scenario(
  N = 360,
  M = 420,
  K = 7,
  group_supports = tree_supports,
  shape_vec = rep(18, length(tree_supports)),
  seed = 47
)

data.frame(group = seq_along(tree_supports),
           true_active_factors = tree$true_patterns)
```

```{r tree-fit}
tree_results = run_benchmark(
  dat = tree,
  K = 7,
  S = length(tree_supports),
  D = 3,
  seed = 2
)

tree_results
```

```{r tree-plots}
plot_benchmark_metric(tree_results, "cluster_accuracy",
                      "Tree scenario: cluster accuracy")
plot_benchmark_metric(tree_results, "support_accuracy",
                      "Tree scenario: support accuracy")
plot_benchmark_metric(tree_results, "mean_factor_cosine",
                      "Tree scenario: mean factor cosine")
```

## Current Interpretation

This benchmark is deliberately small, but it gives the basic empirical table a
statistics paper will need: a reconstruction/factorization baseline, a
per-sample sparse posterior baseline, and the motif-level model. The quantities
to watch are not only factor recovery. The paper claim requires MiSo to
improve motif clustering and support-pattern recovery, while `gamma_bar`
provides additional motif uncertainty diagnostics that the baselines do not
have.

The current result is strongest in the block scenario, where soft MiSo
improves clustering, support recovery, and factor recovery. In the tree
scenario, the result is more nuanced: soft MiSo improves support accuracy
relative to both baselines, but its hard cluster labels from `max.col(omega)`
are less accurate than k-means on the baseline loading scores. That is useful
evidence rather than a failure, because it points to the next methodological
question: whether the motif model needs better initialization, more stable
responsibility updates, or a tree-aware diagnostic that does not reduce shared
branch structure to one hard cluster label.
