---
title: "Motif-Specific Dimension Selection by SuSiE Credible-Set Purity"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

This experiment fits MiSo with a common upper bound
$D_{\mathrm{fit}}=5$ and selects an effective dimension separately for each
motif using the credible-set purity rule from SuSiE. The Gamma prior shape and
rate parameters are still updated by empirical Bayes during fitting, but they
are not used for dimension selection.

```{r source-code}
source("code/miso-benchmark-utils.R")
```

## Credible sets and purity

For each motif $s$ and fitted dimension $d$, order the posterior factor
probabilities $\bar\gamma_{sdk}$ from largest to smallest. The 90% credible set
$\mathrm{CS}_{sd}$ is the smallest subset satisfying

$$
\sum_{k\in\mathrm{CS}_{sd}}\bar\gamma_{sdk}\geq 0.9.
$$

Let $R_{kk'}=\operatorname{cor}(F_k,F_{k'})$ be the Pearson correlation between
the fitted factor rows. Following the standard SuSiE convention, credible-set
purity is the minimum absolute pairwise correlation,

$$
\operatorname{purity}(\mathrm{CS}_{sd})
=
\min_{k\neq k'\in\mathrm{CS}_{sd}}|R_{kk'}|.
$$

A singleton credible set has purity one. We retain credible sets whose purity
is at least 0.5 and remove exact duplicate credible sets within each motif.
The estimated motif dimension is the number of distinct retained credible
sets. The 0.5 cutoff is fixed as the primary rule; other cutoffs are reported
only as sensitivity analysis.

## Simulation design

```{r design}
n_seed = as.integer(Sys.getenv("MISO_CS_SEEDS", "20"))
benchmark_seeds = seq_len(n_seed)
coverage = 0.9
purity_cutoffs = c(0, 0.25, 0.5, 0.75, 0.9)
primary_cutoff = 0.5
train_fraction = 0.8
D_fit = 5

block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

data.frame(
  scenario = c("block", "tree/no-anchor"),
  K = c(5, 7),
  S = c(length(block_supports), length(tree_supports)),
  true_motif_dimensions = c("1 or 3", "2 or 3"),
  fitted_D = D_fit,
  N = 240,
  M = c(250, 280),
  n_seed = n_seed,
  credible_set_coverage = coverage
)
```

Both repeated-surplus and diffuse-surplus initializations are evaluated.
Repeated pure dimensions should be removed as duplicate credible sets, while
diffuse dimensions should produce low-purity credible sets.

## Evaluation helpers

```{r helpers}
match_motifs_to_groups = function(omega, true_group) {
  true_group = as.integer(as.factor(true_group))
  S_fit = ncol(omega)
  S_true = max(true_group)
  if (S_fit != S_true) stop("This experiment requires fitted S to equal true S")

  score = matrix(0, nrow = S_fit, ncol = S_true)
  for (g in seq_len(S_true)) {
    score[, g] = colSums(omega[true_group == g, , drop = FALSE])
  }
  if (requireNamespace("clue", quietly = TRUE)) {
    as.integer(clue::solve_LSAP(max(score) - score))
  } else {
    permutations = all_permutations(seq_len(S_true))
    objective = apply(permutations, 1, function(p) {
      sum(score[cbind(seq_len(S_fit), p)])
    })
    permutations[which.max(objective), ]
  }
}

parse_index_set = function(x) {
  if (is.na(x) || x == "") return(integer(0))
  as.integer(strsplit(x, ",", fixed = TRUE)[[1]])
}

jaccard = function(estimated, truth) {
  if (length(estimated) == 0 && length(truth) == 0) return(1)
  length(intersect(estimated, truth)) / length(union(estimated, truth))
}

evaluate_cutoff = function(credible_set_table, purity_cutoff) {
  do.call(rbind, lapply(split(credible_set_table,
                              credible_set_table$motif), function(tab) {
    retained = which(tab$purity >= purity_cutoff)
    if (length(retained) > 0) {
      retained = retained[!duplicated(tab$credible_set[retained])]
    }

    selected_top_factors = sort(unique(tab$top_true_factor[retained]))
    selected_cs_union = sort(unique(unlist(tab$credible_set_true[retained])))
    truth = sort(unique(tab$true_support[[1]]))

    data.frame(
      scenario = tab$scenario[1],
      initialization = tab$initialization[1],
      seed = tab$seed[1],
      purity_cutoff = purity_cutoff,
      motif = tab$motif[1],
      matched_group = tab$matched_group[1],
      true_D = length(truth),
      D_hat = length(retained),
      dimension_error = length(retained) - length(truth),
      exact_D = length(retained) == length(truth),
      exact_top_factor_support =
        length(selected_top_factors) == length(truth) &&
        setequal(selected_top_factors, truth),
      top_factor_jaccard = jaccard(selected_top_factors, truth),
      credible_set_true_factor_coverage =
        mean(truth %in% selected_cs_union),
      credible_set_union_precision = if (length(selected_cs_union) == 0) 0 else
        mean(selected_cs_union %in% truth),
      selected_credible_sets =
        paste(tab$credible_set[retained], collapse = ";"),
      stringsAsFactors = FALSE
    )
  }))
}

summarize_selection = function(selection_table) {
  keys = interaction(selection_table$scenario,
                     selection_table$initialization,
                     selection_table$purity_cutoff, drop = TRUE)
  do.call(rbind, lapply(split(selection_table, keys), function(tab) {
    run_keys = interaction(tab$seed, tab$scenario, tab$initialization,
                           tab$purity_cutoff, drop = TRUE)
    run_summary = do.call(rbind, lapply(split(tab, run_keys), function(run) {
      data.frame(
        exact_D = mean(run$exact_D),
        D_mae = mean(abs(run$dimension_error)),
        D_bias = mean(run$dimension_error),
        exact_top_factor_support = mean(run$exact_top_factor_support),
        top_factor_jaccard = mean(run$top_factor_jaccard),
        credible_set_true_factor_coverage =
          mean(run$credible_set_true_factor_coverage),
        credible_set_union_precision =
          mean(run$credible_set_union_precision)
      )
    }))

    data.frame(
      scenario = tab$scenario[1],
      initialization = tab$initialization[1],
      purity_cutoff = tab$purity_cutoff[1],
      n_seed = length(unique(tab$seed)),
      exact_D_mean = mean(run_summary$exact_D),
      exact_D_sd = sd(run_summary$exact_D),
      D_mae_mean = mean(run_summary$D_mae),
      D_mae_sd = sd(run_summary$D_mae),
      D_bias_mean = mean(run_summary$D_bias),
      D_bias_sd = sd(run_summary$D_bias),
      exact_top_factor_support_mean =
        mean(run_summary$exact_top_factor_support),
      exact_top_factor_support_sd = sd(run_summary$exact_top_factor_support),
      top_factor_jaccard_mean = mean(run_summary$top_factor_jaccard),
      top_factor_jaccard_sd = sd(run_summary$top_factor_jaccard),
      credible_set_true_factor_coverage_mean =
        mean(run_summary$credible_set_true_factor_coverage),
      credible_set_union_precision_mean =
        mean(run_summary$credible_set_union_precision)
    )
  }))
}
```

## Fit models

EB prior means are retained as diagnostics but do not enter
`evaluate_cutoff()`.

```{r run-helpers}
cache_version = "v3"
cache_dir = file.path("analysis", "cache",
                      paste0("miso-CS-D-", cache_version))
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

run_one_fit = function(scenario, initialization, seed) {
  cache_file = file.path(
    cache_dir,
    sprintf("%s-%s-seed-%03d.rds", scenario, initialization, seed)
  )
  if (file.exists(cache_file)) return(readRDS(cache_file))

  if (scenario == "block") {
    supports = block_supports
    dat = simulate_support_scenario(
      N = 240, M = 250, K = 5,
      group_supports = supports,
      shape_vec = c(rep(10, 5), rep(20, 3)),
      seed = 5000 + seed,
      train_fraction = train_fraction
    )
    K = 5
  } else if (scenario == "tree") {
    supports = tree_supports
    dat = simulate_support_scenario(
      N = 240, M = 280, K = 7,
      group_supports = supports,
      shape_vec = rep(18, length(supports)),
      seed = 6000 + seed,
      train_fraction = train_fraction
    )
    K = 7
  } else {
    stop("Unknown scenario")
  }

  fit = miso(
    Y = dat$Y,
    K = K,
    S = length(supports),
    D = D_fit,
    max_iters = 12,
    n_inner = 3,
    mf_max_iters = 25,
    mf_nmf_iters = 45,
    init_seed = seed,
    update_prior = TRUE,
    update_F = TRUE,
    update_gamma = TRUE,
    update_slot_scale = FALSE,
    gamma_init_floor = 0.05,
    surplus_slots = initialization,
    gamma_step_init = 0.5,
    gamma_step_ramp = 8,
    F_step_init = 0.2,
    F_step_ramp = 12,
    tol = 1e-5,
    min_iters = 5,
    patience = 2,
    block_size = 100
  )

  factor_match = match_factor_rows(fit$F, dat$F0)
  motif_to_group = match_motifs_to_groups(fit$omega, dat$grp)
  credible_sets = motif_credible_sets(
    fit,
    coverage = coverage,
    min_abs_correlation = 0,
    deduplicate = FALSE
  )$dimensions
  prior_mean = fit$alpha0 / pmax(miso_fit_beta0(fit), 1e-12)
  prior_fraction = prior_mean / rowSums(prior_mean)

  credible_sets$scenario = scenario
  credible_sets$initialization = initialization
  credible_sets$seed = seed
  credible_sets$matched_group = motif_to_group[credible_sets$motif]
  credible_sets$true_D = lengths(supports)[credible_sets$matched_group]
  credible_sets$true_support = I(lapply(
    credible_sets$matched_group,
    function(group) supports[[group]]
  ))
  credible_sets$credible_set_fitted = lapply(
    credible_sets$credible_set,
    parse_index_set
  )
  credible_sets$credible_set_true = I(lapply(
    credible_sets$credible_set_fitted,
    function(set) sort(factor_match$learned_to_true[set])
  ))
  credible_sets$top_true_factor =
    factor_match$learned_to_true[credible_sets$top_factor]
  credible_sets$EB_prior_mean = prior_mean[
    cbind(credible_sets$motif, credible_sets$dimension)
  ]
  credible_sets$EB_prior_fraction = prior_fraction[
    cbind(credible_sets$motif, credible_sets$dimension)
  ]

  fit_summary = data.frame(
    scenario = scenario,
    initialization = initialization,
    seed = seed,
    cluster_accuracy = cluster_accuracy(fit$z_hat, dat$grp),
    mean_factor_cosine = mean(factor_match$table$cosine),
    min_factor_cosine = min(factor_match$table$cosine),
    mean_max_responsibility = mean(apply(fit$omega, 1, max))
  )

  result = list(credible_sets = credible_sets, fit_summary = fit_summary)
  saveRDS(result, cache_file)
  result
}
```

```{r run-models}
jobs = expand.grid(
  scenario = c("block", "tree"),
  initialization = c("repeat", "uniform"),
  seed = benchmark_seeds,
  stringsAsFactors = FALSE
)
run_job = function(j) {
  run_one_fit(jobs$scenario[j], jobs$initialization[j], jobs$seed[j])
}

requested_cores = as.integer(Sys.getenv("MISO_CS_CORES", "4"))
available_cores = parallel::detectCores()
if (is.na(requested_cores)) requested_cores = 1L
if (is.na(available_cores)) available_cores = requested_cores
n_cores = max(1L, min(requested_cores, available_cores, nrow(jobs)))
if (.Platform$OS.type == "windows" || n_cores == 1L) {
  result_list = lapply(seq_len(nrow(jobs)), run_job)
} else {
  result_list = parallel::mclapply(seq_len(nrow(jobs)), run_job,
                                   mc.cores = n_cores)
}

credible_set_results = do.call(rbind, lapply(result_list, `[[`,
                                              "credible_sets"))
fit_results = do.call(rbind, lapply(result_list, `[[`, "fit_summary"))
selection_results = do.call(rbind, lapply(result_list, function(result) {
  do.call(rbind, lapply(purity_cutoffs, function(cutoff) {
    evaluate_cutoff(result$credible_sets, cutoff)
  }))
}))
rownames(credible_set_results) = NULL
rownames(selection_results) = NULL
rownames(fit_results) = NULL

saveRDS(
  list(credible_sets = credible_set_results,
       selection = selection_results,
       fits = fit_results),
  file.path(cache_dir, sprintf("combined-%03d-seeds.rds", n_seed))
)
```

## Selection results

```{r selection-summary}
selection_summary = summarize_selection(selection_results)
numeric_columns = vapply(selection_summary, is.numeric, logical(1))
numeric_columns[match(c("purity_cutoff", "n_seed"),
                      names(selection_summary))] = FALSE
selection_summary[numeric_columns] = round(
  selection_summary[numeric_columns], 4
)
selection_summary
```

```{r cutoff-plots}
plot_cutoff_metric = function(summary_table, metric, ylab) {
  op = par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  on.exit(par(op))
  for (scenario in c("block", "tree")) {
    tab = summary_table[summary_table$scenario == scenario, ]
    repeat_tab = tab[tab$initialization == "repeat", ]
    uniform_tab = tab[tab$initialization == "uniform", ]
    plot(repeat_tab$purity_cutoff, repeat_tab[[metric]], type = "b", pch = 19,
         ylim = c(0, 1), xlab = "Minimum CS purity", ylab = ylab,
         main = scenario)
    lines(uniform_tab$purity_cutoff, uniform_tab[[metric]], type = "b",
          pch = 17, col = "firebrick")
    abline(v = primary_cutoff, lty = 3, col = "gray40")
    legend("bottomleft", c("repeat surplus", "diffuse surplus"),
           pch = c(19, 17), col = c("black", "firebrick"), bty = "n")
  }
}

plot_cutoff_metric(selection_summary, "exact_D_mean",
                   "Fraction of motifs with correct D")
plot_cutoff_metric(selection_summary, "exact_top_factor_support_mean",
                   "Fraction of exact top-factor supports")
plot_cutoff_metric(selection_summary,
                   "credible_set_true_factor_coverage_mean",
                   "True-factor coverage by selected CSs")
```

## Primary cutoff

```{r primary-summary}
primary_summary = selection_summary[
  selection_summary$purity_cutoff == primary_cutoff,
  c("scenario", "initialization", "n_seed",
    "exact_D_mean", "exact_D_sd", "D_mae_mean", "D_mae_sd",
    "D_bias_mean", "D_bias_sd",
    "exact_top_factor_support_mean", "exact_top_factor_support_sd",
    "top_factor_jaccard_mean", "top_factor_jaccard_sd",
    "credible_set_true_factor_coverage_mean",
    "credible_set_union_precision_mean")
]
primary_summary
```

```{r dimension-confusion}
primary_selection = selection_results[
  selection_results$purity_cutoff == primary_cutoff, ]
dimension_confusion = as.data.frame(with(
  primary_selection,
  table(scenario, initialization, true_D, D_hat)
))
dimension_confusion[dimension_confusion$Freq > 0, ]
```

## Credible-set diagnostics

```{r credible-set-summary}
credible_set_group = interaction(
  credible_set_results$scenario,
  credible_set_results$initialization,
  credible_set_results$seed,
  credible_set_results$motif,
  drop = TRUE
)
credible_set_results$duplicate_within_motif = ave(
  seq_len(nrow(credible_set_results)),
  credible_set_group,
  FUN = function(index) {
    as.numeric(duplicated(credible_set_results$credible_set[index]))
  }
)

aggregate(
  cbind(credible_set_size, posterior_coverage, purity,
        duplicate_within_motif, EB_prior_fraction) ~
    scenario + initialization,
  data = credible_set_results,
  FUN = mean
)
```

```{r credible-set-plots}
op = par(mfrow = c(1, 2), mar = c(7, 4, 3, 1))
for (scenario in c("block", "tree")) {
  tab = credible_set_results[credible_set_results$scenario == scenario, ]
  boxplot(tab$purity ~ tab$initialization, ylim = c(0, 1),
          ylab = "Minimum absolute correlation", xlab = "Initialization",
          main = scenario, col = "gray85")
  abline(h = primary_cutoff, lty = 3, col = "firebrick")
}
par(op)
```

The EB prior fractions verify that prior updates still occur. They are not used
to retain, reject, or deduplicate credible sets.

## Fit quality

```{r fit-summary}
fit_summary = do.call(rbind, lapply(
  split(fit_results,
        interaction(fit_results$scenario, fit_results$initialization,
                    drop = TRUE)),
  function(tab) {
    data.frame(
      scenario = tab$scenario[1],
      initialization = tab$initialization[1],
      n_seed = nrow(tab),
      cluster_accuracy_mean = mean(tab$cluster_accuracy),
      cluster_accuracy_sd = sd(tab$cluster_accuracy),
      mean_factor_cosine_mean = mean(tab$mean_factor_cosine),
      mean_factor_cosine_sd = sd(tab$mean_factor_cosine),
      mean_max_responsibility = mean(tab$mean_max_responsibility)
    )
  }
))
fit_summary[, vapply(fit_summary, is.numeric, logical(1))] = round(
  fit_summary[, vapply(fit_summary, is.numeric, logical(1))], 4
)
fit_summary
```

## Observed results

At the primary purity cutoff 0.5, repeated-surplus initialization gives the
strongest block result. It selects the exact motif dimension in
`r sprintf("%.1f", 100 * primary_summary$exact_D_mean[primary_summary$scenario == "block" & primary_summary$initialization == "repeat"])`%
of motifs, with dimension MAE
`r sprintf("%.3f", primary_summary$D_mae_mean[primary_summary$scenario == "block" & primary_summary$initialization == "repeat"])`.
The top factors of the selected credible sets recover the exact motif support
in the same fraction of cases, with mean support Jaccard
`r sprintf("%.3f", primary_summary$top_factor_jaccard_mean[primary_summary$scenario == "block" & primary_summary$initialization == "repeat"])`.
Diffuse-surplus initialization gives lower block dimension recovery of
`r sprintf("%.1f", 100 * primary_summary$exact_D_mean[primary_summary$scenario == "block" & primary_summary$initialization == "uniform"])`%.

The tree/no-anchor result remains difficult. Exact dimension recovery is
`r sprintf("%.1f", 100 * primary_summary$exact_D_mean[primary_summary$scenario == "tree" & primary_summary$initialization == "repeat"])`%
under repeated surplus and
`r sprintf("%.1f", 100 * primary_summary$exact_D_mean[primary_summary$scenario == "tree" & primary_summary$initialization == "uniform"])`%
under diffuse surplus. Their dimension MAEs are
`r sprintf("%.3f", primary_summary$D_mae_mean[primary_summary$scenario == "tree" & primary_summary$initialization == "repeat"])`
and
`r sprintf("%.3f", primary_summary$D_mae_mean[primary_summary$scenario == "tree" & primary_summary$initialization == "uniform"])`,
respectively. Exact top-factor support recovery is only
`r sprintf("%.1f", 100 * primary_summary$exact_top_factor_support_mean[primary_summary$scenario == "tree" & primary_summary$initialization == "repeat"])`%
and
`r sprintf("%.1f", 100 * primary_summary$exact_top_factor_support_mean[primary_summary$scenario == "tree" & primary_summary$initialization == "uniform"])`%.

The purity cutoff is not the main source of this failure. Results are nearly
unchanged between cutoffs 0.25 and 0.9. With repeated initialization, every
credible set is a singleton and therefore has purity one; exact duplicate
removal collapses many of them but cannot reject a distinct incorrect
singleton. With diffuse initialization, approximately
`r sprintf("%.1f", 100 * mean(credible_set_results$credible_set_size[credible_set_results$scenario == "tree" & credible_set_results$initialization == "uniform"] == 1))`%
of tree credible sets are still singletons. Consequently, most surplus
dimensions also pass the purity filter.

The tree fits are not globally degenerate: mean clustering accuracy is
`r sprintf("%.3f", mean(fit_results$cluster_accuracy[fit_results$scenario == "tree"]))`
and mean factor cosine is
`r sprintf("%.3f", mean(fit_results$mean_factor_cosine[fit_results$scenario == "tree"]))`.
Rather, the model can fit useful motifs while assigning some surplus dimensions
to confident singleton credible sets.

The corrected conclusion is therefore that SuSiE credible-set purity and
duplicate removal work well for the block scenario but are not sufficient for
motif-specific dimension selection in the tree scenario. Purity answers
whether a credible set contains mutually similar factor alternatives; it does
not answer whether a pure singleton dimension has meaningful loading activity.
The EB updates remained active throughout but did not enter this selection
rule, as intended.

## Interpretation guide

Exact dimension recovery is the primary model-selection metric. Exact
top-factor support additionally asks whether the most probable factor in each
selected credible set recovers the motif support after factor matching.
Credible-set true-factor coverage is less strict: it asks whether each true
factor appears anywhere in at least one selected credible set. Union precision
penalizes credible sets that contain unrelated factors.

The repeated-surplus comparison tests whether SuSiE-style duplicate removal
collapses repeated dimensions. The diffuse-surplus comparison tests whether
unused dimensions form broad, low-purity credible sets and are filtered out.
The threshold sensitivity analysis indicates whether conclusions depend on the
standard 0.5 minimum-correlation cutoff.
