---
title: "Alternating MiSo Poisson SuSiE Updates"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This file tests the alternating hard-motif MiSo-Poisson-SuSiE update. Both
comparisons use the same MF-Poisson-SuSiE initialization. The frozen fit keeps
the initialized `F` and motifs fixed; the alternating fit updates both `F` and
hard motifs inside the MiSo loop.

```{r source-code}
source("code/miso-poi-susie.R")
```

## Helpers

```{r helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)
  perms = all_permutations(seq_len(K))
  scores = apply(perms, 1, function(p) {
    sum(similarity[cbind(seq_len(K), p)])
  })
  learned_to_true = perms[which.max(scores), ]

  list(
    table = data.frame(
      learned_factor = seq_len(K),
      matched_true_factor = learned_to_true,
      cosine = similarity[cbind(seq_len(K), learned_to_true)]
    ),
    learned_to_true = learned_to_true,
    similarity = similarity
  )
}

support_label = function(support) paste(sort(unique(support)), collapse = ",")

simulate_support_scenario = function(N, M, K, group_supports, shape_vec,
                                      factor_shape = 0.1,
                                      factor_rate = 0.01, seed = 1) {
  set.seed(seed)
  S = length(group_supports)
  n_per_group = N / S
  grp = rep(seq_len(S), each = n_per_group)

  F0 = matrix(rgamma(M * K, shape = factor_shape, rate = factor_rate),
              nrow = K, ncol = M)
  F0 = normalize_rows(F0)

  L = matrix(0, nrow = N, ncol = K)
  for (s in seq_len(S)) {
    idx = which(grp == s)
    support = group_supports[[s]]
    L[idx, support] = matrix(
      rgamma(length(idx) * length(support), shape = shape_vec[s], rate = 1),
      nrow = length(idx),
      ncol = length(support)
    )
  }

  Y = matrix(rpois(N * M, L %*% F0), nrow = N, ncol = M)

  list(
    Y = Y,
    F0 = F0,
    L = L,
    grp = grp,
    group_supports = group_supports,
    true_patterns = vapply(group_supports, support_label, character(1))
  )
}

translate_motifs_to_true_space = function(motifs, learned_to_true) {
  matrix(learned_to_true[motifs], nrow = nrow(motifs), ncol = ncol(motifs))
}

fit_frozen_and_alternating = function(dat, K, S, D, init_seed,
                                      mf_max_iters = 50,
                                      mf_nmf_iters = 80,
                                      miso_max_iters = 25,
                                      n_inner = 5) {
  mf_fit = mf_poi_susie(
    Y = dat$Y,
    K = K,
    D = D,
    max_iters = mf_max_iters,
    update_prior = TRUE,
    init_seed = init_seed,
    init_F = "poisson_nmf",
    nmf_iters = mf_nmf_iters,
    init_gamma_from_nmf = TRUE,
    elbo_every = 5,
    tol = 5e-4,
    min_iters = 15,
    patience = 2
  )

  motif_init = init_motifs_from_loading_scores(
    loading_scores_from_mf(mf_fit),
    S = S,
    D = D,
    init_seed = init_seed
  )

  frozen = miso_poi_susie_fixed_motifs(
    Y = dat$Y,
    F = mf_fit$F,
    motifs = motif_init$motifs,
    max_iters = miso_max_iters,
    n_inner = n_inner,
    update_prior = TRUE,
    update_F = FALSE,
    update_motifs = FALSE,
    tol = 1e-5,
    min_iters = 5,
    patience = 2
  )

  alternating = miso_poi_susie_fixed_motifs(
    Y = dat$Y,
    F = mf_fit$F,
    motifs = motif_init$motifs,
    max_iters = miso_max_iters,
    n_inner = n_inner,
    update_prior = TRUE,
    update_F = TRUE,
    update_motifs = TRUE,
    motif_update_every = 1,
    F_step_init = 0.2,
    F_step_ramp = 15,
    tol = 1e-5,
    min_iters = 5,
    patience = 2
  )

  list(mf_fit = mf_fit, motif_init = motif_init, frozen = frozen,
       alternating = alternating)
}

summarize_one_fit = function(name, fit, z_true, F0, motif_init) {
  factor_match = match_factor_rows(fit$F, F0)
  data.frame(
    fit = name,
    converged = fit$converged,
    n_iter = fit$n_iter,
    final_elbo = tail(fit$elbo[!is.na(fit$elbo)], 1),
    cluster_accuracy = cluster_accuracy(fit$z_hat, z_true),
    mean_max_responsibility = mean(apply(fit$omega, 1, max)),
    mean_factor_cosine = mean(factor_match$table$cosine),
    min_factor_cosine = min(factor_match$table$cosine),
    motif_slot_changes = sum(fit$motifs != motif_init$motifs)
  )
}

summarize_pair = function(pair, dat) {
  rbind(
    summarize_one_fit("frozen", pair$frozen, dat$grp, dat$F0,
                      pair$motif_init),
    summarize_one_fit("alternating", pair$alternating, dat$grp, dat$F0,
                      pair$motif_init)
  )
}

motif_table = function(fit, F0, label) {
  factor_match = match_factor_rows(fit$F, F0)
  motifs_true = translate_motifs_to_true_space(fit$motifs,
                                               factor_match$learned_to_true)
  data.frame(
    fit = label,
    fitted_submanifold = seq_len(nrow(fit$motifs)),
    motif_in_fitted_factor_space = apply(fit$motifs, 1, support_label),
    motif_mapped_to_true_factor_space = apply(motifs_true, 1, support_label)
  )
}

plot_group_responsibilities = function(omega, grp, main) {
  group_resp = do.call(rbind, lapply(sort(unique(grp)), function(g) {
    colMeans(omega[grp == g, , drop = FALSE])
  }))

  image(x = seq_len(ncol(group_resp)), y = seq_len(nrow(group_resp)),
        z = t(group_resp), axes = FALSE, xlab = "Fitted submanifold",
        ylab = "True group", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(group_resp)),
       labels = paste0("S", seq_len(ncol(group_resp))))
  axis(2, at = seq_len(nrow(group_resp)),
       labels = paste0("G", seq_len(nrow(group_resp))), las = 1)
  for (g in seq_len(nrow(group_resp))) {
    for (s in seq_len(ncol(group_resp))) {
      text(s, g, sprintf("%.2f", group_resp[g, s]), cex = 0.65)
    }
  }
  box()
}

plot_cosine_heatmap = function(similarity, main) {
  image(x = seq_len(ncol(similarity)), y = seq_len(nrow(similarity)),
        z = t(similarity), axes = FALSE, xlab = "True factor",
        ylab = "Fitted factor", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(similarity)),
       labels = paste0("true ", seq_len(ncol(similarity))))
  axis(2, at = seq_len(nrow(similarity)),
       labels = paste0("fit ", seq_len(nrow(similarity))), las = 1)
  for (i in seq_len(nrow(similarity))) {
    for (j in seq_len(ncol(similarity))) {
      text(j, i, sprintf("%.2f", similarity[i, j]), cex = 0.65)
    }
  }
  box()
}
```

## Block Scenario

```{r block-simulate}
block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

block = simulate_support_scenario(
  N = 480,
  M = 500,
  K = 5,
  group_supports = block_supports,
  shape_vec = c(rep(10, 5), rep(20, 3)),
  seed = 42
)

data.frame(group = seq_along(block_supports),
           true_active_factors = block$true_patterns)
```

```{r block-fit}
block_pair = fit_frozen_and_alternating(
  dat = block,
  K = 5,
  S = length(block_supports),
  D = 3,
  init_seed = 1,
  mf_max_iters = 50,
  mf_nmf_iters = 80,
  miso_max_iters = 25,
  n_inner = 5
)

summarize_pair(block_pair, block)
```

```{r block-motifs}
rbind(
  motif_table(block_pair$frozen, block$F0, "frozen"),
  motif_table(block_pair$alternating, block$F0, "alternating")
)
```

```{r block-plots}
plot_group_responsibilities(
  block_pair$frozen$omega,
  block$grp,
  "Block scenario: frozen MiSo responsibilities"
)

plot_group_responsibilities(
  block_pair$alternating$omega,
  block$grp,
  "Block scenario: alternating MiSo responsibilities"
)

plot_cosine_heatmap(
  match_factor_rows(block_pair$alternating$F, block$F0)$similarity,
  "Block scenario: alternating fitted and true factor cosine"
)
```

## Tree Scenario

```{r tree-simulate}
tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

tree = simulate_support_scenario(
  N = 480,
  M = 600,
  K = 7,
  group_supports = tree_supports,
  shape_vec = rep(18, length(tree_supports)),
  seed = 47
)

data.frame(group = seq_along(tree_supports),
           true_active_factors = tree$true_patterns)
```

```{r tree-fit}
tree_pair = fit_frozen_and_alternating(
  dat = tree,
  K = 7,
  S = length(tree_supports),
  D = 3,
  init_seed = 2,
  mf_max_iters = 60,
  mf_nmf_iters = 100,
  miso_max_iters = 25,
  n_inner = 5
)

summarize_pair(tree_pair, tree)
```

```{r tree-motifs}
rbind(
  motif_table(tree_pair$frozen, tree$F0, "frozen"),
  motif_table(tree_pair$alternating, tree$F0, "alternating")
)
```

```{r tree-plots}
plot_group_responsibilities(
  tree_pair$frozen$omega,
  tree$grp,
  "Tree scenario: frozen MiSo responsibilities"
)

plot_group_responsibilities(
  tree_pair$alternating$omega,
  tree$grp,
  "Tree scenario: alternating MiSo responsibilities"
)

plot_cosine_heatmap(
  match_factor_rows(tree_pair$alternating$F, tree$F0)$similarity,
  "Tree scenario: alternating fitted and true factor cosine"
)
```

## Takeaway

This is still a hard-motif coordinate method. A useful result here is not that
the ELBO is perfectly monotone, but that updating `F` and motifs from a good MF
initialization improves or preserves clustering and gives more coherent
submanifold motifs. If alternating hard motifs starts to reinforce a bad
initialization, the next step is the soft submanifold motif posterior from the
full model.
