---
title: "MiSo Canonical Multi-Seed Recovery Benchmark"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

This benchmark evaluates whether MiSo recovers factors, loading supports,
and motifs when the factor count is correct and the factor matrix is learned.
It includes a block scenario with single-factor anchors and a harder tree
scenario with no single-factor motif. All methods are fitted to the same 80%
Poisson-thinned training counts and evaluated on the remaining 20% of counts.

```{r source-code}
source("code/miso-benchmark-utils.R")
```

## Benchmark design

```{r design}
n_seed = as.integer(Sys.getenv("MISO_BENCHMARK_SEEDS", "20"))
benchmark_seeds = seq_len(n_seed)
train_fraction = 0.8

block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

data.frame(
  scenario = c("block", "tree/no-anchor"),
  K = c(5, 7),
  S = c(length(block_supports), length(tree_supports)),
  fitted_D = c(3, 3),
  true_support_size = c("1 or 3", "2 or 3"),
  N = c(240, 240),
  M = c(250, 280),
  n_seed = n_seed,
  train_fraction = train_fraction
)
```

The comparison includes Poisson NMF followed by k-means,
MF-Poisson-SuSiE followed by k-means, and MiSo. The two baseline pipelines
do not produce motif responsibilities, so the soft motif assignment score is
reported only for MiSo.

## Run fits

Each seed is cached separately, making an interrupted run resumable. Set the
environment variables `MISO_BENCHMARK_SEEDS` or
`MISO_BENCHMARK_CORES` to change the run size or local parallelism.

```{r run-helpers}
cache_version = "v1"
cache_dir = file.path("analysis", "cache",
                      paste0("miso-canonical-", cache_version))
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

run_scenario_seed = function(scenario, seed) {
  cache_file = file.path(cache_dir, sprintf("%s-seed-%03d.rds", scenario,
                                            seed))
  if (file.exists(cache_file)) {
    out = readRDS(cache_file)
    baseline_methods = c("Poisson NMF + kmeans",
                         "MF-Poisson-SuSiE + kmeans")
    out$method[!out$method %in% baseline_methods] = "MiSo"
    return(out)
  }

  if (scenario == "block") {
    dat = simulate_support_scenario(
      N = 240,
      M = 250,
      K = 5,
      group_supports = block_supports,
      shape_vec = c(rep(10, 5), rep(20, 3)),
      seed = 1000 + seed,
      train_fraction = train_fraction
    )
    out = run_one_benchmark(
      dat = dat,
      K = 5,
      S = length(block_supports),
      D = 3,
      seed = seed,
      nmf_iters = 50,
      mf_max_iters = 25,
      mf_nmf_iters = 45,
      miso_max_iters = 12,
      miso_n_inner = 3,
      block_size = 100
    )
  } else if (scenario == "tree") {
    dat = simulate_support_scenario(
      N = 240,
      M = 280,
      K = 7,
      group_supports = tree_supports,
      shape_vec = rep(18, length(tree_supports)),
      seed = 2000 + seed,
      train_fraction = train_fraction
    )
    out = run_one_benchmark(
      dat = dat,
      K = 7,
      S = length(tree_supports),
      D = 3,
      seed = seed,
      nmf_iters = 50,
      mf_max_iters = 25,
      mf_nmf_iters = 45,
      miso_max_iters = 12,
      miso_n_inner = 3,
      block_size = 100
    )
  } else {
    stop("Unknown scenario")
  }

  out$scenario = scenario
  out$seed = seed
  saveRDS(out, cache_file)
  out
}
```

```{r run-benchmark}
jobs = expand.grid(
  scenario = c("block", "tree"),
  seed = benchmark_seeds,
  stringsAsFactors = FALSE
)

run_job = function(j) {
  run_scenario_seed(jobs$scenario[j], jobs$seed[j])
}

requested_cores = as.integer(Sys.getenv("MISO_BENCHMARK_CORES", "4"))
n_cores = max(1L, min(requested_cores, parallel::detectCores(), nrow(jobs)))
if (.Platform$OS.type == "windows" || n_cores == 1L) {
  result_list = lapply(seq_len(nrow(jobs)), run_job)
} else {
  result_list = parallel::mclapply(seq_len(nrow(jobs)), run_job,
                                   mc.cores = n_cores)
}
benchmark_results = do.call(rbind, result_list)
rownames(benchmark_results) = NULL

saveRDS(benchmark_results,
        file.path(cache_dir, sprintf("combined-%03d-seeds.rds", n_seed)))
```

## Recovery and prediction results

```{r summaries}
summary_results = summarize_benchmark_results(benchmark_results)

compact_summary = summary_results[, c(
  "scenario",
  "method",
  "n_seed",
  "cluster_accuracy_mean",
  "cluster_accuracy_sd",
  "soft_cluster_accuracy_mean",
  "soft_cluster_accuracy_sd",
  "support_accuracy_mean",
  "support_accuracy_sd",
  "support_jaccard_mean",
  "support_jaccard_sd",
  "mean_factor_cosine_mean",
  "mean_factor_cosine_sd",
  "heldout_deviance_per_entry_mean",
  "heldout_deviance_per_entry_sd"
)]

numeric_columns = vapply(compact_summary, is.numeric, logical(1))
numeric_columns[match("n_seed", names(compact_summary))] = FALSE
compact_summary[numeric_columns] = round(compact_summary[numeric_columns], 4)
compact_summary
```

The exact support metric selects the known true number of active factors for
each observation and asks whether the complete support is correct. Support
Jaccard gives partial credit when only part of the support is recovered. Lower
held-out Poisson deviance is better.

## Paired comparisons with MiSo

Because all methods use the same simulated datasets, method differences should
be paired by seed. Positive differences favor MiSo for recovery metrics;
negative differences favor MiSo for held-out deviance.

```{r paired-comparisons}
paired_difference = function(results, scenario_name, baseline, metric) {
  tab = results[results$scenario == scenario_name &
                  results$method %in% c("MiSo", baseline),
                c("seed", "method", metric)]
  wide = reshape(tab, idvar = "seed", timevar = "method", direction = "wide")
  difference = wide[[paste0(metric, ".MiSo")]] -
    wide[[paste0(metric, ".", baseline)]]
  data.frame(
    scenario = scenario_name,
    baseline = baseline,
    metric = metric,
    mean_difference = mean(difference),
    lower_95 = mean(difference) - qt(0.975, length(difference) - 1) *
      sd(difference) / sqrt(length(difference)),
    upper_95 = mean(difference) + qt(0.975, length(difference) - 1) *
      sd(difference) / sqrt(length(difference))
  )
}

baselines = c("Poisson NMF + kmeans", "MF-Poisson-SuSiE + kmeans")
comparison_metrics = c(
  "cluster_accuracy",
  "support_accuracy",
  "support_jaccard",
  "mean_factor_cosine",
  "heldout_deviance_per_entry"
)

paired_results = do.call(rbind, lapply(c("block", "tree"), function(scenario) {
  do.call(rbind, lapply(baselines, function(baseline) {
    do.call(rbind, lapply(comparison_metrics, function(metric) {
      paired_difference(benchmark_results, scenario, baseline, metric)
    }))
  }))
}))
paired_results[, c("mean_difference", "lower_95", "upper_95")] =
  round(paired_results[, c("mean_difference", "lower_95", "upper_95")], 4)
paired_results
```

## Visual summary

```{r plotting-helpers}
plot_metric_by_seed = function(results, metric, main, ylim = NULL) {
  scenarios = unique(results$scenario)
  op = par(mfrow = c(1, length(scenarios)), mar = c(7, 4, 3, 1))
  on.exit(par(op))

  for (scenario_name in scenarios) {
    tab = results[results$scenario == scenario_name, ]
    values = split(tab[[metric]], tab$method)
    boxplot(values, las = 2, ylim = ylim, ylab = metric,
            main = paste(main, scenario_name), col = "gray85")
    stripchart(values, vertical = TRUE, method = "jitter", pch = 19,
               add = TRUE, col = "firebrick")
  }
}
```

```{r recovery-plots}
plot_metric_by_seed(benchmark_results, "cluster_accuracy",
                    "Hard clustering:", ylim = c(0, 1))
plot_metric_by_seed(benchmark_results, "support_accuracy",
                    "Exact support:", ylim = c(0, 1))
plot_metric_by_seed(benchmark_results, "support_jaccard",
                    "Support Jaccard:", ylim = c(0, 1))
plot_metric_by_seed(benchmark_results, "mean_factor_cosine",
                    "Factor recovery:", ylim = c(0, 1))
plot_metric_by_seed(benchmark_results, "heldout_deviance_per_entry",
                    "Held-out deviance:")
```

## Observed results

```{r result-values, include=FALSE}
summary_row = function(scenario_name, method_name) {
  summary_results[summary_results$scenario == scenario_name &
                    summary_results$method == method_name, ]
}

block_miso = summary_row("block", "MiSo")
block_nmf = summary_row("block", "Poisson NMF + kmeans")
block_mf = summary_row("block", "MF-Poisson-SuSiE + kmeans")
tree_miso = summary_row("tree", "MiSo")
tree_nmf = summary_row("tree", "Poisson NMF + kmeans")
tree_mf = summary_row("tree", "MF-Poisson-SuSiE + kmeans")

paired_row = function(scenario_name, baseline_name, metric_name) {
  paired_results[paired_results$scenario == scenario_name &
                   paired_results$baseline == baseline_name &
                   paired_results$metric == metric_name, ]
}
```

In the block scenario, MiSo obtains mean hard clustering accuracy
`r sprintf("%.3f", block_miso$cluster_accuracy_mean)`, exact support accuracy
`r sprintf("%.3f", block_miso$support_accuracy_mean)`, and mean factor cosine
`r sprintf("%.3f", block_miso$mean_factor_cosine_mean)`. The corresponding
MF-Poisson-SuSiE values are
`r sprintf("%.3f", block_mf$cluster_accuracy_mean)`,
`r sprintf("%.3f", block_mf$support_accuracy_mean)`, and
`r sprintf("%.3f", block_mf$mean_factor_cosine_mean)`. The paired 95% interval
for the exact-support improvement over MF-Poisson-SuSiE is
[`r sprintf("%.3f", paired_row("block", "MF-Poisson-SuSiE + kmeans", "support_accuracy")$lower_95)`,
`r sprintf("%.3f", paired_row("block", "MF-Poisson-SuSiE + kmeans", "support_accuracy")$upper_95)`].
Thus the clean block result is positive, although much of the problem is
already solved by either baseline.

The tree/no-anchor scenario provides the clearer structural result. MiSo
raises exact support accuracy to
`r sprintf("%.3f", tree_miso$support_accuracy_mean)`, compared with
`r sprintf("%.3f", tree_nmf$support_accuracy_mean)` for Poisson NMF and
`r sprintf("%.3f", tree_mf$support_accuracy_mean)` for MF-Poisson-SuSiE.
Support Jaccard increases to
`r sprintf("%.3f", tree_miso$support_jaccard_mean)`, compared with
`r sprintf("%.3f", tree_nmf$support_jaccard_mean)` and
`r sprintf("%.3f", tree_mf$support_jaccard_mean)`. The paired exact-support
improvement over MF-Poisson-SuSiE is
`r sprintf("%.3f", paired_row("tree", "MF-Poisson-SuSiE + kmeans", "support_accuracy")$mean_difference)`
with a 95% interval of
[`r sprintf("%.3f", paired_row("tree", "MF-Poisson-SuSiE + kmeans", "support_accuracy")$lower_95)`,
`r sprintf("%.3f", paired_row("tree", "MF-Poisson-SuSiE + kmeans", "support_accuracy")$upper_95)`].

This support gain is not explained by a better learned dictionary. Mean factor
cosine in the tree scenario is
`r sprintf("%.3f", tree_miso$mean_factor_cosine_mean)` for MiSo,
`r sprintf("%.3f", tree_nmf$mean_factor_cosine_mean)` for Poisson NMF, and
`r sprintf("%.3f", tree_mf$mean_factor_cosine_mean)` for MF-Poisson-SuSiE.
The similar factor recovery, together with materially better support recovery,
is evidence that sharing a posterior motif across observations adds useful
structure after learning $F$.

Predictive performance does not reveal a tradeoff for the structural gain.
MiSo has the lowest mean held-out deviance in both scenarios. In the tree
setting it is `r sprintf("%.4f", tree_miso$heldout_deviance_per_entry_mean)`,
versus `r sprintf("%.4f", tree_nmf$heldout_deviance_per_entry_mean)` for
Poisson NMF and `r sprintf("%.4f", tree_mf$heldout_deviance_per_entry_mean)`
for MF-Poisson-SuSiE. The improvement over Poisson NMF is clear across seeds;
the paired interval against MF-Poisson-SuSiE reaches approximately zero, so
that smaller predictive difference should not be emphasized.

The result is not uniformly easy across simulation draws. MiSo exact
support accuracy ranges from
`r sprintf("%.3f", min(benchmark_results$support_accuracy[benchmark_results$scenario == "tree" & benchmark_results$method == "MiSo"]))`
to
`r sprintf("%.3f", max(benchmark_results$support_accuracy[benchmark_results$scenario == "tree" & benchmark_results$method == "MiSo"]))`
in the tree setting. This variation is consistent with a nonconvex learned-$F$
problem and argues for reporting all seeds and, later, an initialization
sensitivity experiment.

Finally, hard and soft motif assignment scores are nearly identical, and the
mean maximum responsibility is
`r sprintf("%.3f", block_miso$mean_max_responsibility_mean)` in the block
scenario and `r sprintf("%.3f", tree_miso$mean_max_responsibility_mean)` in the
tree scenario. These correctly specified simulations therefore show
concentrated motif assignments. They are evidence for recovery, not for
uncertainty calibration.

## Interpretation guide

The block scenario tests ordinary sparse recovery when some motifs provide
single-factor anchors. The tree scenario is the more diagnostic MiSo test:
every motif contains two or three factors, factor 1 is shared globally, and
the remaining factors define two branches. A useful motif model should gain
most clearly in exact support recovery there.

The hard clustering score uses the maximum posterior responsibility for
MiSo. Its soft score averages the posterior mass on the correctly aligned
true motif, so a large gap between hard and soft scores indicates assignment
uncertainty. Factor cosine evaluates the learned dictionary independently of
motif labels. Held-out deviance checks whether improved structural recovery is
obtained without sacrificing count prediction.

This experiment evaluates recovery under a favorable, correctly specified
setting. It does not yet establish robustness to initialization, unknown
dimension, unknown motif count, correlated factors, or model misspecification.
Here $D=3$ is the correct maximum motif dimension, but motifs with only one
or two active factors still contain surplus dimensions under the current
common-$D$ fit. The separate over-specified-$D$ experiment will test whether
empirical-Bayes shrinkage can identify those motif-specific dimensions.
