---
title: "MoS-NMF Multi-Seed Uncertainty Benchmark"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This notebook repeats the uncertainty stress tests over several random seeds.
The goal is narrower than the recovery benchmark: here we ask whether
`gamma_bar[s,d,k]` represents posterior uncertainty when factor identities are
ambiguous. To isolate this question, the factor dictionary is held fixed during
MoS fitting.

```{r source-code}
source("code/mos-nmf.R")
```

## Helpers

```{r helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

support_label = function(support) paste(sort(unique(support)), collapse = ",")

make_sparse_factors = function(K, M, shape = 0.1, rate = 0.01, seed = 1) {
  set.seed(seed)
  F0 = matrix(rgamma(K * M, shape = shape, rate = rate), nrow = K, ncol = M)
  normalize_rows(F0)
}

make_correlated_factors = function(K, M, correlated_pair = c(1, 2),
                                   mix = 0.96, seed = 1) {
  F0 = make_sparse_factors(K, M, seed = seed)
  a = correlated_pair[1]
  b = correlated_pair[2]
  F0[b, ] = normalize_rows(matrix(mix * F0[a, ] + (1 - mix) * F0[b, ],
                                  nrow = 1))
  F0
}

simulate_from_fixed_F = function(N, F0, group_supports, shape_vec, seed = 1) {
  set.seed(seed)
  K = nrow(F0)
  S = length(group_supports)
  stopifnot(N %% S == 0)
  n_per_group = N / S
  grp = rep(seq_len(S), each = n_per_group)

  L = matrix(0, nrow = N, ncol = K)
  for (s in seq_len(S)) {
    idx = which(grp == s)
    support = group_supports[[s]]
    L[idx, support] = matrix(
      rgamma(length(idx) * length(support), shape = shape_vec[s], rate = 1),
      nrow = length(idx),
      ncol = length(support)
    )
  }

  Y = matrix(rpois(N * ncol(F0), L %*% F0), nrow = N, ncol = ncol(F0))

  list(
    Y = Y,
    F0 = F0,
    L = L,
    grp = grp,
    group_supports = group_supports,
    true_patterns = vapply(group_supports, support_label, character(1))
  )
}

fit_mos_fixed_dictionary = function(dat, F_init, S, D, init_seed,
                                    max_iters = 16, n_inner = 3,
                                    block_size = 100, gamma_floor = 0.30) {
  set.seed(init_seed)
  nmf_init = poisson_nmf_init(dat$Y, K = nrow(F_init), max_iters = 50,
                              init_seed = init_seed)
  gamma_init = init_soft_gamma_from_loading_scores(
    scores = nmf_init$L,
    S = S,
    D = D,
    init_seed = init_seed,
    gamma_floor = gamma_floor
  )

  fit = mos_nmf_fixed_gamma(
    Y = dat$Y,
    F = F_init,
    gamma_bar = gamma_init$gamma_bar,
    max_iters = max_iters,
    n_inner = n_inner,
    update_prior = TRUE,
    update_F = FALSE,
    update_gamma = TRUE,
    gamma_step_init = 0.25,
    gamma_step_ramp = 10,
    tol = 1e-5,
    min_iters = 6,
    patience = 2,
    block_size = block_size
  )
  fit$gamma_init = gamma_init
  fit
}

gamma_slot_table = function(fit, seed, scenario) {
  ent = motif_gamma_entropy(fit$gamma_bar)
  top = apply(fit$gamma_bar, c(1, 2), max)
  out = do.call(rbind, lapply(seq_len(dim(fit$gamma_bar)[1]), function(s) {
    do.call(rbind, lapply(seq_len(dim(fit$gamma_bar)[2]), function(d) {
      p = fit$gamma_bar[s, d, ]
      data.frame(
        scenario = scenario,
        seed = seed,
        submanifold = s,
        slot = d,
        top_factor = which.max(p),
        top_prob = top[s, d],
        entropy = ent[s, d]
      )
    }))
  }))
  rownames(out) = NULL
  out
}

summarize_values = function(x) {
  c(mean = mean(x), sd = sd(x), median = median(x),
    q25 = unname(quantile(x, 0.25)), q75 = unname(quantile(x, 0.75)),
    max = max(x))
}

summarize_by_scenario = function(tab, metric) {
  out = do.call(rbind, lapply(split(tab, tab$scenario), function(x) {
    vals = summarize_values(x[[metric]])
    data.frame(
      scenario = unique(x$scenario),
      metric = metric,
      t(vals)
    )
  }))
  rownames(out) = NULL
  out
}

plot_metric_by_seed = function(tab, metric, main) {
  vals = split(tab[[metric]], tab$seed)
  boxplot(vals, ylim = c(0, 1), xlab = "Seed", ylab = metric,
          main = main, col = "gray85")
  stripchart(vals, vertical = TRUE, method = "jitter", pch = 19,
             col = "firebrick", add = TRUE)
}

plot_slot_heatmap = function(fit, slots, main) {
  mat = do.call(rbind, lapply(seq_len(nrow(slots)), function(j) {
    fit$gamma_bar[slots$submanifold[j], slots$slot[j], ]
  }))
  row_lab = paste0("S", slots$submanifold, ":d", slots$slot)

  image(x = seq_len(ncol(mat)), y = seq_len(nrow(mat)),
        z = t(mat), axes = FALSE, xlab = "Candidate factor",
        ylab = "Motif slot", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(mat)), labels = paste0("F", seq_len(ncol(mat))))
  axis(2, at = seq_len(nrow(mat)), labels = row_lab, las = 1)
  for (i in seq_len(nrow(mat))) {
    for (k in seq_len(ncol(mat))) {
      text(k, i, sprintf("%.2f", mat[i, k]), cex = 0.65)
    }
  }
  box()
}
```

## Design

```{r design}
benchmark_seeds = 1:3

over_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

corr_supports = list(
  c(1),
  c(2),
  c(3),
  c(1, 3),
  c(2, 3),
  c(1, 2)
)

data.frame(
  scenario = c("overspecified K duplicate dictionary",
               "correlated factors"),
  N = c(240, 180),
  M = c(260, 260),
  K_true = c(5, 5),
  K_fit = c(7, 5),
  S = c(length(over_supports), length(corr_supports)),
  D = c(3, 2),
  n_seed = length(benchmark_seeds)
)
```

## Overspecified `K`: Duplicate Candidate Factors

The data are generated from five true factors, but the fitted dictionary has
seven rows. Rows 6 and 7 are near-duplicates of true factors 1 and 2. Posterior
mass split inside pairs `(1,6)` and `(2,7)` is the desired signal.

```{r overspecified-k-run}
run_over_seed = function(seed) {
  K_true = 5
  M = 260
  F_true = make_sparse_factors(K_true, M, seed = 1000 + seed)
  dup1 = normalize_rows(matrix(
    0.94 * F_true[1, ] + 0.06 * make_sparse_factors(1, M, seed = 1100 + seed),
    nrow = 1
  ))
  dup2 = normalize_rows(matrix(
    0.94 * F_true[2, ] + 0.06 * make_sparse_factors(1, M, seed = 1200 + seed),
    nrow = 1
  ))
  F_over = rbind(F_true, dup1, dup2)

  dat = simulate_from_fixed_F(
    N = 240,
    F0 = F_true,
    group_supports = over_supports,
    shape_vec = c(rep(8, 5), rep(14, 3)),
    seed = 1300 + seed
  )

  fit = fit_mos_fixed_dictionary(
    dat = dat,
    F_init = F_over,
    S = length(over_supports),
    D = 3,
    init_seed = 1400 + seed,
    max_iters = 16,
    n_inner = 3,
    block_size = 100,
    gamma_floor = 0.30
  )

  slot_tab = gamma_slot_table(
    fit = fit,
    seed = seed,
    scenario = "overspecified K"
  )

  pair_tab = do.call(rbind, lapply(seq_len(dim(fit$gamma_bar)[1]), function(s) {
    do.call(rbind, lapply(seq_len(dim(fit$gamma_bar)[2]), function(d) {
      p = fit$gamma_bar[s, d, ]
      data.frame(
        scenario = "overspecified K",
        seed = seed,
        submanifold = s,
        slot = d,
        mass_true1_or_dup = p[1] + p[6],
        split_true1_vs_dup = min(p[1], p[6]),
        mass_true2_or_dup = p[2] + p[7],
        split_true2_vs_dup = min(p[2], p[7]),
        max_duplicate_split = max(min(p[1], p[6]), min(p[2], p[7])),
        cluster_accuracy = cluster_accuracy(fit$z_hat, dat$grp),
        mean_max_responsibility = mean(apply(fit$omega, 1, max))
      )
    }))
  }))

  list(
    fit = fit,
    F_true = F_true,
    F_over = F_over,
    dat = dat,
    slot_tab = slot_tab,
    pair_tab = pair_tab
  )
}

over_runs = lapply(benchmark_seeds, run_over_seed)
over_slot_tab = do.call(rbind, lapply(over_runs, `[[`, "slot_tab"))
over_pair_tab = do.call(rbind, lapply(over_runs, `[[`, "pair_tab"))

head(over_pair_tab[order(-over_pair_tab$max_duplicate_split), ], 12)
```

```{r overspecified-k-summary}
rbind(
  summarize_by_scenario(over_slot_tab, "entropy"),
  summarize_by_scenario(over_slot_tab, "top_prob"),
  summarize_by_scenario(over_pair_tab, "max_duplicate_split"),
  summarize_by_scenario(over_pair_tab, "cluster_accuracy"),
  summarize_by_scenario(over_pair_tab, "mean_max_responsibility")
)
```

```{r overspecified-k-plots}
op = par(mfrow = c(1, 2))
plot_metric_by_seed(over_slot_tab, "entropy",
                    "Overspecified K: gamma_bar entropy")
plot_metric_by_seed(over_pair_tab, "max_duplicate_split",
                    "Overspecified K: duplicate-pair split")
par(op)
```

```{r overspecified-k-example-heatmap}
example_over = over_runs[[which.max(vapply(
  over_runs,
  function(x) max(x$pair_tab$max_duplicate_split),
  numeric(1)
))]]
example_slots = example_over$pair_tab[
  order(-example_over$pair_tab$max_duplicate_split),
][seq_len(10), ]
plot_slot_heatmap(
  example_over$fit,
  example_slots,
  "Overspecified K: slots with largest duplicate split"
)
```

## Correlated True Factors

Here `K` is correct, but true factors 1 and 2 are nearly collinear. The desired
targeted uncertainty signal is posterior mass concentrated on `{1,2}` with a
non-negligible split between the two.

```{r correlated-run}
run_corr_seed = function(seed) {
  K = 5
  M = 260
  F_corr = make_correlated_factors(K, M, correlated_pair = c(1, 2),
                                   mix = 0.96, seed = 2000 + seed)
  dat = simulate_from_fixed_F(
    N = 180,
    F0 = F_corr,
    group_supports = corr_supports,
    shape_vec = c(5, 5, 8, 10, 10, 10),
    seed = 2100 + seed
  )

  fit = fit_mos_fixed_dictionary(
    dat = dat,
    F_init = F_corr,
    S = length(corr_supports),
    D = 2,
    init_seed = 2200 + seed,
    max_iters = 16,
    n_inner = 3,
    block_size = 100,
    gamma_floor = 0.30
  )

  slot_tab = gamma_slot_table(
    fit = fit,
    seed = seed,
    scenario = "correlated factors"
  )

  pair_tab = do.call(rbind, lapply(seq_len(dim(fit$gamma_bar)[1]), function(s) {
    do.call(rbind, lapply(seq_len(dim(fit$gamma_bar)[2]), function(d) {
      p = fit$gamma_bar[s, d, ]
      data.frame(
        scenario = "correlated factors",
        seed = seed,
        submanifold = s,
        slot = d,
        mass_factor_1_or_2 = p[1] + p[2],
        split_between_1_and_2 = min(p[1], p[2]),
        targeted_pair_uncertainty =
          (p[1] + p[2]) * (2 * min(p[1], p[2])),
        cluster_accuracy = cluster_accuracy(fit$z_hat, dat$grp),
        mean_max_responsibility = mean(apply(fit$omega, 1, max))
      )
    }))
  }))

  list(
    fit = fit,
    F_corr = F_corr,
    dat = dat,
    slot_tab = slot_tab,
    pair_tab = pair_tab
  )
}

corr_runs = lapply(benchmark_seeds, run_corr_seed)
corr_slot_tab = do.call(rbind, lapply(corr_runs, `[[`, "slot_tab"))
corr_pair_tab = do.call(rbind, lapply(corr_runs, `[[`, "pair_tab"))

head(corr_pair_tab[order(-corr_pair_tab$targeted_pair_uncertainty), ], 12)
```

```{r correlated-summary}
rbind(
  summarize_by_scenario(corr_slot_tab, "entropy"),
  summarize_by_scenario(corr_slot_tab, "top_prob"),
  summarize_by_scenario(corr_pair_tab, "mass_factor_1_or_2"),
  summarize_by_scenario(corr_pair_tab, "split_between_1_and_2"),
  summarize_by_scenario(corr_pair_tab, "targeted_pair_uncertainty"),
  summarize_by_scenario(corr_pair_tab, "cluster_accuracy"),
  summarize_by_scenario(corr_pair_tab, "mean_max_responsibility")
)
```

```{r correlated-plots}
op = par(mfrow = c(1, 3))
plot_metric_by_seed(corr_slot_tab, "entropy",
                    "Correlated factors: gamma_bar entropy")
plot_metric_by_seed(corr_pair_tab, "mass_factor_1_or_2",
                    "Correlated factors: mass on {1,2}")
plot_metric_by_seed(corr_pair_tab, "targeted_pair_uncertainty",
                    "Correlated factors: targeted uncertainty")
par(op)
```

```{r correlated-example-heatmap}
example_corr = corr_runs[[which.max(vapply(
  corr_runs,
  function(x) max(x$pair_tab$targeted_pair_uncertainty),
  numeric(1)
))]]
example_corr_slots = example_corr$pair_tab[
  order(-example_corr$pair_tab$targeted_pair_uncertainty),
][seq_len(10), ]
plot_slot_heatmap(
  example_corr$fit,
  example_corr_slots,
  "Correlated factors: slots with largest targeted uncertainty"
)
```

## Interpretation

The overspecified-`K` experiment asks a clean uncertainty question because the
extra rows are explicitly constructed as near-duplicates. In this setting, raw
entropy is useful, but duplicate-pair split is more direct. A motif slot can
have low global entropy and still be scientifically uncertain if its posterior
mass is split mostly between an original factor and its duplicate.

The correlated-factor experiment is harder to interpret. High entropy by itself
can mean either targeted ambiguity between factors 1 and 2 or a weak/unidentified
motif slot. The targeted metric
`(mass on factors 1 or 2) * (2 * smaller posterior mass within the pair)` is
therefore a better paper diagnostic than entropy alone: it is large only when
the slot assigns substantial total mass to the correlated pair and does not
choose one factor decisively.

These results should be read alongside the clean multi-seed recovery benchmark.
In clean settings, `gamma_bar` should concentrate. In deliberately ambiguous
settings, the evidence for the uncertainty claim should come from pair-specific
posterior split metrics, not from diffuse posteriors everywhere.
