---
title: "Soft MoS-NMF With Overspecified D"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This page tests whether soft MoS-NMF behaves reasonably when the fitted
submanifold dimension `D` is larger than the true motif dimension. The desired
behavior is that extra motif slots carry small posterior loading size, or at
least contribute little to recovered support patterns.

We compare two `D = 5` initializations and one explicit slot-scale variant. The
original initialization pads short motifs by repeating active slots. The
experimental diffuse-surplus initialization leaves repeated surplus slots nearly
uniform over factors. The slot-scale variant additionally learns nonnegative
usage multipliers for each motif slot.

```{r source-code}
source("code/mos-nmf.R")
```

## Helpers

```{r helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)
  perms = all_permutations(seq_len(K))
  scores = apply(perms, 1, function(p) {
    sum(similarity[cbind(seq_len(K), p)])
  })
  learned_to_true = perms[which.max(scores), ]

  list(
    table = data.frame(
      learned_factor = seq_len(K),
      matched_true_factor = learned_to_true,
      cosine = similarity[cbind(seq_len(K), learned_to_true)]
    ),
    learned_to_true = learned_to_true,
    similarity = similarity
  )
}

support_label = function(support) paste(sort(unique(support)), collapse = ",")

simulate_support_scenario = function(N, M, K, group_supports, shape_vec,
                                      factor_shape = 0.1,
                                      factor_rate = 0.01, seed = 1) {
  set.seed(seed)
  S = length(group_supports)
  n_per_group = N / S
  grp = rep(seq_len(S), each = n_per_group)

  F0 = matrix(rgamma(M * K, shape = factor_shape, rate = factor_rate),
              nrow = K, ncol = M)
  F0 = normalize_rows(F0)

  L = matrix(0, nrow = N, ncol = K)
  for (s in seq_len(S)) {
    idx = which(grp == s)
    support = group_supports[[s]]
    L[idx, support] = matrix(
      rgamma(length(idx) * length(support), shape = shape_vec[s], rate = 1),
      nrow = length(idx),
      ncol = length(support)
    )
  }

  Y = matrix(rpois(N * M, L %*% F0), nrow = N, ncol = M)

  list(
    Y = Y,
    F0 = F0,
    L = L,
    grp = grp,
    group_supports = group_supports,
    true_support_size = rowSums(L > 0),
    true_support_label = apply(L > 0, 1, function(x) {
      paste(which(x), collapse = ",")
    }),
    true_patterns = vapply(group_supports, support_label, character(1))
  )
}

map_scores_to_true = function(scores, learned_to_true) {
  K = length(learned_to_true)
  out = matrix(0, nrow = nrow(scores), ncol = K)
  for (k in seq_len(K)) {
    out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
  }
  out
}

support_labels_from_scores = function(scores_true_space, support_sizes) {
  vapply(seq_len(nrow(scores_true_space)), function(i) {
    active = order(scores_true_space[i, ], decreasing = TRUE)[
      seq_len(support_sizes[i])
    ]
    paste(sort(active), collapse = ",")
  }, character(1))
}

support_accuracy = function(fit, dat) {
  factor_match = match_factor_rows(fit$F, dat$F0)
  scores = fit$omega %*% motif_factor_scores(fit)$scores
  scores_true = map_scores_to_true(scores, factor_match$learned_to_true)

  mean(support_labels_from_scores(scores_true, dat$true_support_size) ==
         dat$true_support_label)
}

group_responsibility = function(omega, grp) {
  do.call(rbind, lapply(sort(unique(grp)), function(g) {
    colMeans(omega[grp == g, , drop = FALSE])
  }))
}

match_submanifolds_to_groups = function(omega, grp) {
  group_resp = group_responsibility(omega, grp)
  apply(group_resp, 2, which.max)
}

slot_diagnostics = function(fit, dat, label) {
  beta_mean = motif_factor_scores(fit)$beta_mean
  if (is.null(fit$slot_scale)) {
    slot_scale = matrix(1, nrow(beta_mean), ncol(beta_mean))
  } else {
    slot_scale = fit$slot_scale
  }
  effective_beta = beta_mean * slot_scale
  prior_mean = fit$alpha0 / pmax(fit$lambda0, 1e-12)
  ent = motif_gamma_entropy(fit$gamma_bar)
  top = apply(fit$gamma_bar, c(1, 2), max)
  matched_group = match_submanifolds_to_groups(fit$omega, dat$grp)

  out = do.call(rbind, lapply(seq_len(nrow(beta_mean)), function(s) {
    ord = order(effective_beta[s, ], decreasing = TRUE)
    true_dim = length(dat$group_supports[[matched_group[s]]])
    prior_fraction = prior_mean[s, ] / pmax(sum(prior_mean[s, ]), 1e-12)
    data.frame(
      fit = label,
      submanifold = s,
      matched_group = matched_group[s],
      true_dim = true_dim,
      slot = ord,
      slot_rank = seq_along(ord),
      beta_mean = beta_mean[s, ord],
      slot_scale = slot_scale[s, ord],
      effective_beta = effective_beta[s, ord],
      beta_fraction = effective_beta[s, ord] / sum(effective_beta[s, ]),
      prior_mean = prior_mean[s, ord],
      prior_fraction = prior_fraction[ord],
      extra_slot = seq_along(ord) > true_dim,
      gamma_entropy = ent[s, ord],
      top_gamma = top[s, ord]
    )
  }))

  rownames(out) = NULL
  out
}

extra_slot_summary = function(slot_tab) {
  do.call(rbind, lapply(split(slot_tab, slot_tab$fit), function(tab) {
    extra_by_submanifold = tapply(seq_len(nrow(tab)), tab$submanifold,
                                  function(idx) {
      sum(tab$beta_fraction[idx][tab$extra_slot[idx]])
    })
    data.frame(
      fit = unique(tab$fit),
      mean_extra_beta_fraction = mean(extra_by_submanifold),
      total_extra_beta_fraction = sum(tab$beta_fraction[tab$extra_slot]) /
        length(unique(tab$submanifold)),
      total_extra_prior_fraction = sum(tab$prior_fraction[tab$extra_slot]) /
        length(unique(tab$submanifold)),
      mean_extra_gamma_entropy = mean(tab$gamma_entropy[tab$extra_slot]),
      mean_extra_top_gamma = mean(tab$top_gamma[tab$extra_slot])
    )
  }))
}

summarize_fit = function(fit, dat, label) {
  factor_match = match_factor_rows(fit$F, dat$F0)
  ent = motif_gamma_entropy(fit$gamma_bar)
  top = apply(fit$gamma_bar, c(1, 2), max)

  data.frame(
    fit = label,
    cluster_accuracy = cluster_accuracy(fit$z_hat, dat$grp),
    support_accuracy = support_accuracy(fit, dat),
    mean_factor_cosine = mean(factor_match$table$cosine),
    min_factor_cosine = min(factor_match$table$cosine),
    mean_max_responsibility = mean(apply(fit$omega, 1, max)),
    mean_gamma_entropy = mean(ent),
    max_gamma_entropy = max(ent),
    mean_top_gamma = mean(top),
    min_top_gamma = min(top)
  )
}

plot_group_responsibilities = function(omega, grp, main) {
  group_resp = group_responsibility(omega, grp)
  image(x = seq_len(ncol(group_resp)), y = seq_len(nrow(group_resp)),
        z = t(group_resp), axes = FALSE, xlab = "Fitted submanifold",
        ylab = "True group", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(group_resp)),
       labels = paste0("S", seq_len(ncol(group_resp))))
  axis(2, at = seq_len(nrow(group_resp)),
       labels = paste0("G", seq_len(nrow(group_resp))), las = 1)
  for (g in seq_len(nrow(group_resp))) {
    for (s in seq_len(ncol(group_resp))) {
      text(s, g, sprintf("%.2f", group_resp[g, s]), cex = 0.65)
    }
  }
  box()
}

plot_slot_fraction = function(slot_tab, main) {
  ordered = slot_tab[order(slot_tab$submanifold, slot_tab$slot_rank), ]
  mat = matrix(ordered$beta_fraction,
               nrow = length(unique(ordered$submanifold)), byrow = TRUE)
  image(x = seq_len(ncol(mat)), y = seq_len(nrow(mat)), z = t(mat),
        axes = FALSE, xlab = "Slot rank by beta mean",
        ylab = "Fitted submanifold", main = main,
        col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(mat)), labels = seq_len(ncol(mat)))
  axis(2, at = seq_len(nrow(mat)), labels = paste0("S", seq_len(nrow(mat))),
       las = 1)
  for (s in seq_len(nrow(mat))) {
    for (d in seq_len(ncol(mat))) {
      text(d, s, sprintf("%.2f", mat[s, d]), cex = 0.65)
    }
  }
  box()
}

plot_slot_entropy = function(slot_tab, main) {
  ordered = slot_tab[order(slot_tab$submanifold, slot_tab$slot_rank), ]
  mat = matrix(ordered$gamma_entropy,
               nrow = length(unique(ordered$submanifold)), byrow = TRUE)
  image(x = seq_len(ncol(mat)), y = seq_len(nrow(mat)), z = t(mat),
        axes = FALSE, xlab = "Slot rank by beta mean",
        ylab = "Fitted submanifold", main = main,
        col = hcl.colors(100, "Blues"))
  axis(1, at = seq_len(ncol(mat)), labels = seq_len(ncol(mat)))
  axis(2, at = seq_len(nrow(mat)), labels = paste0("S", seq_len(nrow(mat))),
       las = 1)
  for (s in seq_len(nrow(mat))) {
    for (d in seq_len(ncol(mat))) {
      text(d, s, sprintf("%.2f", mat[s, d]), cex = 0.65)
    }
  }
  box()
}

plot_slot_prior_fraction = function(slot_tab, main) {
  ordered = slot_tab[order(slot_tab$submanifold, slot_tab$slot_rank), ]
  mat = matrix(ordered$prior_fraction,
               nrow = length(unique(ordered$submanifold)), byrow = TRUE)
  image(x = seq_len(ncol(mat)), y = seq_len(nrow(mat)), z = t(mat),
        axes = FALSE, xlab = "Slot rank by beta mean",
        ylab = "Fitted submanifold", main = main,
        col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(mat)), labels = seq_len(ncol(mat)))
  axis(2, at = seq_len(nrow(mat)), labels = paste0("S", seq_len(nrow(mat))),
       las = 1)
  for (s in seq_len(nrow(mat))) {
    for (d in seq_len(ncol(mat))) {
      text(d, s, sprintf("%.2f", mat[s, d]), cex = 0.65)
    }
  }
  box()
}
```

## Simulate Data

```{r simulate}
block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

dat = simulate_support_scenario(
  N = 320,
  M = 350,
  K = 5,
  group_supports = block_supports,
  shape_vec = c(rep(10, 5), rep(20, 3)),
  seed = 52
)

data.frame(group = seq_along(block_supports),
           true_active_factors = dat$true_patterns)
```

## Fit Correct and Overspecified D

```{r fit-models}
fit_D3 = mos_nmf(
  Y = dat$Y,
  K = 5,
  S = length(block_supports),
  D = 3,
  max_iters = 18,
  n_inner = 3,
  mf_max_iters = 35,
  mf_nmf_iters = 60,
  init_seed = 3,
  update_prior = TRUE,
  update_F = TRUE,
  update_gamma = TRUE,
  gamma_init_floor = 0.05,
  gamma_step_init = 0.5,
  gamma_step_ramp = 8,
  F_step_init = 0.2,
  F_step_ramp = 15,
  tol = 1e-5,
  min_iters = 5,
  patience = 2,
  block_size = 100
)

fit_D5_repeat = mos_nmf(
  Y = dat$Y,
  K = 5,
  S = length(block_supports),
  D = 5,
  max_iters = 18,
  n_inner = 3,
  mf_max_iters = 35,
  mf_nmf_iters = 60,
  init_seed = 3,
  update_prior = TRUE,
  update_F = TRUE,
  update_gamma = TRUE,
  gamma_init_floor = 0.05,
  gamma_step_init = 0.5,
  gamma_step_ramp = 8,
  F_step_init = 0.2,
  F_step_ramp = 15,
  tol = 1e-5,
  min_iters = 5,
  patience = 2,
  block_size = 100
)

fit_D5_uniform = mos_nmf(
  Y = dat$Y,
  K = 5,
  S = length(block_supports),
  D = 5,
  max_iters = 18,
  n_inner = 3,
  mf_max_iters = 35,
  mf_nmf_iters = 60,
  init_seed = 3,
  update_prior = TRUE,
  update_F = TRUE,
  update_gamma = TRUE,
  gamma_init_floor = 0.05,
  surplus_slots = "uniform",
  gamma_step_init = 0.5,
  gamma_step_ramp = 8,
  F_step_init = 0.2,
  F_step_ramp = 15,
  tol = 1e-5,
  min_iters = 5,
  patience = 2,
  block_size = 100
)

fit_D5_scaled = mos_nmf(
  Y = dat$Y,
  K = 5,
  S = length(block_supports),
  D = 5,
  max_iters = 18,
  n_inner = 3,
  mf_max_iters = 35,
  mf_nmf_iters = 60,
  init_seed = 3,
  update_prior = TRUE,
  update_F = TRUE,
  update_gamma = TRUE,
  update_slot_scale = TRUE,
  gamma_init_floor = 0.05,
  surplus_slots = "uniform",
  slot_scale_step_init = 0.2,
  slot_scale_step_ramp = 8,
  gamma_step_init = 0.5,
  gamma_step_ramp = 8,
  F_step_init = 0.2,
  F_step_ramp = 15,
  tol = 1e-5,
  min_iters = 5,
  patience = 2,
  block_size = 100
)

rbind(
  summarize_fit(fit_D3, dat, "D = 3"),
  summarize_fit(fit_D5_repeat, dat, "D = 5 repeat surplus"),
  summarize_fit(fit_D5_uniform, dat, "D = 5 diffuse surplus"),
  summarize_fit(fit_D5_scaled, dat, "D = 5 diffuse + scale")
)
```

## Extra Slot Diagnostics

```{r slot-diagnostics}
slot_D3 = slot_diagnostics(fit_D3, dat, "D = 3")
slot_D5_repeat = slot_diagnostics(fit_D5_repeat, dat, "D = 5 repeat surplus")
slot_D5_uniform = slot_diagnostics(fit_D5_uniform, dat,
                                   "D = 5 diffuse surplus")
slot_D5_scaled = slot_diagnostics(fit_D5_scaled, dat,
                                  "D = 5 diffuse + scale")

extra_slot_summary(rbind(slot_D3, slot_D5_repeat, slot_D5_uniform,
                         slot_D5_scaled))
```

```{r ard-d-selection}
selection_with_truth = function(fit, label) {
  sel = select_motif_dimensions(fit)$selection
  matched_group = match_submanifolds_to_groups(fit$omega, dat$grp)
  sel$matched_group = matched_group[sel$submanifold]
  sel$true_dim = lengths(dat$group_supports)[sel$matched_group]
  data.frame(fit = label, sel[, c(
    "submanifold",
    "matched_group",
    "true_dim",
    "D_hat_ard",
    "D_hat_beta",
    "D_hat_ard_or_beta",
    "D_hat_diffuse_heuristic",
    "max_prior_fraction",
    "min_active_prior_fraction"
  )])
}

selection_summary = rbind(
  selection_with_truth(fit_D5_repeat, "D = 5 repeat surplus"),
  selection_with_truth(fit_D5_uniform, "D = 5 diffuse surplus"),
  selection_with_truth(fit_D5_scaled, "D = 5 diffuse + scale")
)

selection_summary
```

```{r top-extra-slots}
slot_D5_scaled[slot_D5_scaled$extra_slot, ][
  order(-slot_D5_scaled$beta_fraction[slot_D5_scaled$extra_slot]), ][
    seq_len(12), ]
```

```{r slot-plots}
plot_group_responsibilities(
  fit_D5_scaled$omega,
  dat$grp,
  "Overspecified D with slot scales: group responsibilities"
)

plot_slot_fraction(
  slot_D5_scaled,
  "Slot-scale fit: effective beta mass by ranked motif slot"
)

plot_slot_prior_fraction(
  slot_D5_scaled,
  "Slot-scale fit: EB prior mean fraction by ranked motif slot"
)

plot_slot_entropy(
  slot_D5_scaled,
  "Slot-scale fit: gamma entropy by ranked motif slot"
)
```

## Interpretation

This diagnostic distinguishes two ways an overspecified dimension can be
benign. The strongest behavior would be extra slots with both small posterior
loading mass and diffuse `gamma_bar`. A weaker but still useful behavior is
small posterior loading mass even if the extra slot's factor identity is
confident. In that case the extra slot is interpretable as inactive rather than
uncertain.

For the paper, the important quantity is therefore the beta-weighted motif
score, not only raw `gamma_bar`. If an extra slot has `top_gamma` near one but
`beta_fraction` near zero, it should not materially affect support recovery.
The empirical-Bayes prior mean `alpha0[s,d] / lambda0[s,d]` gives a second,
more SuSiE-like way to quantify the same idea. If the prior mean for a slot is
near zero relative to the other slots in submanifold `s`, then the fitted model
has effectively switched that dimension off.

The ARD-style rule used above counts a slot as active if its prior-mean
fraction within a submanifold is at least `0.05`. The diffuse-posterior rule is
more conservative: it treats a slot as inactive only when its posterior loading
fraction is below `0.05` and `gamma_bar[s,d,]` has high entropy. These two rules
answer different questions. ARD estimates how many loading dimensions each
motif uses; diffuse `gamma_bar` tells us whether an inactive or weak slot also
lacks a clear factor identity.

In the repeated-surplus run, the overfitted `D = 5` model does not show that
ideal behavior. Several extra slots have nontrivial beta fraction, and
`gamma_bar` collapses to point masses. This is a useful failure mode: with
repeated motif slots and exchangeable Gamma priors, duplicated slots can split
one true effect instead of being shrunk away.

The diffuse-surplus initialization tests whether this is mainly an
initialization artifact. If the extra beta fraction remains large, then
initialization alone is insufficient and overspecifying `D` is not automatically
benign for the current soft MoS-NMF implementation.

In this run, diffuse-surplus initialization largely fixes the extra-slot mass
problem: the extra beta fraction drops sharply and extra slots have higher
`gamma_bar` entropy. The tradeoff is that support accuracy is worse than with
the repeated-surplus initialization. This suggests that surplus-slot handling is
not merely a cosmetic initialization choice; it affects the fit's ability to
represent multi-factor motifs. The paper version should treat this as a model
selection/shrinkage problem rather than relying only on initialization.

The experimental slot-scale update pushes this further: extra slots have very
small effective beta mass and highly diffuse `gamma_bar`. However, in this
seeded run it badly degrades hard clustering accuracy. That means the simple
normalized scale update is not yet a satisfactory shrinkage solution. It is
useful as a diagnostic because it confirms that slot usage can be separated from
motif identity, but it also shows that stronger shrinkage needs to be integrated
more carefully with the responsibility update and the Gamma prior.

The next model-development step suggested by this result is to add an explicit
slot-usage/shrinkage mechanism. Possible options are submanifold-slot weights,
ARD-style priors on `E[beta_sd]`, or an initialization/update that keeps surplus
slots diffuse and small unless the likelihood strongly activates them.
