---
title: "MoS Poisson SuSiE Scenario Tests"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This file tests the first hard-motif MoS-Poisson-SuSiE implementation on two
simulation settings we have been using:

1. the block support-pattern scenario with one-factor and three-factor groups;
2. the harder tree scenario where no group occupies a single factor.

For each scenario, the oracle-motif fit asks whether the MoS mixture layer can
separate observations when the true factor matrix and submanifold motifs are
available. The initialized fit runs MF-Poisson-SuSiE first, initializes motifs
from posterior loading scores, and then fits the MoS layer.

The motif initializer allows an effective submanifold dimension smaller than
`D`: after clustering posterior loading scores, factors whose normalized center
weight is below the threshold are omitted, and the selected factors are padded
with repeats. This matters for singleton and two-factor groups.

```{r source-code}
source("code/mos-poi-susie.R")
```

## Helpers

```{r helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)
  perms = all_permutations(seq_len(K))
  scores = apply(perms, 1, function(p) {
    sum(similarity[cbind(seq_len(K), p)])
  })
  learned_to_true = perms[which.max(scores), ]

  list(
    table = data.frame(
      learned_factor = seq_len(K),
      matched_true_factor = learned_to_true,
      cosine = similarity[cbind(seq_len(K), learned_to_true)]
    ),
    learned_to_true = learned_to_true,
    similarity = similarity
  )
}

pad_supports = function(supports, D) {
  do.call(rbind, lapply(supports, function(support) {
    rep(support, length.out = D)
  }))
}

support_label = function(support) paste(sort(unique(support)), collapse = ",")

simulate_support_scenario = function(N, M, K, group_supports, shape_vec,
                                      factor_shape = 0.1,
                                      factor_rate = 0.01, seed = 1) {
  set.seed(seed)
  S = length(group_supports)
  n_per_group = N / S
  grp = rep(seq_len(S), each = n_per_group)

  F0 = matrix(rgamma(M * K, shape = factor_shape, rate = factor_rate),
              nrow = K, ncol = M)
  F0 = normalize_rows(F0)

  L = matrix(0, nrow = N, ncol = K)
  for (s in seq_len(S)) {
    idx = which(grp == s)
    support = group_supports[[s]]
    L[idx, support] = matrix(
      rgamma(length(idx) * length(support), shape = shape_vec[s], rate = 1),
      nrow = length(idx),
      ncol = length(support)
    )
  }

  Y = matrix(rpois(N * M, L %*% F0), nrow = N, ncol = M)

  list(
    Y = Y,
    F0 = F0,
    L = L,
    grp = grp,
    group_supports = group_supports,
    true_patterns = vapply(group_supports, support_label, character(1))
  )
}

plot_responsibility_heatmap = function(omega, grp, main) {
  N = nrow(omega)
  S = ncol(omega)
  n_groups = length(unique(grp))
  n_per_group = N / n_groups

  image(x = seq_len(N), y = seq_len(S), z = omega,
        axes = FALSE, xlab = "Sample", ylab = "Submanifold",
        main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", seq_len(n_groups)))
  axis(2, at = seq_len(S), labels = seq_len(S), las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  box()
}

plot_group_responsibilities = function(omega, grp, main) {
  group_resp = do.call(rbind, lapply(sort(unique(grp)), function(g) {
    colMeans(omega[grp == g, , drop = FALSE])
  }))

  image(x = seq_len(ncol(group_resp)), y = seq_len(nrow(group_resp)),
        z = t(group_resp), axes = FALSE, xlab = "Fitted submanifold",
        ylab = "True group", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(group_resp)),
       labels = paste0("S", seq_len(ncol(group_resp))))
  axis(2, at = seq_len(nrow(group_resp)),
       labels = paste0("G", seq_len(nrow(group_resp))), las = 1)
  for (g in seq_len(nrow(group_resp))) {
    for (s in seq_len(ncol(group_resp))) {
      text(s, g, sprintf("%.2f", group_resp[g, s]), cex = 0.65)
    }
  }
  box()

  invisible(group_resp)
}

plot_cosine_heatmap = function(similarity, main) {
  image(x = seq_len(ncol(similarity)), y = seq_len(nrow(similarity)),
        z = t(similarity), axes = FALSE, xlab = "True factor",
        ylab = "Fitted factor", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(similarity)),
       labels = paste0("true ", seq_len(ncol(similarity))))
  axis(2, at = seq_len(nrow(similarity)),
       labels = paste0("fit ", seq_len(nrow(similarity))), las = 1)
  for (i in seq_len(nrow(similarity))) {
    for (j in seq_len(ncol(similarity))) {
      text(j, i, sprintf("%.2f", similarity[i, j]), cex = 0.65)
    }
  }
  box()
}

translate_motifs_to_true_space = function(motifs, learned_to_true) {
  matrix(learned_to_true[motifs], nrow = nrow(motifs), ncol = ncol(motifs))
}

summarize_fit = function(fit, z_true) {
  data.frame(
    converged = fit$converged,
    n_iter = fit$n_iter,
    final_elbo = tail(fit$elbo[!is.na(fit$elbo)], 1),
    cluster_accuracy = cluster_accuracy(fit$z_hat, z_true),
    mean_max_responsibility = mean(apply(fit$omega, 1, max))
  )
}
```

## Scenario 1: Block Support Patterns

```{r block-simulate}
block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

block = simulate_support_scenario(
  N = 480,
  M = 500,
  K = 5,
  group_supports = block_supports,
  shape_vec = c(rep(10, 5), rep(20, 3)),
  seed = 42
)

data.frame(
  scenario = "block",
  N = nrow(block$Y),
  M = ncol(block$Y),
  K = nrow(block$F0),
  S = length(block_supports),
  total_count = sum(block$Y),
  nonzero_entries = sum(block$Y > 0)
)
```

```{r block-patterns}
data.frame(
  group = seq_along(block_supports),
  true_active_factors = block$true_patterns
)
```

### Oracle Motifs

```{r block-oracle-fit}
block_D = 3
block_motifs = pad_supports(block_supports, block_D)

block_oracle = mos_poi_susie_fixed_motifs(
  Y = block$Y,
  F = block$F0,
  motifs = block_motifs,
  max_iters = 30,
  n_inner = 6,
  update_prior = TRUE,
  tol = 1e-5,
  min_iters = 5,
  patience = 2
)

summarize_fit(block_oracle, block$grp)
```

```{r block-oracle-plots}
plot_group_responsibilities(
  block_oracle$omega,
  block$grp,
  "Block scenario: oracle motif responsibilities"
)

plot_responsibility_heatmap(
  block_oracle$omega,
  block$grp,
  "Block scenario: oracle responsibilities by sample"
)
```

### MF-Initialized Motifs

```{r block-learned-fit}
block_learned = mos_poi_susie(
  Y = block$Y,
  K = 5,
  S = length(block_supports),
  D = block_D,
  max_iters = 20,
  n_inner = 5,
  mf_max_iters = 50,
  mf_nmf_iters = 80,
  init_seed = 1,
  update_prior = TRUE,
  update_F = FALSE,
  update_motifs = FALSE,
  tol = 1e-5,
  min_iters = 5,
  patience = 2
)

summarize_fit(block_learned, block$grp)
```

```{r block-learned-factor-match}
block_match = match_factor_rows(block_learned$F, block$F0)
block_match$table

data.frame(
  mean_cosine = mean(block_match$table$cosine),
  min_cosine = min(block_match$table$cosine)
)
```

```{r block-learned-motifs}
block_motifs_true_space =
  translate_motifs_to_true_space(block_learned$motifs,
                                 block_match$learned_to_true)

data.frame(
  fitted_submanifold = seq_len(nrow(block_learned$motifs)),
  motif_in_fitted_factor_space = apply(block_learned$motifs, 1, support_label),
  motif_mapped_to_true_factor_space =
    apply(block_motifs_true_space, 1, support_label)
)
```

```{r block-learned-plots}
plot_cosine_heatmap(
  block_match$similarity,
  "Block scenario: fitted and true factor cosine"
)

plot_group_responsibilities(
  block_learned$omega,
  block$grp,
  "Block scenario: MF-initialized MoS responsibilities"
)
```

## Scenario 2: Tree Support Patterns

```{r tree-simulate}
tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

tree = simulate_support_scenario(
  N = 480,
  M = 600,
  K = 7,
  group_supports = tree_supports,
  shape_vec = rep(18, length(tree_supports)),
  seed = 47
)

data.frame(
  scenario = "tree",
  N = nrow(tree$Y),
  M = ncol(tree$Y),
  K = nrow(tree$F0),
  S = length(tree_supports),
  total_count = sum(tree$Y),
  nonzero_entries = sum(tree$Y > 0)
)
```

```{r tree-patterns}
data.frame(
  group = seq_along(tree_supports),
  true_active_factors = tree$true_patterns
)
```

### Oracle Motifs

```{r tree-oracle-fit}
tree_D = 3
tree_motifs = pad_supports(tree_supports, tree_D)

tree_oracle = mos_poi_susie_fixed_motifs(
  Y = tree$Y,
  F = tree$F0,
  motifs = tree_motifs,
  max_iters = 30,
  n_inner = 6,
  update_prior = TRUE,
  tol = 1e-5,
  min_iters = 5,
  patience = 2
)

summarize_fit(tree_oracle, tree$grp)
```

```{r tree-oracle-plots}
plot_group_responsibilities(
  tree_oracle$omega,
  tree$grp,
  "Tree scenario: oracle motif responsibilities"
)

plot_responsibility_heatmap(
  tree_oracle$omega,
  tree$grp,
  "Tree scenario: oracle responsibilities by sample"
)
```

### MF-Initialized Motifs

```{r tree-learned-fit}
tree_learned = mos_poi_susie(
  Y = tree$Y,
  K = 7,
  S = length(tree_supports),
  D = tree_D,
  max_iters = 20,
  n_inner = 5,
  mf_max_iters = 60,
  mf_nmf_iters = 100,
  init_seed = 2,
  update_prior = TRUE,
  update_F = FALSE,
  update_motifs = FALSE,
  tol = 1e-5,
  min_iters = 5,
  patience = 2
)

summarize_fit(tree_learned, tree$grp)
```

```{r tree-learned-factor-match}
tree_match = match_factor_rows(tree_learned$F, tree$F0)
tree_match$table

data.frame(
  mean_cosine = mean(tree_match$table$cosine),
  min_cosine = min(tree_match$table$cosine)
)
```

```{r tree-learned-motifs}
tree_motifs_true_space =
  translate_motifs_to_true_space(tree_learned$motifs,
                                 tree_match$learned_to_true)

data.frame(
  fitted_submanifold = seq_len(nrow(tree_learned$motifs)),
  motif_in_fitted_factor_space = apply(tree_learned$motifs, 1, support_label),
  motif_mapped_to_true_factor_space =
    apply(tree_motifs_true_space, 1, support_label)
)
```

```{r tree-learned-plots}
plot_cosine_heatmap(
  tree_match$similarity,
  "Tree scenario: fitted and true factor cosine"
)

plot_group_responsibilities(
  tree_learned$omega,
  tree$grp,
  "Tree scenario: MF-initialized MoS responsibilities"
)
```

## Takeaways To Check

The oracle-motif fits are sanity checks for the MoS layer: if they fail, the
mixture/responsibility updates are wrong. The MF-initialized fits are more
demanding. In the block scenario, the single-factor groups provide anchors, so
the factor rows and motifs should be easier to initialize. In the tree scenario,
the no-anchor structure can still inherit the factor-identifiability issue we
saw in `tree-scenario.Rmd`; good clustering there would suggest the
submanifold mixture can partially compensate for imperfect factor recovery.

The remaining limitation is that motifs are frozen after initialization. The
next version should replace hard initialized motifs with soft submanifold-level
posterior motif probabilities and update them jointly with `F`.
