---
title: "MoS-NMF Multi-Seed Benchmark"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This notebook turns the single-seed benchmark into a compact multi-seed pilot.
It is still intentionally small, but it gives a more honest read on robustness
than a single lucky or unlucky simulation.

```{r source-code}
source("code/mos-nmf-benchmark-utils.R")
```

## Benchmark Design

```{r design}
benchmark_seeds = 1:3

block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

data.frame(
  scenario = c("block", "tree"),
  K = c(5, 7),
  S = c(length(block_supports), length(tree_supports)),
  D = c(3, 3),
  N = c(240, 240),
  M = c(250, 280),
  n_seed = length(benchmark_seeds)
)
```

## Run Fits

```{r run-benchmark}
run_block_seed = function(seed) {
  dat = simulate_support_scenario(
    N = 240,
    M = 250,
    K = 5,
    group_supports = block_supports,
    shape_vec = c(rep(10, 5), rep(20, 3)),
    seed = 100 + seed
  )

  out = run_one_benchmark(
    dat = dat,
    K = 5,
    S = length(block_supports),
    D = 3,
    seed = seed,
    nmf_iters = 50,
    mf_max_iters = 25,
    mf_nmf_iters = 45,
    mos_max_iters = 12,
    mos_n_inner = 3,
    block_size = 100
  )
  out$scenario = "block"
  out$seed = seed
  out
}

run_tree_seed = function(seed) {
  dat = simulate_support_scenario(
    N = 240,
    M = 280,
    K = 7,
    group_supports = tree_supports,
    shape_vec = rep(18, length(tree_supports)),
    seed = 200 + seed
  )

  out = run_one_benchmark(
    dat = dat,
    K = 7,
    S = length(tree_supports),
    D = 3,
    seed = seed,
    nmf_iters = 50,
    mf_max_iters = 25,
    mf_nmf_iters = 45,
    mos_max_iters = 12,
    mos_n_inner = 3,
    block_size = 100
  )
  out$scenario = "tree"
  out$seed = seed
  out
}

block_results = do.call(rbind, lapply(benchmark_seeds, run_block_seed))
tree_results = do.call(rbind, lapply(benchmark_seeds, run_tree_seed))
benchmark_results = rbind(block_results, tree_results)

benchmark_results
```

## Summary Tables

```{r summaries}
summary_results = summarize_benchmark_results(benchmark_results)
summary_results
```

```{r compact-summary}
compact_summary = summary_results[, c(
  "scenario",
  "method",
  "n_seed",
  "cluster_accuracy_mean",
  "cluster_accuracy_sd",
  "support_accuracy_mean",
  "support_accuracy_sd",
  "mean_factor_cosine_mean",
  "mean_factor_cosine_sd",
  "mean_gamma_entropy_mean"
)]

compact_summary
```

## Visual Summary

```{r plotting-helpers}
plot_metric_by_seed = function(results, metric, main) {
  methods = unique(results$method)
  scenarios = unique(results$scenario)
  op = par(mfrow = c(1, length(scenarios)), mar = c(7, 4, 3, 1))
  on.exit(par(op))

  for (scenario_name in scenarios) {
    tab = results[results$scenario == scenario_name, ]
    vals = split(tab[[metric]], tab$method)
    boxplot(vals, las = 2, ylim = c(0, 1), ylab = metric,
            main = paste(main, scenario_name), col = "gray80")
    stripchart(vals, vertical = TRUE, method = "jitter", pch = 19,
               add = TRUE, col = "firebrick")
  }
}
```

```{r plots}
plot_metric_by_seed(benchmark_results, "cluster_accuracy",
                    "Cluster accuracy:")
plot_metric_by_seed(benchmark_results, "support_accuracy",
                    "Support accuracy:")
plot_metric_by_seed(benchmark_results, "mean_factor_cosine",
                    "Mean factor cosine:")
```

## Interpretation

This is a pilot benchmark, not the final paper table. Its purpose is to reveal
whether the single-seed conclusions survive basic random-seed variation.

In the block setting, soft MoS-NMF has the best mean hard clustering accuracy
and factor recovery in this small pilot. Its support accuracy is not clearly
better than the baselines here: the three-seed mean is slightly below
MF-Poisson-SuSiE + k-means and close to Poisson NMF + k-means. This is an
important caveat for the paper framing. The block result supports feasibility,
but not yet a broad superiority claim for support recovery.

In the tree/no-anchor setting, the signal is more encouraging for the MoS
model. Soft MoS-NMF gives support recovery around `0.76` on average, while the
two baseline pipelines are around `0.41`. Hard clustering is also comparable in
this smaller multi-seed pilot. This is exactly the kind of setting where motif
sharing should help: no group is attached to a single anchor factor, so the
method has to learn repeated combinations of factors rather than isolated
single-factor groups.

The `gamma_bar` entropy is essentially zero in both clean settings. That should
not be interpreted as a failure to quantify uncertainty. With correct `K`, a
moderate number of samples per motif, and reasonably separated factors, the
motif factor identities are strongly identified, so the posterior should
concentrate. The uncertainty claim should instead be evaluated in deliberately
ambiguous settings such as over-specified `K`, duplicated factors, correlated
factors, or weak motifs.

The next version of this benchmark should expand the number of seeds and add
held-out Poisson deviance or predictive log likelihood. The current pilot is
useful evidence for the roadmap, but it is still too small to be a final
simulation table.
