---
title: "Soft-Gamma MoS-NMF Alternating Updates"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This file tests the soft-gamma MoS-NMF model. The fitted motif is no longer a
hard matrix of factor labels. Instead, each submanifold slot has
`gamma_bar[s, d, k]`, the posterior probability that motif slot `d` in
submanifold `s` uses factor `k`.

The local Jensen allocation `xi[i, s, d, m]` is still observation-specific, but
the implementation streams it over feature blocks. It only stores sufficient
statistics, so the full `N x S x D x M` tensor is never materialized.

```{r source-code}
source("code/mos-nmf.R")
```

## Helpers

```{r helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)
  perms = all_permutations(seq_len(K))
  scores = apply(perms, 1, function(p) {
    sum(similarity[cbind(seq_len(K), p)])
  })
  learned_to_true = perms[which.max(scores), ]

  list(
    table = data.frame(
      learned_factor = seq_len(K),
      matched_true_factor = learned_to_true,
      cosine = similarity[cbind(seq_len(K), learned_to_true)]
    ),
    learned_to_true = learned_to_true,
    similarity = similarity
  )
}

support_label = function(support) paste(sort(unique(support)), collapse = ",")

simulate_support_scenario = function(N, M, K, group_supports, shape_vec,
                                      factor_shape = 0.1,
                                      factor_rate = 0.01, seed = 1) {
  set.seed(seed)
  S = length(group_supports)
  n_per_group = N / S
  grp = rep(seq_len(S), each = n_per_group)

  F0 = matrix(rgamma(M * K, shape = factor_shape, rate = factor_rate),
              nrow = K, ncol = M)
  F0 = normalize_rows(F0)

  L = matrix(0, nrow = N, ncol = K)
  for (s in seq_len(S)) {
    idx = which(grp == s)
    support = group_supports[[s]]
    L[idx, support] = matrix(
      rgamma(length(idx) * length(support), shape = shape_vec[s], rate = 1),
      nrow = length(idx),
      ncol = length(support)
    )
  }

  Y = matrix(rpois(N * M, L %*% F0), nrow = N, ncol = M)

  list(
    Y = Y,
    F0 = F0,
    L = L,
    grp = grp,
    group_supports = group_supports,
    true_patterns = vapply(group_supports, support_label, character(1))
  )
}

map_scores_to_true = function(scores, learned_to_true) {
  K = length(learned_to_true)
  out = matrix(0, nrow = nrow(scores), ncol = K)
  for (k in seq_len(K)) {
    out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
  }
  out
}

fit_soft_pair = function(dat, K, S, D, init_seed,
                         mf_max_iters = 50,
                         mf_nmf_iters = 80,
                         mos_max_iters = 25,
                         n_inner = 4,
                         block_size = 100) {
  mf_fit = mf_poi_susie(
    Y = dat$Y,
    K = K,
    D = D,
    max_iters = mf_max_iters,
    update_prior = TRUE,
    init_seed = init_seed,
    init_F = "poisson_nmf",
    nmf_iters = mf_nmf_iters,
    init_gamma_from_nmf = TRUE,
    elbo_every = 5,
    tol = 5e-4,
    min_iters = 15,
    patience = 2
  )

  gamma_init = init_soft_gamma_from_loading_scores(
    loading_scores_from_mf(mf_fit),
    S = S,
    D = D,
    init_seed = init_seed,
    min_share = 0.10,
    gamma_floor = 0.05
  )

  fixed_F = mos_nmf_fixed_gamma(
    Y = dat$Y,
    F = mf_fit$F,
    gamma_bar = gamma_init$gamma_bar,
    max_iters = mos_max_iters,
    n_inner = n_inner,
    update_prior = TRUE,
    update_F = FALSE,
    update_gamma = TRUE,
    gamma_step_init = 0.5,
    gamma_step_ramp = 8,
    tol = 1e-5,
    min_iters = 5,
    patience = 2,
    block_size = block_size
  )

  alternating = mos_nmf_fixed_gamma(
    Y = dat$Y,
    F = mf_fit$F,
    gamma_bar = gamma_init$gamma_bar,
    max_iters = mos_max_iters,
    n_inner = n_inner,
    update_prior = TRUE,
    update_F = TRUE,
    update_gamma = TRUE,
    gamma_step_init = 0.5,
    gamma_step_ramp = 8,
    F_step_init = 0.2,
    F_step_ramp = 15,
    tol = 1e-5,
    min_iters = 5,
    patience = 2,
    block_size = block_size
  )

  list(mf_fit = mf_fit, gamma_init = gamma_init,
       fixed_F = fixed_F, alternating = alternating)
}

summarize_one_fit = function(name, fit, z_true, F0) {
  factor_match = match_factor_rows(fit$F, F0)
  ent = motif_gamma_entropy(fit$gamma_bar)
  data.frame(
    fit = name,
    converged = fit$converged,
    n_iter = fit$n_iter,
    final_elbo = tail(fit$elbo[!is.na(fit$elbo)], 1),
    cluster_accuracy = cluster_accuracy(fit$z_hat, z_true),
    mean_max_responsibility = mean(apply(fit$omega, 1, max)),
    mean_factor_cosine = mean(factor_match$table$cosine),
    min_factor_cosine = min(factor_match$table$cosine),
    mean_motif_entropy = mean(ent),
    max_motif_entropy = max(ent)
  )
}

summarize_pair = function(pair, dat) {
  rbind(
    summarize_one_fit("soft fixed F", pair$fixed_F, dat$grp, dat$F0),
    summarize_one_fit("soft alternating F", pair$alternating, dat$grp, dat$F0)
  )
}

motif_score_table = function(fit, F0, label, top_n = 3) {
  factor_match = match_factor_rows(fit$F, F0)
  scores = motif_factor_scores(fit)$scores
  true_scores = map_scores_to_true(scores, factor_match$learned_to_true)

  do.call(rbind, lapply(seq_len(nrow(true_scores)), function(s) {
    ord = order(true_scores[s, ], decreasing = TRUE)[seq_len(top_n)]
    data.frame(
      fit = label,
      fitted_submanifold = s,
      top_true_factors = paste(ord, collapse = ","),
      top_scores = paste(sprintf("%.2f", true_scores[s, ord]),
                         collapse = ",")
    )
  }))
}

plot_group_responsibilities = function(omega, grp, main) {
  group_resp = do.call(rbind, lapply(sort(unique(grp)), function(g) {
    colMeans(omega[grp == g, , drop = FALSE])
  }))

  image(x = seq_len(ncol(group_resp)), y = seq_len(nrow(group_resp)),
        z = t(group_resp), axes = FALSE, xlab = "Fitted submanifold",
        ylab = "True group", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(group_resp)),
       labels = paste0("S", seq_len(ncol(group_resp))))
  axis(2, at = seq_len(nrow(group_resp)),
       labels = paste0("G", seq_len(nrow(group_resp))), las = 1)
  for (g in seq_len(nrow(group_resp))) {
    for (s in seq_len(ncol(group_resp))) {
      text(s, g, sprintf("%.2f", group_resp[g, s]), cex = 0.65)
    }
  }
  box()
}

plot_cosine_heatmap = function(similarity, main) {
  image(x = seq_len(ncol(similarity)), y = seq_len(nrow(similarity)),
        z = t(similarity), axes = FALSE, xlab = "True factor",
        ylab = "Fitted factor", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(similarity)),
       labels = paste0("true ", seq_len(ncol(similarity))))
  axis(2, at = seq_len(nrow(similarity)),
       labels = paste0("fit ", seq_len(nrow(similarity))), las = 1)
  for (i in seq_len(nrow(similarity))) {
    for (j in seq_len(ncol(similarity))) {
      text(j, i, sprintf("%.2f", similarity[i, j]), cex = 0.65)
    }
  }
  box()
}

plot_motif_scores = function(fit, F0, main) {
  factor_match = match_factor_rows(fit$F, F0)
  scores = motif_factor_scores(fit)$scores
  true_scores = map_scores_to_true(scores, factor_match$learned_to_true)
  true_scores = true_scores / pmax(rowSums(true_scores), 1e-12)

  image(x = seq_len(ncol(true_scores)), y = seq_len(nrow(true_scores)),
        z = t(true_scores), axes = FALSE, xlab = "True factor",
        ylab = "Fitted submanifold", main = main,
        col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(true_scores)),
       labels = paste0("true ", seq_len(ncol(true_scores))))
  axis(2, at = seq_len(nrow(true_scores)),
       labels = paste0("S", seq_len(nrow(true_scores))), las = 1)
  for (s in seq_len(nrow(true_scores))) {
    for (k in seq_len(ncol(true_scores))) {
      text(k, s, sprintf("%.2f", true_scores[s, k]), cex = 0.65)
    }
  }
  box()
}

plot_motif_entropy = function(fit, main) {
  ent = motif_gamma_entropy(fit$gamma_bar)
  image(x = seq_len(ncol(ent)), y = seq_len(nrow(ent)),
        z = t(ent), axes = FALSE, xlab = "Slot d",
        ylab = "Fitted submanifold", main = main,
        col = hcl.colors(100, "Blues"))
  axis(1, at = seq_len(ncol(ent)), labels = paste0("d", seq_len(ncol(ent))))
  axis(2, at = seq_len(nrow(ent)), labels = paste0("S", seq_len(nrow(ent))),
       las = 1)
  for (s in seq_len(nrow(ent))) {
    for (d in seq_len(ncol(ent))) {
      text(d, s, sprintf("%.2f", ent[s, d]), cex = 0.65)
    }
  }
  box()
}
```

## Block Scenario

```{r block-simulate}
block_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

block = simulate_support_scenario(
  N = 320,
  M = 350,
  K = 5,
  group_supports = block_supports,
  shape_vec = c(rep(10, 5), rep(20, 3)),
  seed = 42
)

data.frame(group = seq_along(block_supports),
           true_active_factors = block$true_patterns)
```

```{r block-fit}
block_pair = fit_soft_pair(
  dat = block,
  K = 5,
  S = length(block_supports),
  D = 3,
  init_seed = 1,
  mf_max_iters = 35,
  mf_nmf_iters = 60,
  mos_max_iters = 18,
  n_inner = 3,
  block_size = 100
)

summarize_pair(block_pair, block)
```

```{r block-motif-table}
rbind(
  motif_score_table(block_pair$fixed_F, block$F0, "soft fixed F"),
  motif_score_table(block_pair$alternating, block$F0, "soft alternating F")
)
```

```{r block-plots}
plot_group_responsibilities(
  block_pair$fixed_F$omega,
  block$grp,
  "Block scenario: soft fixed-F responsibilities"
)

plot_group_responsibilities(
  block_pair$alternating$omega,
  block$grp,
  "Block scenario: soft alternating responsibilities"
)

plot_cosine_heatmap(
  match_factor_rows(block_pair$alternating$F, block$F0)$similarity,
  "Block scenario: alternating fitted and true factor cosine"
)

plot_motif_scores(
  block_pair$alternating,
  block$F0,
  "Block scenario: posterior motif mass in true factor space"
)

plot_motif_entropy(
  block_pair$alternating,
  "Block scenario: normalized entropy of gamma_bar[s,d,]"
)
```

## Tree Scenario

```{r tree-simulate}
tree_supports = list(
  c(1, 2, 3),
  c(1, 2, 4),
  c(1, 5, 6),
  c(1, 5, 7),
  c(1, 2),
  c(1, 5)
)

tree = simulate_support_scenario(
  N = 360,
  M = 420,
  K = 7,
  group_supports = tree_supports,
  shape_vec = rep(18, length(tree_supports)),
  seed = 47
)

data.frame(group = seq_along(tree_supports),
           true_active_factors = tree$true_patterns)
```

```{r tree-fit}
tree_pair = fit_soft_pair(
  dat = tree,
  K = 7,
  S = length(tree_supports),
  D = 3,
  init_seed = 2,
  mf_max_iters = 40,
  mf_nmf_iters = 70,
  mos_max_iters = 18,
  n_inner = 3,
  block_size = 120
)

summarize_pair(tree_pair, tree)
```

```{r tree-motif-table}
rbind(
  motif_score_table(tree_pair$fixed_F, tree$F0, "soft fixed F"),
  motif_score_table(tree_pair$alternating, tree$F0, "soft alternating F")
)
```

```{r tree-plots}
plot_group_responsibilities(
  tree_pair$fixed_F$omega,
  tree$grp,
  "Tree scenario: soft fixed-F responsibilities"
)

plot_group_responsibilities(
  tree_pair$alternating$omega,
  tree$grp,
  "Tree scenario: soft alternating responsibilities"
)

plot_cosine_heatmap(
  match_factor_rows(tree_pair$alternating$F, tree$F0)$similarity,
  "Tree scenario: alternating fitted and true factor cosine"
)

plot_motif_scores(
  tree_pair$alternating,
  tree$F0,
  "Tree scenario: posterior motif mass in true factor space"
)

plot_motif_entropy(
  tree_pair$alternating,
  "Tree scenario: normalized entropy of gamma_bar[s,d,]"
)
```

## Takeaway

The main object to inspect is `gamma_bar`. Concentrated rows indicate that a
submanifold slot has selected a factor confidently. Diffuse rows indicate
posterior uncertainty over competing fitted factors. This is different from
`omega`, which measures uncertainty over the observation's submanifold label.

The first comparison keeps `F` fixed after MF-Poisson-SuSiE initialization and
only learns the soft motifs. The second comparison alternates the soft motif
posterior with a responsibility-weighted update of `F`.
