---
title: "Soft MoS-NMF Uncertainty Stress Tests"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This file targets the paper's uncertainty claim. In clean simulations,
`gamma_bar[s,d,]` often collapses to a point mass because each motif is strongly
identified by many samples. Here we create settings where uncertainty should be
visible: overfitted factor rows and correlated true factors.

```{r source-code}
source("code/mos-nmf.R")
```

## Helpers

```{r helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

support_label = function(support) paste(sort(unique(support)), collapse = ",")

simulate_support_scenario = function(N, M, F0, group_supports, shape_vec,
                                      seed = 1) {
  set.seed(seed)
  K = nrow(F0)
  S = length(group_supports)
  n_per_group = N / S
  grp = rep(seq_len(S), each = n_per_group)

  L = matrix(0, nrow = N, ncol = K)
  for (s in seq_len(S)) {
    idx = which(grp == s)
    support = group_supports[[s]]
    L[idx, support] = matrix(
      rgamma(length(idx) * length(support), shape = shape_vec[s], rate = 1),
      nrow = length(idx),
      ncol = length(support)
    )
  }

  Y = matrix(rpois(N * M, L %*% F0), nrow = N, ncol = M)

  list(
    Y = Y,
    F0 = F0,
    L = L,
    grp = grp,
    group_supports = group_supports,
    true_patterns = vapply(group_supports, support_label, character(1))
  )
}

make_sparse_factors = function(K, M, shape = 0.1, rate = 0.01, seed = 1) {
  set.seed(seed)
  F0 = matrix(rgamma(K * M, shape = shape, rate = rate), nrow = K, ncol = M)
  normalize_rows(F0)
}

make_correlated_factors = function(K, M, correlated_pair = c(1, 2),
                                   mix = 0.92, seed = 1) {
  F0 = make_sparse_factors(K, M, seed = seed)
  a = correlated_pair[1]
  b = correlated_pair[2]
  F0[b, ] = normalize_rows(matrix(mix * F0[a, ] + (1 - mix) * F0[b, ],
                                  nrow = 1))
  F0
}

fit_mos_from_true_F = function(dat, F_init, K_fit, S, D, init_seed,
                               max_iters = 20, n_inner = 3,
                               block_size = 100,
                               gamma_floor = 0.20,
                               update_F = FALSE) {
  set.seed(init_seed)
  nmf_init = poisson_nmf_init(dat$Y, K_fit, max_iters = 60,
                              init_seed = init_seed)
  scores = nmf_init$L
  gamma_init = init_soft_gamma_from_loading_scores(
    scores = scores,
    S = S,
    D = D,
    init_seed = init_seed,
    gamma_floor = gamma_floor
  )

  fit = mos_nmf_fixed_gamma(
    Y = dat$Y,
    F = F_init,
    gamma_bar = gamma_init$gamma_bar,
    max_iters = max_iters,
    n_inner = n_inner,
    update_prior = TRUE,
    update_F = update_F,
    update_gamma = TRUE,
    gamma_step_init = 0.25,
    gamma_step_ramp = 10,
    F_step_init = 0.1,
    F_step_ramp = 15,
    tol = 1e-5,
    min_iters = 6,
    patience = 2,
    block_size = block_size
  )

  fit$gamma_init = gamma_init
  fit
}

top_two_gap = function(prob) {
  sorted = sort(prob, decreasing = TRUE)
  sorted[1] - sorted[2]
}

gamma_uncertainty_table = function(fit, label = "fit") {
  S = dim(fit$gamma_bar)[1]
  D = dim(fit$gamma_bar)[2]
  out = do.call(rbind, lapply(seq_len(S), function(s) {
    do.call(rbind, lapply(seq_len(D), function(d) {
      p = fit$gamma_bar[s, d, ]
      data.frame(
        fit = label,
        submanifold = s,
        slot = d,
        top_factor = which.max(p),
        top_prob = max(p),
        top_two_gap = top_two_gap(p),
        entropy = -sum(p * log(p)) / log(length(p))
      )
    }))
  }))
  rownames(out) = NULL
  out
}

plot_gamma_slot_heatmap = function(fit, slots, main) {
  mat = do.call(rbind, lapply(seq_len(nrow(slots)), function(j) {
    fit$gamma_bar[slots$submanifold[j], slots$slot[j], ]
  }))
  row_lab = paste0("S", slots$submanifold, ":d", slots$slot)

  image(x = seq_len(ncol(mat)), y = seq_len(nrow(mat)),
        z = t(mat), axes = FALSE, xlab = "Fitted factor",
        ylab = "Motif slot", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(mat)), labels = paste0("fit ", seq_len(ncol(mat))))
  axis(2, at = seq_len(nrow(mat)), labels = row_lab, las = 1)
  for (i in seq_len(nrow(mat))) {
    for (k in seq_len(ncol(mat))) {
      text(k, i, sprintf("%.2f", mat[i, k]), cex = 0.65)
    }
  }
  box()
}

plot_group_responsibilities = function(omega, grp, main) {
  group_resp = do.call(rbind, lapply(sort(unique(grp)), function(g) {
    colMeans(omega[grp == g, , drop = FALSE])
  }))

  image(x = seq_len(ncol(group_resp)), y = seq_len(nrow(group_resp)),
        z = t(group_resp), axes = FALSE, xlab = "Fitted submanifold",
        ylab = "True group", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(group_resp)),
       labels = paste0("S", seq_len(ncol(group_resp))))
  axis(2, at = seq_len(nrow(group_resp)),
       labels = paste0("G", seq_len(nrow(group_resp))), las = 1)
  for (g in seq_len(nrow(group_resp))) {
    for (s in seq_len(ncol(group_resp))) {
      text(s, g, sprintf("%.2f", group_resp[g, s]), cex = 0.65)
    }
  }
  box()
}

plot_factor_cosine = function(F, F_ref, main) {
  cosine = row_cosine(F, F_ref)
  image(x = seq_len(ncol(cosine)), y = seq_len(nrow(cosine)),
        z = t(cosine), axes = FALSE, xlab = "Reference factor",
        ylab = "Fitted factor", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(cosine)),
       labels = paste0("ref ", seq_len(ncol(cosine))))
  axis(2, at = seq_len(nrow(cosine)),
       labels = paste0("fit ", seq_len(nrow(cosine))), las = 1)
  for (i in seq_len(nrow(cosine))) {
    for (j in seq_len(ncol(cosine))) {
      text(j, i, sprintf("%.2f", cosine[i, j]), cex = 0.65)
    }
  }
  box()
}

motif_scores_true_space = function(fit, learned_to_true) {
  scores = motif_factor_scores(fit)$scores
  K_true = max(learned_to_true, na.rm = TRUE)
  out = matrix(0, nrow = nrow(scores), ncol = K_true)
  for (k in seq_along(learned_to_true)) {
    if (!is.na(learned_to_true[k])) {
      out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
    }
  }
  out
}

summarize_fit = function(fit, dat, label) {
  ent = gamma_uncertainty_table(fit, label)
  data.frame(
    fit = label,
    cluster_accuracy = cluster_accuracy(fit$z_hat, dat$grp),
    mean_max_responsibility = mean(apply(fit$omega, 1, max)),
    mean_gamma_entropy = mean(ent$entropy),
    max_gamma_entropy = max(ent$entropy),
    mean_top_prob = mean(ent$top_prob),
    min_top_prob = min(ent$top_prob),
    final_elbo = tail(fit$elbo[!is.na(fit$elbo)], 1)
  )
}
```

## Overspecified K With Duplicate Candidate Factors

Here the data are generated from `K_true = 5` factors, but the fitted
dictionary has `K_fit = 7`. The two extra rows are near-duplicates of true
factors 1 and 2. We hold this overcomplete dictionary fixed in the first test.
If the soft motif posterior is doing its job, slots involving true factors 1 or
2 should be allowed to spread mass across the original and duplicate rows.

```{r overspecified-k-simulate}
K_true = 5
K_fit = 7
M = 300

F_true = make_sparse_factors(K_true, M, seed = 22)
set.seed(23)
dup1 = normalize_rows(matrix(0.94 * F_true[1, ] +
                               0.06 * make_sparse_factors(1, M, seed = 24),
                             nrow = 1))
dup2 = normalize_rows(matrix(0.94 * F_true[2, ] +
                               0.06 * make_sparse_factors(1, M, seed = 25),
                             nrow = 1))
F_over = rbind(F_true, dup1, dup2)

over_supports = list(
  c(1),
  c(2),
  c(3),
  c(4),
  c(5),
  c(1, 2, 3),
  c(1, 2, 4),
  c(2, 3, 5)
)

over_dat = simulate_support_scenario(
  N = 240,
  M = M,
  F0 = F_true,
  group_supports = over_supports,
  shape_vec = c(rep(8, 5), rep(14, 3)),
  seed = 26
)

data.frame(
  group = seq_along(over_supports),
  true_active_factors = over_dat$true_patterns
)
```

```{r overspecified-k-fit}
over_fit = fit_mos_from_true_F(
  dat = over_dat,
  F_init = F_over,
  K_fit = K_fit,
  S = length(over_supports),
  D = 3,
  init_seed = 27,
  max_iters = 18,
  n_inner = 3,
  block_size = 100,
  gamma_floor = 0.30,
  update_F = FALSE
)

summarize_fit(over_fit, over_dat, "fixed overcomplete F")
```

```{r overspecified-k-diagnostics}
plot_factor_cosine(
  F_over,
  F_true,
  "Overcomplete dictionary: candidate and true factor cosine"
)

plot_group_responsibilities(
  over_fit$omega,
  over_dat$grp,
  "Overspecified K: group responsibilities"
)

over_uncertainty = gamma_uncertainty_table(over_fit, "fixed overcomplete F")
over_uncertainty[order(over_uncertainty$top_prob), ][seq_len(12), ]
```

```{r overspecified-k-gamma-heatmap}
plot_gamma_slot_heatmap(
  over_fit,
  slots = over_uncertainty[order(over_uncertainty$top_prob), ][seq_len(12), ],
  "Overspecified K: most uncertain motif slots"
)
```

The extra rows are candidate factors 6 and 7. They are near-duplicates of true
factors 1 and 2 respectively. Posterior mass split between rows `(1,6)` or
`(2,7)` is the desirable behavior here: a hard motif would be forced to choose
one duplicate, while `gamma_bar` can represent the ambiguity.

```{r duplicate-mass}
duplicate_mass = do.call(rbind, lapply(seq_len(dim(over_fit$gamma_bar)[1]),
                                        function(s) {
  do.call(rbind, lapply(seq_len(dim(over_fit$gamma_bar)[2]), function(d) {
    data.frame(
      submanifold = s,
      slot = d,
      mass_true1_or_dup = over_fit$gamma_bar[s, d, 1] +
        over_fit$gamma_bar[s, d, 6],
      split_true1_vs_dup = min(over_fit$gamma_bar[s, d, 1],
                               over_fit$gamma_bar[s, d, 6]),
      mass_true2_or_dup = over_fit$gamma_bar[s, d, 2] +
        over_fit$gamma_bar[s, d, 7],
      split_true2_vs_dup = min(over_fit$gamma_bar[s, d, 2],
                               over_fit$gamma_bar[s, d, 7])
    )
  }))
}))

duplicate_mass[order(-pmax(duplicate_mass$split_true1_vs_dup,
                           duplicate_mass$split_true2_vs_dup)), ][seq_len(12), ]
```

In this seeded experiment, the most useful diagnostic is the split mass within
the duplicate pairs. A large value of `split_true2_vs_dup`, for example, means
that the motif slot has not arbitrarily selected either fitted row 2 or fitted
row 7; it has posterior support for both rows. This is the kind of behavior we
would want to emphasize in a paper, because it distinguishes posterior motif
uncertainty from hard support recovery.

## Correlated True Factors

Here `K` is correctly specified, but true factors 1 and 2 are highly correlated.
Groups that use either factor should be harder to distinguish, and uncertainty
should concentrate around those slots.

```{r correlated-simulate}
K = 5
M = 300
F_corr = make_correlated_factors(K, M, correlated_pair = c(1, 2),
                                 mix = 0.96, seed = 32)

corr_supports = list(
  c(1),
  c(2),
  c(3),
  c(1, 3),
  c(2, 3),
  c(1, 2)
)

corr_dat = simulate_support_scenario(
  N = 180,
  M = M,
  F0 = F_corr,
  group_supports = corr_supports,
  shape_vec = c(5, 5, 8, 10, 10, 10),
  seed = 33
)

data.frame(
  group = seq_along(corr_supports),
  true_active_factors = corr_dat$true_patterns
)
```

```{r correlated-fit}
corr_fit = fit_mos_from_true_F(
  dat = corr_dat,
  F_init = F_corr,
  K_fit = K,
  S = length(corr_supports),
  D = 2,
  init_seed = 34,
  max_iters = 18,
  n_inner = 3,
  block_size = 100,
  gamma_floor = 0.30,
  update_F = FALSE
)

summarize_fit(corr_fit, corr_dat, "fixed correlated F")
```

```{r correlated-diagnostics}
plot_factor_cosine(
  F_corr,
  F_corr,
  "Correlated scenario: true factor cosine"
)

plot_group_responsibilities(
  corr_fit$omega,
  corr_dat$grp,
  "Correlated factors: group responsibilities"
)

corr_uncertainty = gamma_uncertainty_table(corr_fit, "fixed correlated F")
corr_uncertainty[order(corr_uncertainty$top_prob), ]
```

```{r correlated-gamma-heatmap}
plot_gamma_slot_heatmap(
  corr_fit,
  slots = corr_uncertainty[order(corr_uncertainty$top_prob), ],
  "Correlated factors: motif slot posteriors"
)
```

```{r correlated-pair-mass}
corr_pair_mass = do.call(rbind, lapply(seq_len(dim(corr_fit$gamma_bar)[1]),
                                        function(s) {
  do.call(rbind, lapply(seq_len(dim(corr_fit$gamma_bar)[2]), function(d) {
    p = corr_fit$gamma_bar[s, d, ]
    data.frame(
      submanifold = s,
      slot = d,
      mass_factor_1_or_2 = p[1] + p[2],
      split_between_1_and_2 = min(p[1], p[2]),
      entropy = -sum(p * log(p)) / log(length(p))
    )
  }))
}))

corr_pair_mass[order(-corr_pair_mass$split_between_1_and_2), ]
```

This correlated-factor stress test is intentionally difficult. In this seeded
run, some motif slots remain nearly uniform, which produces high entropy but is
not the same as targeted uncertainty between factors 1 and 2. The more
convincing paper diagnostic is therefore not raw entropy alone, but entropy
together with concentrated mass on a scientifically meaningful competing set,
such as `mass_factor_1_or_2` and `split_between_1_and_2`.

## Takeaway

These are targeted stress tests for uncertainty, not final benchmark
experiments. The desired signal is not only high clustering accuracy. We also
want `gamma_bar` to behave differently from a hard motif assignment: when
multiple factor rows are statistically interchangeable, motif slots should show
larger entropy or explicit posterior mass split across the competing factors.
