---
title: "MF Poisson SuSiE With Overspecified K"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This page stress-tests the MF-Poisson-SuSiE fit when the number of fitted
factors is larger than the true number of factors. The main question is whether
the fitted model still recovers the true factor rows and sparse supports, and
whether the extra factor rows remain weakly used or become duplicates.

```{r source-code}
source("code/joint-learn-susie-poi-F.R")
```

## Simulate Data

The data-generating mechanism matches the current joint-learning experiment:
five true factor rows, three possible effects per observation, five one-factor
groups, and three three-factor groups.

```{r simulate}
set.seed(42)

N = 1600
M = 2000
K_true = 5
D = 3
K_fit = 7

F0 = matrix(rgamma(M * K_true, shape = 0.1, rate = 0.01),
            nrow = K_true, ncol = M)
F0 = F0 / rowSums(F0)

n_per_group = N / 8
L = matrix(0, nrow = N, ncol = K_true)
grp = rep(1:8, each = n_per_group)

for (g in 1:5) {
  L[grp == g, g] = matrix(rgamma(n_per_group, 10, 1), n_per_group, 1)
}

L[grp == 6, c(1, 2, 3)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 7, c(1, 2, 4)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 8, c(2, 3, 5)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)

rate = L %*% F0
Y = matrix(rpois(N * M, rate), nrow = N, ncol = M)

data.frame(
  N = N,
  M = M,
  K_true = K_true,
  K_fit = K_fit,
  D = D,
  total_count = sum(Y),
  nonzero_entries = sum(Y > 0)
)
```

```{r true-patterns}
true_patterns = vapply(1:8, function(g) {
  paste(which(colSums(L[grp == g, , drop = FALSE]) > 0), collapse = ",")
}, character(1))

data.frame(group = 1:8, true_active_factors = true_patterns)
```

## Diagnostics

With `K_fit > K_true`, factor recovery is rectangular. We match each true factor
to one learned factor row, then separately inspect the usage and nearest true
factor for every learned row.

```{r diagnostic-helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_true_to_learned = function(F_hat, F_true) {
  K_fit = nrow(F_hat)
  K_true = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)

  best_score = -Inf
  best_learned_for_true = rep(NA_integer_, K_true)

  learned_sets = combn(seq_len(K_fit), K_true, simplify = FALSE)
  for (learned_set in learned_sets) {
    perms = all_permutations(learned_set)
    scores = apply(perms, 1, function(p) {
      sum(similarity[cbind(p, seq_len(K_true))])
    })
    idx = which.max(scores)
    if (scores[idx] > best_score) {
      best_score = scores[idx]
      best_learned_for_true = perms[idx, ]
    }
  }

  learned_to_true = rep(NA_integer_, K_fit)
  learned_to_true[best_learned_for_true] = seq_len(K_true)

  nearest_true = apply(similarity, 1, which.max)
  nearest_cosine = similarity[cbind(seq_len(K_fit), nearest_true)]

  list(
    match = data.frame(
      true_factor = seq_len(K_true),
      learned_factor = best_learned_for_true,
      cosine = similarity[cbind(best_learned_for_true, seq_len(K_true))]
    ),
    learned_rows = data.frame(
      learned_factor = seq_len(K_fit),
      matched_true_factor = learned_to_true,
      nearest_true_factor = nearest_true,
      nearest_true_cosine = nearest_cosine
    ),
    learned_to_true = learned_to_true,
    nearest_true = nearest_true
  )
}

loading_scores = function(fit) {
  E_beta = fit$alpha / fit$lambda
  apply(fit$gamma_bar * E_beta, c(1, 3), sum)
}

assignment_scores = function(fit) {
  apply(fit$gamma_bar, c(1, 3), sum)
}

posterior_inclusion_scores = function(fit) {
  1 - apply(1 - fit$gamma_bar, c(1, 3), prod)
}

map_scores_to_true = function(scores, learned_to_true, K_true) {
  out = matrix(0, nrow = nrow(scores), ncol = K_true)
  for (k in seq_along(learned_to_true)) {
    if (!is.na(learned_to_true[k])) {
      out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
    }
  }
  out
}

support_labels_from_scores = function(scores, support_sizes) {
  vapply(seq_len(nrow(scores)), function(i) {
    active = order(scores[i, ], decreasing = TRUE)[seq_len(support_sizes[i])]
    paste(sort(active), collapse = ",")
  }, character(1))
}

normalized_entropy = function(scores) {
  prob = scores / rowSums(scores)
  entropy = -prob * log(prob)
  entropy[is.nan(entropy)] = 0
  mean(rowSums(entropy)) / log(ncol(scores))
}

plot_cosine_heatmap = function(cosine_mat, main) {
  image(x = seq_len(ncol(cosine_mat)), y = seq_len(nrow(cosine_mat)),
        z = t(cosine_mat), axes = FALSE, xlab = "True factor",
        ylab = "Fitted factor", main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(ncol(cosine_mat)),
       labels = paste0("true ", seq_len(ncol(cosine_mat))))
  axis(2, at = seq_len(nrow(cosine_mat)),
       labels = paste0("fit ", seq_len(nrow(cosine_mat))), las = 1)
  for (k_fit in seq_len(nrow(cosine_mat))) {
    for (k_true in seq_len(ncol(cosine_mat))) {
      text(k_true, k_fit, sprintf("%.2f", cosine_mat[k_fit, k_true]),
           cex = 0.75)
    }
  }
  box()
}

plot_usage = function(loading_usage, assignment_usage, nearest_true) {
  usage_mat = rbind(loading = loading_usage, assignment = assignment_usage)
  barplot(usage_mat, beside = TRUE, names.arg = paste0("fit ", seq_along(loading_usage)),
          xlab = "Fitted factor", ylab = "Usage fraction",
          main = "Fitted factor usage", col = c("firebrick", "gray50"),
          ylim = c(0, max(usage_mat) * 1.2))
  legend("topright", legend = rownames(usage_mat),
         fill = c("firebrick", "gray50"), bty = "n")
  mtext(paste("nearest true:", paste(nearest_true, collapse = " ")),
        side = 1, line = 4, cex = 0.8)
}

plot_recovered_support = function(scores_true_space, main) {
  support_size = true_support_size
  recovered = matrix(0, nrow = K_true, ncol = N)
  for (i in seq_len(N)) {
    active = order(scores_true_space[i, ], decreasing = TRUE)[seq_len(support_size[i])]
    recovered[active, i] = 1
  }

  image(x = seq_len(N), y = seq_len(K_true), z = t(recovered),
        axes = FALSE, xlab = "Sample", ylab = "Recovered true factor",
        main = main, col = c("white", "firebrick"))
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", 1:8))
  axis(2, at = seq_len(K_true), labels = seq_len(K_true), las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  box()
}

plot_recovered_support_fit = function(scores_fit_space, main) {
  recovered = matrix(0, nrow = ncol(scores_fit_space), ncol = N)
  for (i in seq_len(N)) {
    active = order(scores_fit_space[i, ], decreasing = TRUE)[
      seq_len(true_support_size[i])
    ]
    recovered[active, i] = 1
  }

  image(x = seq_len(N), y = seq_len(ncol(scores_fit_space)), z = t(recovered),
        axes = FALSE, xlab = "Sample", ylab = "Recovered fitted factor",
        main = main, col = c("white", "firebrick"))
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", 1:8))
  axis(2, at = seq_len(ncol(scores_fit_space)),
       labels = seq_len(ncol(scores_fit_space)), las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  box()
}

conditional_scores = function(scores, factor_set, sample_idx) {
  out = scores[sample_idx, factor_set, drop = FALSE]
  out = out / rowSums(out)
  out[is.nan(out)] = 0
  out
}

row_normalized_entropy = function(prob) {
  entropy = -prob * log(prob)
  entropy[is.nan(entropy)] = 0
  rowSums(entropy) / log(ncol(prob))
}

plot_competing_factor_shares = function(prob, sample_idx, factor_set, main) {
  sample_group = grp[sample_idx]
  group_run = rle(sample_group)
  group_end = cumsum(group_run$lengths)
  group_mid = group_end - group_run$lengths / 2

  image(x = seq_along(sample_idx), y = seq_along(factor_set), z = prob,
        axes = FALSE, xlab = "Sample with target true factor active",
        ylab = "Competing fitted factor", main = main,
        col = hcl.colors(100, "YlOrRd"))
  axis(1, at = group_mid, labels = paste("G", group_run$values))
  axis(2, at = seq_along(factor_set),
       labels = paste0("fit ", factor_set), las = 1)
  if (length(group_end) > 1) {
    abline(v = group_end[-length(group_end)] + 0.5, col = "gray80", lty = 3)
  }
  box()
}

plot_competing_entropy = function(entropy, sample_idx, main) {
  boxplot(entropy ~ factor(grp[sample_idx]), xlab = "Group",
          ylab = "Normalized entropy", main = main)
}

candidate_competing_factors = function(true_factor, top_n = 3) {
  nearest_set = which(factor_match$nearest_true == true_factor)
  top_set = order(cosine_fit_true[, true_factor], decreasing = TRUE)[
    seq_len(min(top_n, nrow(cosine_fit_true)))
  ]
  sort(unique(c(nearest_set, top_set)))
}

competing_factor_diagnostics = function(true_factor, factor_set = NULL,
                                        top_n = 3) {
  if (is.null(factor_set)) {
    factor_set = candidate_competing_factors(true_factor, top_n)
  }
  sample_idx = which(L[, true_factor] > 0)

  pip_share = conditional_scores(pip_scores, factor_set, sample_idx)
  loading_share = conditional_scores(fit_loading_scores, factor_set, sample_idx)
  pip_entropy = row_normalized_entropy(pip_share)
  loading_entropy = row_normalized_entropy(loading_share)

  list(
    true_factor = true_factor,
    factor_set = factor_set,
    sample_idx = sample_idx,
    pip_share = pip_share,
    loading_share = loading_share,
    pip_entropy = pip_entropy,
    loading_entropy = loading_entropy,
    summary = data.frame(
      true_factor = true_factor,
      candidate_fitted_factors = paste(factor_set, collapse = ","),
      n_samples_with_true_factor = length(sample_idx),
      mean_candidate_cosine = mean(cosine_fit_true[factor_set, true_factor]),
      max_candidate_cosine = max(cosine_fit_true[factor_set, true_factor]),
      total_candidate_loading_usage = sum(loading_usage[factor_set]),
      mean_pip_entropy = mean(pip_entropy),
      median_pip_entropy = median(pip_entropy),
      mean_top_pip_share = mean(apply(pip_share, 1, max)),
      mean_loading_entropy = mean(loading_entropy),
      median_loading_entropy = median(loading_entropy),
      mean_top_loading_share = mean(apply(loading_share, 1, max))
    )
  )
}

plot_competing_diagnostics = function(diag, label) {
  plot_competing_factor_shares(
    diag$pip_share,
    diag$sample_idx,
    diag$factor_set,
    paste("Conditional PIP share among candidate factors for", label)
  )
  plot_competing_factor_shares(
    diag$loading_share,
    diag$sample_idx,
    diag$factor_set,
    paste("Conditional loading share among candidate factors for", label)
  )
  plot_competing_entropy(
    diag$pip_entropy,
    diag$sample_idx,
    paste("PIP uncertainty among candidate factors for", label)
  )
}

true_support_size = rowSums(L > 0)
true_support_label = apply(L > 0, 1, function(x) {
  paste(which(x), collapse = ",")
})
```

## Fit Overspecified K

```{r fit}
fit = mf_poi_susie(
  Y = Y,
  K = K_fit,
  D = D,
  max_iters = 100,
  update_prior = TRUE,
  break_symmetry = FALSE,
  init_seed = 1,
  init_F = "poisson_nmf",
  nmf_iters = 200,
  init_gamma_from_nmf = TRUE,
  gamma_step_init = 0.1,
  gamma_step_ramp = 50,
  F_step_init = 0.05,
  F_step_ramp = 75,
  elbo_every = 5,
  tol = 5e-4,
  min_iters = 20,
  patience = 2
)
```

```{r elbo}
idx = fit$elbo_iter
plot(idx, fit$elbo[idx], type = "l", xlab = "Iteration", ylab = "ELBO",
     main = "Variational objective")
points(idx, fit$elbo[idx], pch = 19, cex = 0.5)

data.frame(
  converged = fit$converged,
  n_iter = fit$n_iter,
  final_elbo = tail(fit$elbo[fit$elbo_iter], 1)
)
```

## Factor Recovery

```{r factor-recovery}
factor_match = match_true_to_learned(fit$F, F0)

factor_match$match
```

```{r factor-cosine-heatmap}
cosine_fit_true = row_cosine(fit$F, F0)
plot_cosine_heatmap(
  cosine_fit_true,
  "Cosine similarity between fitted and true factors"
)
```

```{r learned-row-diagnostics}
fit_loading_scores = loading_scores(fit)
fit_assignment_scores = assignment_scores(fit)

loading_usage = colSums(fit_loading_scores) / sum(fit_loading_scores)
assignment_usage = colSums(fit_assignment_scores) / sum(fit_assignment_scores)

learned_row_summary = transform(
  factor_match$learned_rows,
  loading_usage = loading_usage,
  assignment_usage = assignment_usage
)

learned_row_summary[order(learned_row_summary$loading_usage, decreasing = TRUE), ]

data.frame(
  mean_true_match_cosine = mean(factor_match$match$cosine),
  min_true_match_cosine = min(factor_match$match$cosine),
  total_unmatched_loading_usage =
    sum(loading_usage[is.na(factor_match$learned_to_true)]),
  max_unmatched_loading_usage =
    max(loading_usage[is.na(factor_match$learned_to_true)])
)
```

```{r fitted-factor-usage}
plot_usage(
  loading_usage = loading_usage,
  assignment_usage = assignment_usage,
  nearest_true = factor_match$nearest_true
)
```

```{r learned-factor-shapes}
ord = order(loading_usage, decreasing = TRUE)
matplot(t(fit$F[ord, 1:150]), type = "l", lty = 1,
        xlab = "m", ylab = "F[k, m]",
        main = "Learned factor rows ordered by usage")
legend("topright", legend = paste0("learned ", ord), col = seq_along(ord),
       lty = 1, bty = "n")
```

## Support Recovery

Support recovery is evaluated after mapping learned factor rows to their nearest
true factor. This gives a generous view of sparsity recovery even if the
overspecified model splits one true factor across multiple learned rows.

```{r support-recovery}
scores_nearest_true = map_scores_to_true(
  fit_loading_scores,
  factor_match$nearest_true,
  K_true
)

support_recovery = data.frame(
  exact_support_accuracy = mean(
    support_labels_from_scores(scores_nearest_true, true_support_size) ==
      true_support_label
  ),
  mean_normalized_loading_entropy = normalized_entropy(scores_nearest_true)
)

support_recovery
```

The first plot below keeps all fitted factor rows, so extra factors appear
explicitly if the overspecified fit uses them in the strongest posterior loading
scores. The second plot collapses fitted rows to the nearest true factor. Thus,
the all-fitted plot diagnoses factor splitting, whereas the collapsed plot asks
whether the correct true support is still recovered after accounting for that
splitting.

```{r recovered-fitted-factors}
plot_recovered_support_fit(
  fit_loading_scores,
  "Recovered supports over all fitted factors"
)
```

```{r recovered-factors}
plot_recovered_support(
  scores_nearest_true,
  "Recovered supports with overspecified K"
)
```

The cosine heatmap and all-fitted support plot show that the extra factors are
not simply unused. Some fitted rows compete to represent the same true factor.
The collapsed support plot can still look accurate because those competing rows
are mapped back to the same true factor.

## Posterior Uncertainty Among Competing Factors

SuSiE-style posterior uncertainty about factor identity is represented most
directly by `gamma_bar`, or by posterior inclusion probabilities derived from
`gamma_bar`. The auxiliary variables `xi` allocate observed counts across
effects and features; they are useful for fitting, but they are not the most
direct summary of uncertainty among highly correlated fitted factor rows.

Here we identify true factors whose nearest fitted rows are duplicated. For the
duplicated true factor with the largest fitted loading usage, we inspect how the
posterior mass is split among its competing fitted rows.

```{r competing-factor-summary}
pip_scores = posterior_inclusion_scores(fit)

competing_sets = split(seq_len(K_fit), factor_match$nearest_true)
competing_sets = competing_sets[vapply(competing_sets, length, integer(1)) > 1]

competing_summary = do.call(rbind, lapply(names(competing_sets), function(k0) {
  factor_set = competing_sets[[k0]]
  data.frame(
    true_factor = as.integer(k0),
    fitted_factors = paste(factor_set, collapse = ","),
    n_fitted_factors = length(factor_set),
    mean_nearest_cosine =
      mean(cosine_fit_true[factor_set, as.integer(k0)]),
    total_loading_usage = sum(loading_usage[factor_set]),
    total_assignment_usage = sum(assignment_usage[factor_set]),
    row.names = NULL
  )
}))

competing_summary[order(competing_summary$total_loading_usage,
                        decreasing = TRUE), ]
```

```{r selected-competing-factor}
selected_true_factor = competing_summary$true_factor[
  which.max(competing_summary$total_loading_usage)
]
selected_fitted_factors = competing_sets[[as.character(selected_true_factor)]]
selected_diag = competing_factor_diagnostics(
  selected_true_factor,
  factor_set = selected_fitted_factors
)

selected_diag$summary
```

```{r competing-pip-heatmap}
plot_competing_factor_shares(
  selected_diag$pip_share,
  selected_diag$sample_idx,
  selected_diag$factor_set,
  paste("Conditional PIP share among competitors for true factor",
        selected_true_factor)
)
```

```{r competing-loading-heatmap}
plot_competing_factor_shares(
  selected_diag$loading_share,
  selected_diag$sample_idx,
  selected_diag$factor_set,
  paste("Conditional loading share among competitors for true factor",
        selected_true_factor)
)
```

```{r competing-entropy-by-group}
plot_competing_entropy(
  selected_diag$pip_entropy,
  selected_diag$sample_idx,
  paste("PIP uncertainty among competitors for true factor",
        selected_true_factor)
)
```

High conditional entropy means that the posterior does not strongly distinguish
between the competing fitted rows. Low entropy means one fitted row dominates.
The PIP-share heatmap is the closest diagnostic to SuSiE's usual uncertainty
quantification; the loading-share heatmap additionally asks whether that
uncertainty matters for reconstructed loading mass.

### Candidate Competitors For True Factors 3 And 4

The duplicated-nearest-factor rule is strict. We also inspect true factors 3 and
4 by taking the fitted rows whose factor shapes are most cosine-similar to each
target true factor. This asks whether the weaker visually apparent competition
corresponds to genuine posterior uncertainty or only to small shape similarity.

```{r competing-factor-3-summary}
factor3_diag = competing_factor_diagnostics(3, top_n = 3)
factor3_diag$summary
```

```{r competing-factor-3-plots, fig.height=9}
old_par = par(mfrow = c(3, 1), mar = c(4.5, 4.5, 3, 1))
plot_competing_diagnostics(factor3_diag, "true factor 3")
par(old_par)
```

```{r competing-factor-4-summary}
factor4_diag = competing_factor_diagnostics(4, top_n = 3)
factor4_diag$summary
```

```{r competing-factor-4-plots, fig.height=9}
old_par = par(mfrow = c(3, 1), mar = c(4.5, 4.5, 3, 1))
plot_competing_diagnostics(factor4_diag, "true factor 4")
par(old_par)
```

For these candidate sets, compare PIP entropy with loading entropy. If PIP
entropy is elevated but loading entropy is low, the model is uncertain about
which duplicate row receives the label, but the reconstructed loading mass is
still concentrated. If both are elevated, the overfitted rows are materially
splitting the signal.

## Interpreting `gamma_bar` As Loading Uncertainty

The posterior assignment probabilities `gamma_bar[i, d, k]` are the right object
for quantifying posterior uncertainty in the loadings conditional on the fitted
factor matrix `F`. Given `F`, each observation has a loading representation
\[
  L_{ik} \approx \sum_{d=1}^D \beta_{id} 1\{\gamma_{id}=k\}.
\]
Thus `gamma_bar[i, d, k]` measures how much posterior probability effect `d` in
observation `i` assigns to fitted factor `k`. When fitted factors are highly
similar because `K` is overspecified, diffuse posterior mass across those
similar fitted rows means the loading support is uncertain between them.

In this file we summarize that uncertainty in two ways. First, the fitted-factor
PIP
\[
  \mathrm{PIP}_{ik} = 1 - \prod_{d=1}^D (1 - \bar\gamma_{idk})
\]
is the posterior probability that at least one effect in observation `i` uses
fitted factor `k`. Conditional PIP shares among a set of competing fitted factors
show which duplicate row receives the support probability. Entropy of those
shares is high when the posterior cannot confidently choose among the competing
rows.

Second, the posterior expected loading score
\[
  S_{ik} = \sum_{d=1}^D \bar\gamma_{idk}
    E_q[\beta_{id} \mid \gamma_{id}=k]
\]
weights assignment uncertainty by expected effect size. This distinguishes pure
label ambiguity from uncertainty that materially affects the reconstructed
loading mass. High PIP entropy but low loading-share entropy means the posterior
is uncertain about labels but the loading mass is mostly concentrated. High
entropy in both summaries means the overfitted factors are genuinely splitting
the loading signal.
