---
title: "MF Poisson SuSiE With Overspecified D"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This page stress-tests the MF-Poisson-SuSiE fit when the number of fitted
single-effect components is larger than the true sparsity dimension. The main
question is whether the fitted model still recovers the true supports when it
has extra effects available, and whether those extra effects carry small
posterior loading mass.

```{r source-code}
source("code/joint-learn-susie-poi-F.R")
```

## Simulate Data

The data-generating mechanism has five true factor rows and at most three active
effects per observation. The fitted model uses the correct `K` but fits a larger
`D`.

```{r simulate}
set.seed(42)

N = 1600
M = 2000
K = 5
D_true = 3
D_fit = 5

F0 = matrix(rgamma(M * K, shape = 0.1, rate = 0.01), nrow = K, ncol = M)
F0 = F0 / rowSums(F0)

n_per_group = N / 8
L = matrix(0, nrow = N, ncol = K)
grp = rep(1:8, each = n_per_group)

for (g in 1:5) {
  L[grp == g, g] = matrix(rgamma(n_per_group, 10, 1), n_per_group, 1)
}

L[grp == 6, c(1, 2, 3)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 7, c(1, 2, 4)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)
L[grp == 8, c(2, 3, 5)] = matrix(rgamma(n_per_group * 3, 20, 1),
                                  n_per_group, 3)

rate = L %*% F0
Y = matrix(rpois(N * M, rate), nrow = N, ncol = M)

data.frame(
  N = N,
  M = M,
  K = K,
  D_true = D_true,
  D_fit = D_fit,
  total_count = sum(Y),
  nonzero_entries = sum(Y > 0)
)
```

```{r true-patterns}
true_patterns = vapply(1:8, function(g) {
  paste(which(colSums(L[grp == g, , drop = FALSE]) > 0), collapse = ",")
}, character(1))

data.frame(group = 1:8, true_active_factors = true_patterns)
```

## Diagnostics

When `D` is overspecified, raw assignment probabilities can be misleading:
every fitted effect must choose a factor. We therefore score support recovery
using posterior expected loading mass,
\[
  \sum_d q(\gamma_{id} = k) E_q[\beta_{id} \mid \gamma_{id}=k],
\]
so that extra effects with small posterior size have little influence.

```{r diagnostic-helpers}
row_cosine = function(A, B) {
  A_norm = sqrt(rowSums(A^2))
  B_norm = sqrt(rowSums(B^2))
  (A %*% t(B)) / (A_norm %o% B_norm)
}

all_permutations = function(x) {
  if (length(x) == 1) return(matrix(x, nrow = 1))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[i], all_permutations(x[-i]))
  }))
}

match_factor_rows = function(F_hat, F_true) {
  K = nrow(F_true)
  similarity = row_cosine(F_hat, F_true)
  perms = all_permutations(seq_len(K))
  scores = apply(perms, 1, function(p) {
    sum(similarity[cbind(seq_len(K), p)])
  })
  learned_to_true = perms[which.max(scores), ]

  data.frame(
    learned_factor = seq_len(K),
    matched_true_factor = learned_to_true,
    cosine = similarity[cbind(seq_len(K), learned_to_true)]
  )
}

loading_scores = function(fit) {
  E_beta = fit$alpha / fit$lambda
  apply(fit$gamma_bar * E_beta, c(1, 3), sum)
}

effect_sizes = function(fit) {
  E_beta = fit$alpha / fit$lambda
  rowSums(fit$gamma_bar * E_beta, dims = 2)
}

effect_gamma_entropy = function(fit) {
  entropy = -fit$gamma_bar * log(fit$gamma_bar)
  entropy[is.nan(entropy)] = 0
  rowSums(entropy, dims = 2) / log(dim(fit$gamma_bar)[3])
}

effect_max_gamma = function(fit) {
  apply(fit$gamma_bar, c(1, 2), max)
}

effect_top_factor = function(fit) {
  apply(fit$gamma_bar, c(1, 2), which.max)
}

rank_by_effect_size = function(effect_mat, size_mat) {
  out = matrix(0, nrow = nrow(effect_mat), ncol = ncol(effect_mat))
  for (i in seq_len(nrow(effect_mat))) {
    ord = order(size_mat[i, ], decreasing = TRUE)
    out[i, ] = effect_mat[i, ord]
  }
  out
}

rank_gamma_bar_by_size = function(fit, size_mat) {
  N = dim(fit$gamma_bar)[1]
  D = dim(fit$gamma_bar)[2]
  K = dim(fit$gamma_bar)[3]
  out = array(0, dim = c(N, D, K))
  for (i in seq_len(N)) {
    ord = order(size_mat[i, ], decreasing = TRUE)
    out[i, , ] = fit$gamma_bar[i, ord, ]
  }
  out
}

map_scores_to_true = function(scores, learned_to_true) {
  K = length(learned_to_true)
  out = matrix(0, nrow = nrow(scores), ncol = K)
  for (k in seq_len(K)) {
    out[, learned_to_true[k]] = out[, learned_to_true[k]] + scores[, k]
  }
  out
}

support_labels_from_scores = function(scores, support_sizes) {
  vapply(seq_len(nrow(scores)), function(i) {
    active = order(scores[i, ], decreasing = TRUE)[seq_len(support_sizes[i])]
    paste(sort(active), collapse = ",")
  }, character(1))
}

normalized_entropy = function(scores) {
  prob = scores / rowSums(scores)
  entropy = -prob * log(prob)
  entropy[is.nan(entropy)] = 0
  mean(rowSums(entropy)) / log(ncol(scores))
}

extra_effect_mass_fraction = function(fit, true_support_size) {
  sizes = effect_sizes(fit)
  vapply(seq_len(nrow(sizes)), function(i) {
    sorted_sizes = sort(sizes[i, ], decreasing = TRUE)
    extra_idx = seq.int(true_support_size[i] + 1, length(sorted_sizes))
    sum(sorted_sizes[extra_idx]) / sum(sorted_sizes)
  }, numeric(1))
}

plot_recovered_support = function(scores_true_space, main) {
  recovered = matrix(0, nrow = K, ncol = N)
  for (i in seq_len(N)) {
    active = order(scores_true_space[i, ], decreasing = TRUE)[
      seq_len(true_support_size[i])
    ]
    recovered[active, i] = 1
  }

  image(x = seq_len(N), y = seq_len(K), z = t(recovered),
        axes = FALSE, xlab = "Sample", ylab = "Recovered true factor",
        main = main, col = c("white", "firebrick"))
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", 1:8))
  axis(2, at = seq_len(K), labels = seq_len(K), las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  box()
}

plot_rank_heatmap = function(rank_mat, main, ylab, col) {
  image(x = seq_len(N), y = seq_len(ncol(rank_mat)), z = rank_mat,
        axes = FALSE, xlab = "Sample", ylab = ylab,
        main = main, col = col)
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", 1:8))
  axis(2, at = seq_len(ncol(rank_mat)), labels = seq_len(ncol(rank_mat)),
       las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  box()
}

plot_factor_id_heatmap = function(rank_top_factor, main) {
  image(x = seq_len(N), y = seq_len(ncol(rank_top_factor)), z = rank_top_factor,
        axes = FALSE, xlab = "Sample", ylab = "Effect rank",
        main = main, col = seq_len(K))
  axis(1, at = seq(n_per_group / 2, N, by = n_per_group),
       labels = paste("G", 1:8))
  axis(2, at = seq_len(ncol(rank_top_factor)),
       labels = seq_len(ncol(rank_top_factor)), las = 1)
  abline(v = seq(n_per_group, N - n_per_group, by = n_per_group),
         col = "gray80", lty = 3)
  legend("topright", legend = paste("factor", seq_len(K)), fill = seq_len(K),
         bty = "n", cex = 0.8)
  box()
}

plot_single_gamma_heatmap = function(gamma_mat, main) {
  D = nrow(gamma_mat)
  K = ncol(gamma_mat)
  image(x = seq_len(K), y = seq_len(D), z = t(gamma_mat),
        axes = FALSE, xlab = "Factor", ylab = "Effect rank",
        main = main, col = hcl.colors(100, "YlOrRd"))
  axis(1, at = seq_len(K), labels = seq_len(K))
  axis(2, at = seq_len(D), labels = seq_len(D), las = 1)
  box()
}

plot_representative_gamma = function(rank_gamma_bar, representative_idx) {
  op = par(mfrow = c(2, 4), mar = c(4, 4, 3, 1))
  on.exit(par(op))
  for (j in seq_along(representative_idx)) {
    i = representative_idx[j]
    plot_single_gamma_heatmap(
      rank_gamma_bar[i, , ],
      paste("Group", grp[i], "sample", i)
    )
  }
}

true_support_size = rowSums(L > 0)
true_support_label = apply(L > 0, 1, function(x) {
  paste(which(x), collapse = ",")
})
```

## Fit Overspecified D

```{r fit}
fit = mf_poi_susie(
  Y = Y,
  K = K,
  D = D_fit,
  max_iters = 100,
  update_prior = TRUE,
  break_symmetry = FALSE,
  init_seed = 1,
  init_F = "poisson_nmf",
  nmf_iters = 200,
  init_gamma_from_nmf = TRUE,
  gamma_step_init = 0.1,
  gamma_step_ramp = 50,
  F_step_init = 0.05,
  F_step_ramp = 75,
  elbo_every = 5,
  tol = 5e-4,
  min_iters = 20,
  patience = 2
)
```

```{r elbo}
idx = fit$elbo_iter
plot(idx, fit$elbo[idx], type = "l", xlab = "Iteration", ylab = "ELBO",
     main = "Variational objective")
points(idx, fit$elbo[idx], pch = 19, cex = 0.5)

data.frame(
  converged = fit$converged,
  n_iter = fit$n_iter,
  final_elbo = tail(fit$elbo[fit$elbo_iter], 1)
)
```

## Factor Recovery

```{r factor-recovery}
factor_match = match_factor_rows(fit$F, F0)
learned_to_true = factor_match$matched_true_factor

factor_match

data.frame(
  mean_matched_cosine = mean(factor_match$cosine),
  min_matched_cosine = min(factor_match$cosine)
)
```

## Support And Extra-Effect Diagnostics

```{r support-recovery}
scores_true_space = map_scores_to_true(loading_scores(fit), learned_to_true)
extra_mass = extra_effect_mass_fraction(fit, true_support_size)

support_recovery = data.frame(
  exact_support_accuracy = mean(
    support_labels_from_scores(scores_true_space, true_support_size) ==
      true_support_label
  ),
  mean_normalized_loading_entropy = normalized_entropy(scores_true_space),
  mean_extra_effect_mass_fraction = mean(extra_mass),
  median_extra_effect_mass_fraction = median(extra_mass),
  q90_extra_effect_mass_fraction = as.numeric(quantile(extra_mass, 0.9))
)

support_recovery
```

```{r extra-effect-by-group}
extra_effect_by_group = aggregate(
  extra_mass,
  by = list(group = grp),
  FUN = function(x) {
    c(mean = mean(x), median = median(x), q90 = quantile(x, 0.9))
  }
)
extra_effect_by_group = do.call(data.frame, extra_effect_by_group)
names(extra_effect_by_group) = c("group", "mean", "median", "q90")

extra_effect_by_group
```

```{r recovered-factors}
plot_recovered_support(
  scores_true_space,
  "Recovered supports with overspecified D"
)
```

```{r effect-size-plot}
sizes = effect_sizes(fit)
sorted_sizes = t(apply(sizes, 1, sort, decreasing = TRUE))

boxplot(
  split(as.vector(sorted_sizes), col(sorted_sizes)),
  xlab = "Effect rank within observation",
  ylab = "Posterior expected effect size",
  main = "Effect sizes after overspecifying D"
)
```

## Behavior Of The Extra Effects

Because fitted effects are exchangeable, we sort the `D_fit` effects within each
observation by posterior expected effect size. Rank 1 is the largest fitted
effect for that observation, and ranks beyond the true support size are the
overfitted dimensions.

```{r effect-rank-diagnostics}
sizes = effect_sizes(fit)
gamma_entropy = effect_gamma_entropy(fit)
max_gamma = effect_max_gamma(fit)
top_factor = effect_top_factor(fit)
rank_gamma_bar = rank_gamma_bar_by_size(fit, sizes)

rank_size = rank_by_effect_size(sizes, sizes)
rank_entropy = rank_by_effect_size(gamma_entropy, sizes)
rank_max_gamma = rank_by_effect_size(max_gamma, sizes)
rank_top_factor = rank_by_effect_size(top_factor, sizes)

rank_summary = do.call(rbind, lapply(seq_len(D_fit), function(r) {
  data.frame(
    effect_rank = r,
    mean_expected_size = mean(rank_size[, r]),
    median_expected_size = median(rank_size[, r]),
    mean_gamma_entropy = mean(rank_entropy[, r]),
    median_gamma_entropy = median(rank_entropy[, r]),
    mean_max_gamma = mean(rank_max_gamma[, r]),
    median_max_gamma = median(rank_max_gamma[, r]),
    selected_as_extra_fraction = mean(r > true_support_size)
  )
}))

rank_summary
```

```{r effect-rank-entropy-boxplot}
boxplot(
  split(as.vector(rank_entropy), col(rank_entropy)),
  xlab = "Effect rank within observation",
  ylab = "Normalized entropy of gamma_bar",
  main = "Posterior factor-assignment uncertainty by effect rank"
)
abline(v = D_true + 0.5, col = "gray50", lty = 2)
```

```{r effect-rank-size-entropy}
plot(
  as.vector(rank_size),
  as.vector(rank_entropy),
  pch = 20,
  col = rgb(0.6, 0, 0, 0.25),
  xlab = "Posterior expected effect size",
  ylab = "Normalized entropy of gamma_bar",
  main = "Small extra effects tend to have diffuse gamma_bar"
)
```

The boxplot and scatterplot ask whether overfitted dimensions are being used as
real effects or left diffuse. High entropy means an effect has no clear factor
assignment. Small size with high entropy is the cleanest behavior for an
overfitted dimension: the effect exists in the variational family but does not
carry much loading mass.

```{r effect-rank-size-heatmap}
plot_rank_heatmap(
  rank_size,
  "Posterior expected effect size by rank",
  "Effect rank",
  hcl.colors(100, "YlOrRd")
)
```

```{r effect-rank-entropy-heatmap}
plot_rank_heatmap(
  rank_entropy,
  "Normalized gamma_bar entropy by effect rank",
  "Effect rank",
  hcl.colors(100, "YlGnBu")
)
```

```{r effect-rank-max-gamma-heatmap}
plot_rank_heatmap(
  rank_max_gamma,
  "Maximum gamma_bar probability by effect rank",
  "Effect rank",
  hcl.colors(100, "YlOrRd")
)
```

```{r effect-rank-top-factor-heatmap}
plot_factor_id_heatmap(
  rank_top_factor,
  "Top posterior factor by effect rank"
)
```

The entropy heatmap and maximum-probability heatmap are complementary. A diffuse
extra effect should have high entropy and low maximum `gamma_bar`. The top-factor
heatmap shows what factor would be selected if we forced every overfitted effect
to choose a label; this plot should be interpreted together with the entropy and
size heatmaps, because a top label is not meaningful when `gamma_bar` is nearly
uniform.

## Representative Posterior Assignment Heatmaps

The following heatmaps mimic the single-observation view in `poisson-susie.Rmd`,
but for representative observations from each group. Rows are fitted effects
sorted by posterior expected size. Columns are factors. Extra low-size effects
should appear more diffuse across factors than the leading effects.

```{r representative-gamma-heatmaps, fig.width=10, fig.height=6}
representative_idx = vapply(seq_len(8), function(g) {
  which(grp == g)[1]
}, integer(1))

plot_representative_gamma(rank_gamma_bar, representative_idx)
```

## Interpreting Overfitted D

For fixed learned `F`, `gamma_bar[i, d, k]` quantifies posterior uncertainty in
which factor receives effect `d` for observation `i`. When `D` is overspecified,
there are more effects than the true loading support needs. A well-behaved fit
can handle this in two ways: extra effects can have small expected size, or they
can remain diffuse in `gamma_bar` so that no factor receives strong posterior
support from them.

The most useful summaries here are therefore effect size and entropy together.
Expected effect size,
\[
  E_q[\beta_{id}] =
  \sum_k \bar\gamma_{idk} E_q[\beta_{id} \mid \gamma_{id}=k],
\]
measures whether effect `d` carries meaningful loading mass. Normalized entropy
of `gamma_bar[i, d, ]` measures whether the factor identity of that effect is
resolved. Extra dimensions are benign when they have low expected size, high
assignment entropy, or both. They are problematic when they have sizable
expected loading mass and low entropy, because then the overfitted dimension is
acting like an additional confident component rather than an uncertain residual
effect.
