---
title: "Poisson SuSiE"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

This page demonstrates the single-observation Poisson SuSiE algorithm implemented
in `code/poisson-susie.R`.

```{r source-code}
source("code/poisson-susie.R")
```

## Model

For a count vector \(y = (y_1,\ldots,y_M)\) and a known nonnegative dictionary
\(F \in \mathbb{R}_+^{K \times M}\),
\[
  y_m \sim \mathrm{Poisson}\left(\sum_{d=1}^D \beta_d F_{\gamma_d m}\right),
  \qquad
  \gamma_d \sim \mathrm{Categorical}(1/K,\ldots,1/K),
  \qquad
  \beta_d \mid \gamma_d \sim \mathrm{Gamma}(\alpha_d^0,\lambda_d^0).
\]
The variational posterior assigns each effect \(d\) a categorical distribution
over the rows of \(F\), stored as `gamma_bar[d, k]`.

## Simulate Data

We simulate \(K = 10\) candidate factors and generate one count vector from three
active factors.

```{r simulate}
set.seed(5)

M = 2000
D = 5
K = 10

F = matrix(rgamma(M * K, shape = 0.1, rate = 0.01), nrow = K, ncol = M)

true_l = rep(0, K)
true_l[1:3] = c(1, 2, 3)

y_mean = as.numeric(true_l %*% F)
y = rpois(M, y_mean)

data.frame(
  M = M,
  K = K,
  D = D,
  total_count = sum(y),
  nonzero_bins = sum(y > 0)
)
```

```{r factor-shapes}
matplot(t(F[1:5, 1:150]), type = "l", lty = 1,
        xlab = "m", ylab = "F[k, m]",
        main = "First five dictionary rows")
legend("topright", legend = paste("factor", 1:5), col = 1:5, lty = 1,
       bty = "n")
```

## Fit

```{r fit}
alpha0 = rep(1, D)
lambda0 = rep(1, D)

fit = poi_susie(
  y = y,
  F = F,
  alpha0 = alpha0,
  lambda0 = lambda0,
  max_iters = 100,
  update_prior = TRUE,
  init_seed = 3,
  break_symmetry = FALSE
)
```

```{r elbo}
plot(fit$elbo, type = "l", xlab = "Iteration", ylab = "ELBO",
     main = "Variational objective")
```

## Recovery

Rows of `gamma_bar` correspond to the \(D\) single effects, and columns
correspond to candidate factors. The true active factors are 1, 2, and 3.

```{r gamma-table}
round(fit$gamma_bar, 3)
```

```{r top-factors}
top_factor = apply(fit$gamma_bar, 1, which.max)
posterior_mass = apply(fit$gamma_bar, 1, max)
estimated_beta = rowSums(fit$gamma_bar * fit$alpha / fit$lambda)

data.frame(
  effect = seq_len(D),
  top_factor = top_factor,
  posterior_mass = round(posterior_mass, 3),
  estimated_beta = round(estimated_beta, 3)
)
```

```{r gamma-heatmap}
op = par(mar = c(5, 4, 3, 6))
image(t(fit$gamma_bar[D:1, ]), axes = FALSE, xlab = "Factor", ylab = "Effect",
      main = "Posterior factor assignment probabilities")
axis(1, at = seq(0, 1, length.out = K), labels = seq_len(K))
axis(2, at = seq(0, 1, length.out = D), labels = D:1)
box()
par(op)
```

The algorithm recovers the active factors up to permutation of the \(D\) effects.
Extra effects can remain diffuse or attach weakly depending on initialization and
the learned prior scale.

## Fit With Symmetry Breaking

The `break_symmetry` option checks whether two active effects have nearly the
same posterior assignment vector. If they do, one effect is averaged with the
other and the duplicate is reset to the uniform distribution. This is intended
to discourage two effects from explaining the same factor.

```{r fit-break-symmetry}
fit_break = poi_susie(
  y = y,
  F = F,
  alpha0 = alpha0,
  lambda0 = lambda0,
  max_iters = 100,
  update_prior = TRUE,
  init_seed = 3,
  break_symmetry = TRUE
)
```

```{r compare-elbo}
plot(fit$elbo, type = "l", col = "black", lwd = 2,
     xlab = "Iteration", ylab = "ELBO",
     main = "ELBO with and without symmetry breaking")
lines(fit_break$elbo, col = "red", lwd = 2)
legend("bottomright", legend = c("break_symmetry = FALSE",
                                 "break_symmetry = TRUE"),
       col = c("black", "red"), lwd = 2, bty = "n")
```

```{r gamma-table-break}
round(fit_break$gamma_bar, 3)
```

```{r top-factors-break}
top_factor_break = apply(fit_break$gamma_bar, 1, which.max)
posterior_mass_break = apply(fit_break$gamma_bar, 1, max)
estimated_beta_break = rowSums(fit_break$gamma_bar *
                                 fit_break$alpha / fit_break$lambda)

data.frame(
  effect = seq_len(D),
  top_factor = top_factor_break,
  posterior_mass = round(posterior_mass_break, 3),
  estimated_beta = round(estimated_beta_break, 3)
)
```

```{r compare-gamma-heatmaps, fig.width=9, fig.height=4}
op = par(mfrow = c(1, 2), mar = c(5, 4, 3, 2))
image(t(fit$gamma_bar[D:1, ]), axes = FALSE, xlab = "Factor", ylab = "Effect",
      main = "break_symmetry = FALSE")
axis(1, at = seq(0, 1, length.out = K), labels = seq_len(K))
axis(2, at = seq(0, 1, length.out = D), labels = D:1)
box()

image(t(fit_break$gamma_bar[D:1, ]), axes = FALSE,
      xlab = "Factor", ylab = "Effect",
      main = "break_symmetry = TRUE")
axis(1, at = seq(0, 1, length.out = K), labels = seq_len(K))
axis(2, at = seq(0, 1, length.out = D), labels = D:1)
box()
par(op)
```

With this seed, symmetry breaking keeps effects from collapsing onto identical
posterior assignment vectors. The active factors are still recovered up to
permutation, but the extra effects are more likely to be reset toward diffuse
assignments instead of duplicating an already-explained factor. This can make
the interpretation cleaner when \(D\) is larger than the true number of active
components, although the reset step can also introduce small ELBO jumps because
it perturbs the current coordinate-ascent solution.
